---
title: "Introduction to R Graphing"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

This brief tutorial will demonstrate how to create a basic plot in R from a text file of data.  This introduction provides an entry point for those unfamiliar with R (or a refresher for those who are rusty).  We will start with a very minimal piece of code and work our way up to code that automates the creation of 12 different pdf files, each with a different X-Y scatterplot.

####Learning objectives:####

* Gain familiarity with R
* Read data from a file
* Visualize data in a graph
* Understand the principle of control flow

In this tutorial, we will be using the 'gapminder' dataset, available here: <a href="http://tinyurl.com/qb83k3z"> http://tinyurl.com/qb83k3z</a> (right-click or Ctrl-click on link and Save As...).

###Your first plot###
```{r}
#######################################
# Plot points
#######################################
# Read in data
gapminder.all <- read.csv(file = "data/gapminder-FiveYearData.csv")
```
When reading data into R, it is always a good idea to make sure you read in the data correctly.  There are a number of ways to investigate your data, but three common methods are:

* `head` shows us the first 6 rows in a data frame.
* `str` (structure) provides some information about the data stored in the data frame.
* `summary` provides even more information about the data, including some summary statistics for numerical data.

```{r}
# Investigate data
head(gapminder.all)
str(gapminder.all)
summary(gapminder.all)
```

Because we are only interested in graphing data from 2002, we just pull out those data.
```{r}
# Subset data, retaining only those data from 2002
gapminder.2002 <- gapminder.all[gapminder.all$year == 2002, ]
```

Last but not least, we use the `plot` function to draw the plot.
```{r}
# Plot points
plot(x = gapminder.2002$gdpPercap, 
     y = gapminder.2002$lifeExp)
```

***
###Make it pretty###
A good start, but those axis labels are pretty unseemly.  We can fix those up by setting values in the `plot` function call.  Namely, we will be passing values to the `main`, `xlab` (x-axis label), and `ylab` (y-axis label) parameters.
```{r}
#######################################
# Add better axis titles
#######################################
# Read in data
gapminder.all <- read.csv(file = "data/gapminder-FiveYearData.csv")

# Subset data
gapminder.2002 <- gapminder.all[gapminder.all$year == 2002, ]

plot(x = gapminder.2002$gdpPercap, 
     y = gapminder.2002$lifeExp, 
     main = "Life expectancy v. GDP", 
     xlab = "GDP Per capita", 
     ylab = "Life expectancy (years)")
```

***
###Changing scales###
This does not look like a linear relationship, but we can try a simple log-transformation on the GDP data to see how that looks.
```{r}
#######################################
# Plot x-axis on log-scale
#######################################
# Read in data
gapminder.all <- read.csv(file = "data/gapminder-FiveYearData.csv")

# Subset data
gapminder.2002 <- gapminder.all[gapminder.all$year == 2002, ]

# Make new vector of Log GDP
gapminder.2002$Log10GDP <- log10(gapminder.2002$gdpPercap)

plot(x = gapminder.2002$Log10GDP, 
     y = gapminder.2002$lifeExp, 
     main = "Life expectancy v. GDP", 
     xlab = "Log(GDP Per capita)", 
     ylab = "Life expectancy (years)")
```

***
###Make it <em>prettier</em>###
We are not restricted to black and white colors.  Here we will color points by the continent each country is located on.
```{r}
#######################################
# Color points by continent + add legend
#######################################
# Read in data
gapminder.all <- read.csv(file = "data/gapminder-FiveYearData.csv")

# Subset data
gapminder.2002 <- gapminder.all[gapminder.all$year == 2002, ]

# Make new vector of Log GDP
gapminder.2002$Log10GDP <- log10(gapminder.2002$gdpPercap)
```

Start by looking at the different values in the `gapminder.2002$continent` vector.
```{r}
# What are the possible values for continent?
levels(gapminder.2002$continent)
```

There are five possible values, so we will need five different colors.  The first step is to create a new vector in the `gapminder.2002` data frame; call it `colors` and fill it with `NA` values.  Then assign colors based on the value in the `gapminder.2002$continent` vector.
```{r}
# Create new vector for colors
gapminder.2002$colors <- NA

# Assign colors based on gapminder.2002$continent
gapminder.2002$colors[gapminder.2002$continent == "Africa"] <- "red"
gapminder.2002$colors[gapminder.2002$continent == "Americas"] <- "orange"
gapminder.2002$colors[gapminder.2002$continent == "Asia"] <- "forestgreen"
gapminder.2002$colors[gapminder.2002$continent == "Europe"] <- "darkblue"
gapminder.2002$colors[gapminder.2002$continent == "Oceania"] <- "violet"

# Create main plot
plot(x = gapminder.2002$Log10GDP, 
     y = gapminder.2002$lifeExp, 
     main = "Life expectancy v. GDP", 
     xlab = "Log(GDP Per capita)", 
     ylab = "Life expectancy (years)",
     col = gapminder.2002$colors,
     pch = 18) # A diamond symbol

# We will also need to add a legend, so we know what the colors mean.
# Here we have to be sure the order of the colors matches the order 
# of the different levels of gapminder.2002$continents.
legend("topleft", 
       legend = levels(gapminder.2002$continent), 
       col = c("red", "orange3", "forestgreen", "darkblue", "violet"),
       pch = 18)
```

***
###Abstract code###
When writing code, abstraction can be quite useful, to ensure consistency throughout your code.
```{r}
#######################################
# Abstraction + a regression line
#######################################
# Read in data
gapminder.all <- read.csv(file = "data/gapminder-FiveYearData.csv")

# Subset data
gapminder.2002 <- gapminder.all[gapminder.all$year == 2002, ]

# Make new vector of Log GDP
gapminder.2002$Log10GDP <- log10(gapminder.2002$gdpPercap)

# Store values to use for some plotting parameters
symbol <- 18
sym.size <- 1.2
continents <- levels(gapminder.2002$continent)
continent.colors <- c("red", "orange", "forestgreen", "darkblue", "violet")

# Assign colors for each continent
gapminder.2002$colors <- NA
gapminder.2002$colors[gapminder.2002$continent == continents[1]] <- continent.colors[1]
gapminder.2002$colors[gapminder.2002$continent == continents[2]] <- continent.colors[2]
gapminder.2002$colors[gapminder.2002$continent == continents[3]] <- continent.colors[3]
gapminder.2002$colors[gapminder.2002$continent == continents[4]] <- continent.colors[4]
gapminder.2002$colors[gapminder.2002$continent == continents[5]] <- continent.colors[5]

# Create main plot
plot(x = gapminder.2002$Log10GDP, 
     y = gapminder.2002$lifeExp, 
     main = "Life expectancy v. GDP", 
     xlab = "Log(GDP Per capita)", 
     ylab = "Life expectancy (years)",
     col = gapminder.2002$colors,
     pch = symbol,
     cex = sym.size,
     lwd = 1.5)

# Add a legend - we don't have to worry about getting the colors 
# in the right order because we are using the same vector that we used 
# when assigning colors in the first place.
legend("topleft", 
       legend = continents, 
       col = continent.colors,
       pch = symbol)

# Add a regression line
lifeExp.lm <- lm(gapminder.2002$lifeExp ~ gapminder.2002$Log10GDP)
abline(lifeExp.lm, lty = 2, lwd = 2)
```

***
###More abstraction & loops###
Continuing with abstraction, we can simplify our code with control flow - here we use a `for` loop to accomplish the color assignment task.
```{r}
#######################################
# Control flow (a for loop)
#######################################
# Read in data
gapminder.all <- read.csv(file = "data/gapminder-FiveYearData.csv")

# Subset data
gapminder.2002 <- gapminder.all[gapminder.all$year == 2002, ]

# Make new vector of Log GDP
gapminder.2002$Log10GDP <- log10(gapminder.2002$gdpPercap)

# Store values to use for some plotting parameters
symbol <- 18
sym.size <- 1.2
continents <- levels(gapminder.2002$continent)
continent.colors <- c("red", "orange", "forestgreen", "darkblue", "violet")

# Establish first color as default
gapminder.2002$colors <- NA
# Loop over continents and assign colors
for (i in 1:length(continents)) {
  gapminder.2002$colors[gapminder.2002$continent == continents[i]] <- continent.colors[i]
}

# Create main plot
plot(x = gapminder.2002$Log10GDP, 
     y = gapminder.2002$lifeExp, 
     main = "Life expectancy v. GDP", 
     xlab = "Log(GDP Per capita)", 
     ylab = "Life expectancy (years)",
     col = gapminder.2002$colors,
     pch = symbol,
     cex = sym.size,
     lwd = 1.5)

# Add a legend
legend("topleft", 
       legend = continents, 
       col = continent.colors,
       pch = symbol)

# Add a regression line
lifeExp.lm <- lm(gapminder.2002$lifeExp ~ gapminder.2002$Log10GDP)
abline(lifeExp.lm, lty = 2, lwd = 2)
```

***
###Saving graphical output###
To save the file, we can redirect the output to a graphics device.  In this example we use a pdf writer; many other graphics devices are availble for writing different file formats, including `svg`, `jpeg`, and `png`.
```{r eval = FALSE}
#######################################
# Send to pdf
#######################################
# Read in data
gapminder.all <- read.csv(file = "data/gapminder-FiveYearData.csv")

# Subset data
gapminder.2002 <- gapminder.all[gapminder.all$year == 2002, ]

# Make new vector of Log GDP
gapminder.2002$Log10GDP <- log10(gapminder.2002$gdpPercap)

# Store values to use for some plotting parameters
symbol <- 18
sym.size <- 1.2
continents <- levels(gapminder.2002$continent)
continent.colors <- c("red", "orange", "forestgreen", "darkblue", "violet")

# Establish first color as default
gapminder.2002$colors <- NA
# Loop over continents and assign colors
for (i in 1:length(continents)) {
  gapminder.2002$colors[gapminder.2002$continent == continents[i]] <- continent.colors[i]
}

# Open PDF device
pdf(file = "output/Life_expectancy_graph.pdf", useDingbats = FALSE)
# Create main plot
plot(x = gapminder.2002$Log10GDP, 
     y = gapminder.2002$lifeExp, 
     main = "Life expectancy v. GDP", 
     xlab = "Log(GDP Per capita)", 
     ylab = "Life expectancy (years)",
     col = gapminder.2002$colors,
     pch = symbol,
     cex = sym.size,
     lwd = 1.5)

# Add a legend
legend("topleft", 
       legend = continents, 
       col = continent.colors,
       pch = symbol)

# Add a regression line
lifeExp.lm <- lm(gapminder.2002$lifeExp ~ gapminder.2002$Log10GDP)
abline(lifeExp.lm, lty = 2, lwd = 2)

# Close PDF device
dev.off()
```

***
###Automation!###
Finally (well, almost finally, see below), we can automate this process to create a separate pdf of the same graph for each year of data in the gapminder datasets (there are 12 years of data for each country).
```{r eval = FALSE}
#######################################
# Make one pdf for each of 12 years
#######################################
# Read in data
gapminder.all <- read.csv(file = "data/gapminder-FiveYearData.csv")

# Hold off on subsetting the data
# Make new vector of Log GDP
gapminder.all$Log10GDP <- log10(gapminder.all$gdpPercap)

# Store values to use for some plotting parameters
symbol <- 18
sym.size <- 1.2
continents <- levels(gapminder.all$continent)
continent.colors <- c("red", "orange", "forestgreen", "darkblue", "violet")

# Establish first color as default
gapminder.all$colors <- NA
# Loop over continents and assign colors
for (i in 1:length(continents)) {
  gapminder.all$colors[gapminder.all$continent == continents[i]] <- continent.colors[i]
}

# Find the unique values in the gapminder.all$year vector
years <- unique(gapminder.all$year)

# Now loop over each of the different years to create the pdfs.
for (curr.year in years) {
  # Subset data
  gapminder.one.year <- gapminder.all[gapminder.all$year == curr.year, ]
  
  # Open PDF device
  filename <- paste0("output/Life_exp_", curr.year, "_graph.pdf")
  pdf(file = filename, useDingbats = FALSE)
  # Create main plot
  plot(x = gapminder.one.year$Log10GDP, 
       y = gapminder.one.year$lifeExp, 
       main = "Life expectancy v. GDP", 
       sub = curr.year, 
       xlab = "Log(GDP Per capita)", 
       ylab = "Life expectancy (years)",
       col = gapminder.one.year$colors,
       pch = symbol,
       cex = sym.size,
       lwd = 1.5)
  
  # Add a legend
  legend("topleft", 
         legend = continents, 
         col = continent.colors,
         pch = symbol)
  
  # Add a regression line
  lifeExp.lm <- lm(gapminder.one.year$lifeExp ~ gapminder.one.year$Log10GDP)
  abline(lifeExp.lm, lty = 2, lwd = 2)
  
  # Close PDF device
  dev.off()
}
```

***
###Advanced topic: making it even prettier###
And for advanced users, we can add some transparency to the points (skipping the step were the plot is saved to a PDF file).
```{r}
#######################################
# Add transparency to colors
#######################################
# Read in data
gapminder.all <- read.csv(file = "data/gapminder-FiveYearData.csv")

# Subset data
gapminder.2002 <- gapminder.all[gapminder.all$year == 2002, ]

# Make new vector of Log GDP
gapminder.2002$Log10GDP <- log10(gapminder.2002$gdpPercap)

# Store values to use for some plotting parameters
symbol <- 18
sym.size <- 1.2
continents <- levels(gapminder.2002$continent)
continent.colors <- c("red", "orange", "forestgreen", "darkblue", "violet")
```

To add transparency, color names will not be enough - we will have to use RGB values and add an alpha (transparency) value.  Start by using the `col2rgb` function to see what the red, green, and blue values are for our five colors.
```{r}
# Investigating colors
col2rgb(continent.colors)
```
In this output, we see that each column corresponds to a color in our `continents.colors` vector (first column corresponds to "red", second column corresponds to "orange", etc.).  Each row corresponds to one of the three primary colors.

```{r}
# Convert colors to RGB, so we can add an alpha (transparency) value
continent.rgb <- col2rgb(continent.colors)
continent.colors <- NULL
opacity <- 150
# Loop over each column (i.e. color) in continent.rgb and extract Red, Green, and Blue values
for (color.column in 1:ncol(continent.rgb)) {
  new.color <- rgb(red = continent.rgb['red', color.column],
                   green = continent.rgb['green', color.column], 
                   blue = continent.rgb['blue', color.column],
                   alpha = opacity,
                   maxColorValue = 255)
  continent.colors[color.column] <- new.color
}

# Establish first color as default
gapminder.2002$colors <- NA
# loop over continents and assign colors
for (i in 1:length(continents)) {
  gapminder.2002$colors[gapminder.2002$continent == continents[i]] <- continent.colors[i]
}

# Create main plot
plot(x = gapminder.2002$Log10GDP, 
     y = gapminder.2002$lifeExp, 
     main = "Life expectancy v. GDP", 
     xlab = "Log(GDP Per capita)", 
     ylab = "Life expectancy (years)",
     col = gapminder.2002$colors,
     bg = gapminder.2002$colors, 
     pch = symbol,
     cex = sym.size, 
     lwd = 1.5)

# Add a legend
legend("topleft", 
       legend = continents, 
       col = continent.colors,
       pch = symbol)

# Add a regression line
lifeExp.lm <- lm(gapminder.2002$lifeExp ~ gapminder.2002$Log10GDP)
abline(lifeExp.lm, lty = 2, lwd = 2)
```

***
For advanced graphing, the <a href="http://ggplot2.org/">ggplot2</a> package is *extremely* useful.

Questions?  e-mail me at <a href="mailto:jcoliver@email.arizona.edu">jcoliver@email.arizona.edu</a>.