---
title: "Introduction to R Statistics"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

##What still needs to be covered?
* data structures (at least vectors and data.frames)
* boxplot (right before introducing script files)

An introduction to using the R statistics package and the RStudio interface.

####Learning objectives
1. Read data from files and output results
2. Extract relevant portions of datasets
3. Run standard statistical tests in R, including Student's t, linear regression
and analysis of variance
  
###Setup
First we need to setup our development environment. We need to create two folders: 'data' will store the data we will be analyzing, and 'output' will store the results of our analyses.
```{r, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

###Analysis of Variance (ANOVA)
For our first set of analyses, we'll use a dataset that comes pre-loaded in R. The `iris` data is from early statistical work of R.A. Fisher [doi: 10.1111%2Fj.1469-1809.1936.tb02137.x](https://dx.doi.org/10.1111%2Fj.1469-1809.1936.tb02137.x), who used three species of _Iris_ flowers to develop linear discriminant analysis. To begin, we'll look at the `data.frame` called `iris` using the `head` command:
```{r}
head(x = iris)
```
The `data.frame` is probably the most commonly used data type in R. It is basically a table where each column is a variable and each row has one set of values for each of those variables. In the `iris` data, there are five columns: Sepal.Length, Sepal.Width, Petal.Length, Petal.Width, and Species. Each row corresponds to the measurements for an individual flower. Note that all the values in a column of a `data.frame` must be of the same type - if you try to mix numbers and words in the same column, R will "coerce" the data to a single type, which may cause problems for downstream analyses.  

An investigation of our call to the `head` command illustrates two fundamental concepts in R: variables and functions.
```{r, eval = FALSE}
head(x = iris)
```
* `iris` is a variable. That is, it is a name we use to refer to some information in our computer's memory. In this case, the information is a table of flower measurements.
* `head` is the name of the function that prints out the first six rows of a `data.frame`. Most functions require some form of input; in this example, we provided one piece of input to `head`: the name of the variable for which we want the first six lines.  


```{r}
boxplot(Petal.Length ~ Species, data = iris)
```

To keep track of what we do, we will switch from running commands directly in the console to writing R scripts that we can execute. These scripts are just text files with R commands. Create a new script file called 'iris-anova.R'. Now add some key information to the top of the script, using the comment character, #, so R will know to ignore these lines.
```{r}
# ANOVA on iris data set
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2016-09-09
```
Commenting your code is critical in understanding why and how you did analyses when you return to the code two years from now.  
  
The question we will first address is: _are there differences in petal length among the three species?_  
We start by building an analysis of variance model with the `aov` function:
```{r}
aov(formula = Petal.Length ~ Species, data = iris)
```
In this case, we pass _two_ arguments to the `aov` function:

1. For the `formula` parameter, we pass `Petal.Length ~ Species`. This format is used throughout R for describing relationships we are testing. The format is `y ~ x`, where the response variables (e.g. `y`) are to the left of the tilde (~) and the predictor variables (e.g. `x`) are to the right of the tilde. In this example, we are asking if petal length is significantly different among the three species.
2. We also need to tell R where to find the `Petal.Length` and `Species` data, so we pass the name of the `iris data.frame` to the `data` parameter.

But we want to store the model, not just print it to the screen, so we use the assignment operator `<-` to store the product of the `aov` function in a variable of our choice
```{r}
petal.length.aov <- aov(formula = Petal.Length ~ Species, data = iris)
```
Notice how when we execute this command, nothing printed in the console because we instead sent it to a variable. If you just type the variable name,
```{r}
petal.length.aov
```
you will see the familiar output from the `aov` function.  
To see the results of the ANOVA, we call the `summary` function:
```{r}
summary(object = petal.length.aov)
```
The species _do_ have significantly different petal lengths (P < 0.001).

###Necessary?
```{r}
petal.length.anova <- anova(object = petal.length.aov)
petal.length.anova
```

###Student's _t_
So the species are different, but how? To investigate this, we need to perform pairwise comparisons between the species. Make a new R script file called 'iris-t-test.R' and add the header information.
```{r}
# T-test on iris petal lengths
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2016-09-09

# Compare setosa and versicolor
```
We'll start by comparing the data of _Iris setosa_ and _Iris versicolor_, so we need to create two new data objects, one corresponding to the _I. setosa_ data and one for the _I. versicolor_ data.
```{r}
setosa <- iris[iris$Species == "setosa", ]
versicolor <- iris[iris$Species == "versicolor", ]
```
OK, a lot happened with those two lines. Let's take a look:

* `iris` is the `data.frame` we worked with before.
* `iris$Species` refers to one column in `iris`, that is, the column with the name of the species (setosa, versicolor, or virginica).
* The square brackets `[<position 1>, <position 2>]` are used to indicate a subset of the `iris` data. A `data.frame` is effectively a two-dimensional structure - it has some number of rows (the first dimension) and some number of columns (the second dimension). We can see how many rows and columns are in a `data.frame` with the `dim` command. `dim(iris)` prints out the number of rows (`r nrow(iris)`) and the number of columns (`r ncol(iris)`): `r dim(iris)`. We use the square brackets to essentially give an address for the data we are interested in. We tell R which rows we want in the first position and which columns we want in the second position. If a dimension is left blank, then all rows/columns are returned. For example, this would return all columns for the third row of data:
```{r, eval = FALSE}
iris[3, ]
```
So the code 
```{r, eval = FALSE}
setosa <- iris[iris$Species == "setosa", ]
```
will extract all columns (because there is nothing after the comma) in the `iris` data for those rows where the value in the `Species` column is "setosa" _and_ assign that information to a variable called `setosa`.  
  
Comparing the `iris` data and the `setosa` data, we see that there are indeed fewer rows in the `setosa` data:
```{r}
nrow(iris)
nrow(setosa)
```


```{r}

# Use the iris data
# Subset setosa and versicolor data
# Compare Petal.Length of these two species
setosa.v.versicolor <- t.test(x = setosa$Petal.Length, y = versicolor$Petal.Length)
setosa.v.versicolor
```
As a challenge, write the code to test for significant differences in petal lengths between _I. setosa_ and _I. virginica_ and between _I. versicolor_ and _I. virginica_.

###Linear regression
```{r}
# Download gapminder data
# summary() for QA
# Subset for a single year
# Quick plot of data, showing non-linearity
# Transformation of gdpPercap
# Test lifeExp vs. log.gdpPercap
```

###Answer to challenges
####Test for significant differences between _I. setosa_ and _I. virginica_ petal lengths.
```{r}
# Subset setosa data
setosa <- iris[iris$Species == "setosa", ]
# Subset virginica data
virginica <- iris[iris$Species == "virginica", ]
# Run t-test
setosa.v.virginica <- t.test(x = setosa$Petal.Length, y = virginica$Petal.Length)
# Print the results
setosa.v.virginica
```
####Test for significant differences between _I. versicolor_ and _I. virginica_ petal lengths.
```{r}
# Subset versicolor data
versicolor <- iris[iris$Species == "versicolor", ]
# Subset virginica data
virginica <- iris[iris$Species == "virginica", ]
# Run t-test
versicolor.v.virginica <- t.test(x = versicolor$Petal.Length, y = virginica$Petal.Length)
# Print the results
versicolor.v.virginica
```
