---
title: "Introduction to R Statistics"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

An introduction to using the R statistics package and the RStudio interface.

####Learning objectives
1. Read data from files and output results
2. Extract relevant portions of datasets
3. Run standard statistical tests in R, including Student's t, linear regression
and analysis of variance
  
What has to be covered?
* organization data/code/output
* data structures (at least vectors and data.frames)
* passing information to a function

###Setup
First we need to setup our development environment. We need to create two folders: 'data' will store the data we will be analyzing, and 'output' will store the results of our analyses.
```{r, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

###Analysis of Variance (ANOVA)
For our first set of analyses, we'll use a dataset that comes pre-loaded in R. The `iris` data is from early statistical work of R.A. Fisher [doi: 10.1111%2Fj.1469-1809.1936.tb02137.x](https://dx.doi.org/10.1111%2Fj.1469-1809.1936.tb02137.x), who used three species of _Iris_ flowers to develop linear discriminant analysis. To begin, we'll look at the `data.frame` called `iris` using the `head` command:
```{r}
head(x = iris)
```
The `data.frame` is probably the most commonly used data type in R. It is basically a table where each column is a variable and each row has one set of values for each of those variables. In the `iris` data, there are five columns: Sepal.Length, Sepal.Width, Petal.Length, Petal.Width, and Species. Each row corresponds to the measurements for an individual flower. Note that all the values in a column of a `data.frame` must be of the same type - if you try to mix numbers and words in the same column, R will "coerce" the data to a single type, which may cause problems for downstream analyses.  

An investigation of our call to the `head` command illustrates two fundamental concepts in R: variables and functions.
```{r, eval = FALSE}
head(x = iris)
```
* `iris` is a variable. That is, it is a name we use to refer to some information in our computer's memory. In this case, the information is a table of flower measurements.
* `head` is the name of the function that prints out the first six rows of a `data.frame`. Most functions require some form of input; in this example, we provided one piece of input to `head`: the name of the variable for which we want the first six lines.  
  
To keep track of what we do, we will switch from running commands directly in the console to writing R scripts that we can execute. These scripts are just text files with R commands. Create a new script file called 'iris-anova.R'. Now add some key information to the top of the script, using the comment character, #, so R will know to ignore these lines.
```{r}
# ANOVA on iris data set
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2016-09-09
```
Commenting your code is critical in understanding why and how you did analyses when you return to the code two years from now.  
The question we will first address is: _are there differences in petal length among the three species?_  
We start by building an analysis of variance model with the `aov` function:
```{r}
aov(formula = Petal.Length ~ Species, data = iris)
```
But this 

```{r}
petal.length.aov <- aov(formula = Petal.Length ~ Species, data = iris)
```


```{r}
# Run anova on iris data set
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2016-09-09

petal.length.aov <- aov(formula = Petal.Length ~ Species, data = iris)
petal.length.anova <- anova(object = petal.length.aov)
petal.length.anova
```


###Student's _t_
```{r}
# Use the iris data
# Subset setosa and versicolor data
# Compare Petal.Length of these two species
setosa <- iris[iris$Species == "setosa", ]
versicolor <- iris[iris$Species == "versicolor", ]
setosa.v.versicolor <- t.test(x = setosa$Petal.Length, y = versicolor$Petal.Length)
setosa.v.versicolor
```

###Linear regression
```{r}
# Download gapminder data
# Subset for a single year
# Quick plot of data
# Transformation of gdpPercap
# Test lifeExp vs. log.gdpPercap
```

