---
title: "Introduction to Multivariate Statistics in R"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    css: stylesheets/markdown-styles.css
---

## TODO:
1. ~~Add instructions for downloading data then exporting as CSV file~~
2. Break apart PCA code in stepwise fashion
3. ~~Provide rationale for why PCA~~
4. Add (advanced) code for detecting the number of clusters
5. ~~Provide spreadsheet software info (Excel, LibreOffice) on 000-setup-instructions page~~

##See: 

###PCA:

+ http://www.statmethods.net/advstats/factor.html
+ https://www.r-bloggers.com/computing-and-visualizing-pca-in-r/
+ [Why PCA? (or "how to explain PCA to your grandmother")](http://stats.stackexchange.com/questions/2691/making-sense-of-principal-component-analysis-eigenvectors-eigenvalues)
+ [Principal Components Regression](https://www.r-bloggers.com/performing-principal-components-regression-pcr-in-r/)

###Clustering, K-means:

+ https://www.r-bloggers.com/k-means-clustering-in-r/
+ http://www.statmethods.net/advstats/cluster.html
+ https://stat.ethz.ch/R-manual/R-devel/library/stats/html/kmeans.html
+ http://stackoverflow.com/questions/15376075/cluster-analysis-in-r-determine-the-optimal-number-of-clusters
+ [Wikipedia entry for the various ways of determining the optimal number of clusters](https://en.wikipedia.org/wiki/Determining_the_number_of_clusters_in_a_data_set)
+ [Programmatically identifying the "elbow"](http://stackoverflow.com/questions/2018178/finding-the-best-trade-off-point-on-a-curve)

###Ignore these:

+ http://www.statmethods.net/stats/regression.html _ignore_
+ https://www.r-bloggers.com/r-tutorial-series-anova-tables/ _ignore_

An introduction to using the R statistics package and the RStudio interface for multivariate statistics.

####Learning objectives
1. Prepare data in spreadsheet program (e.g. Excel, LibreOffice Calc) for export to R
2. Read data from files into R
3. Run Principal Components Analysis (PCA) and graphically display results
4. Run K-means clustering analysis

##Setup
###Workspace organization
First we need to setup our development environment. We need to create two folders: 'data' will store the data we will be analyzing, and 'output' will store the results of our analyses.
```{r, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

###Preparing data in a format R can read
+ Download data file from [https://jcoliver.github.io/learn-r/data/otter-mandible-data.xlsx](https://jcoliver.github.io/learn-r/data/otter-mandible-data.xlsx) or [http://tinyurl.otter-data](http://tinyurl.otter-data) (the latter just re-directs to the former). These data are a subset of those used in a study on skull morphology and diet specialization in otters [doi: 10.1371/journal.pone.0143236](http://dx.doi.org/10.1371/journal.pone.0143236).
+ Open this file, otter-mandible-data.xlsx, in spreadsheet program like Microsoft Excel<sup>&reg;</sup> or LibreOffice Calc.
+ Save a copy of the file as a CSV (comma-separated values) file named 'otter-mandible-data.csv' in the data folder you created above:
    + In MS Excel<sup>&reg;</sup>, select File > Save As... and in the dialog that appears, select CSV from the type dropdown menu.
    + In LibreOffice Calc, select File > Save As... and in the dialog that appears, select Text CSV (.csv) in the Format dropdown in the lower-right portion of the dialog.

###Reading data into R
```{r}
otter <- read.csv(file = "data/otter-mandible-data.csv", header = TRUE)
```
  
Missing data can cause problems in downstream analyses, so we will just remove any rows that have missing data. Here we replace the original data object `otter` with one in which there are no missing values. Note, this _does not_ alter the data in the original file we read into R; it only alters the data object `otter` currently in R's memory.
```{r}
otter <- na.omit(otter)
```

##PCA
+ Do pca
+ point out
    + summary for the proportion of variance (remember if you want to show how this is calculated, the standard deviation has to be squared to get variance)
    + loadings for what points along an axis actually mean
+ do call to biplot (ugly)?
+ do plot from fit$x
  + make plot nicer

Why PCA? Very briefly, Principal Components Analysis is a way of re-describing the variation observed in your data. It serves as a means of reducing the dimensionality of data (i.e. reducing the number of predictor variables) and is often used for exploratory analyses. The full rationale and mathematically underpinnings are waaaaaaaay beyond the scope of this lesson, and other resources already do a fairly good job of explaining PCA. If you want a few perspectives for a relatively novice audience, check out this [Why PCA? (or "how to explain PCA to your grandmother")](http://stats.stackexchange.com/questions/2691/making-sense-of-principal-component-analysis-eigenvectors-eigenvalues) thread at Stack Overflow. If you are more inclined to print media, I highly recommend B.F.J. Manly's _Multivariate Statistical Methods: A primer_ [*TODO: REF*], which provides an excellent introduction to a variety of multivariate statistic topics.
  
So, on to the code:
```{r}
pca.fit <- prcomp(x = otter[, -1], scale. = TRUE)
```
That's PCA. We pass the data to the `x` parameter, skipping the first column `[, -1]` because that has the name of the species the data came from (and is not something we can analyze with the PCA). We also set the `scale.` parameter to `TRUE` because we want to transform the data so each column has a mean of zero and a variance of one.
  
To look at the results, we use the `summary` command and assign the output to a variable.
```{r}
pca.summary <- summary(pca.fit)
ls(pca.summary) # Lists the objects produced by summary
```

We are interested to know (1) what are the important factors that emerge from the PCA (i.e. which ones explain a lot of variation) and (2) what do these factors actually say about the variation observed in our data. For (1), look at the `importance` object in the summary:
```{r}
pca.summary$importance
```
The second row, `Proportion of Variance`, shows how much variation in the data is described by each component; notice that the first component, PC1, explains the most variance, `r round(pca.summary$sdev[1]^2/sum(pca.summary$sdev^2), 4)`, or `r round(pca.summary$sdev[1]^2/sum(pca.summary$sdev^2) * 100, 2)`% of the total variance, the second components explains the second most variance (`r round(pca.summary$sdev[2]^2/sum(pca.summary$sdev^2) * 100, 2)`%), and so on, with each successive component explaining a lower proportion of the total variance. For the remainder of the lesson, we will focus on the first two principal components, which together explain `r round((pca.summary$sdev[1]^2 + pca.summary$sdev[2]^2)/sum(pca.summary$sdev^2) * 100, 2)`% of the observed variation in the skull measurements.
  
But what about that variation? What are the principal components actually explaining? To address this (point 2 from above), we need to look at the loadings of the PCA. The `rotation` object from the summary call has the information we are interested in. Focus on the values in the PC1 column:
```{r}
pca.summary$rotation
```

```{r echo = FALSE}
# Crude, but should be fine
direction <- "positive"
if (sum(pca.summary$rotation) < 0) {
  direction <- "negative" 
}
```
Looking at the signs of the loadings, we see they are all the same (`r direction`), thus this first component, explaining most of the variation in the measurements, is really just reflecting variation in size. That is, since all the loadings have the same sign, large values for one skull measurement generally coincide with large values for other skull measurements for this first component.
  
The second principal component is a little more interesting. Two of the variables, mandibular ramus width and outlever at carnassial don't contribute much to the component (the magnitudes of their respective loadings (.173 and .031) are small compared to the other four skull measurements). The remaining four indicate a shape difference, with one pair of variables having positive loadings and one pair having negative loadings. This interpretation of the second principal component would benefit greatly from a graphical representation.
  
Plotting the results of a PCA can be done using a simple call to the `biplot` function:
```{r}
biplot(x = pca.fit)
```

But that figure really leaves much to be desired, and gets messier with larger sample sizes and more variables. If you want to find out more about how that figure is produced, look at the documentation for `biplot` (`?biplot`).

Instead, we can plot the scores of the first two principal components using the standard `plot` command, using the scores that are stored in the `x` object of `pca.fit`:
```{r}
plot(x = pca.fit$x[, 1],
     y = pca.fit$x[, 2],
     xlab = "PC 1",
     ylab = "PC 2")
```

Well, maybe that plot isn't so useful either. It does help a bit if we color the points by species, though.
```{r}
# Pull out the unique values in the 'species' column
species <- unique(otter$species)
# Set up a vector of colors for the legend
legend.cols <- c("black", "forestgreen", "cadetblue", "darkred") 
################################################################################
# Set up a vector of colors for the actual plot, based on values in the 
# 'species' column and the legend colors vector. This vector has one element 
# corresponding to each row of the otter data frame. 
pt.cols <- rep(x = legend.cols[1], length = nrow(otter))
pt.cols[otter$species == species[2]] <- legend.cols[2]
pt.cols[otter$species == species[3]] <- legend.cols[3]
pt.cols[otter$species == species[4]] <- legend.cols[4]
plot(x = pca.fit$x[, 1],
     y = pca.fit$x[, 2],
     xlab = "PC 1",
     ylab = "PC 2",
     pch = 19,
     col = pt.cols)
legend("bottomleft", legend = species, pch = 19, col = legend.cols, cex = 0.8)
```

Ah, now this is a little more interesting. Note the first principal component scores along the x-axis, and there is actually pretty clear separation among some of species. As mentioned above, this principal component is really an index of size, which we can visualize by looking at the actual skull measurements across the four species

```{r}
par(mfrow = c(2, 3), las = 2) # To set-up a multi-panel graph (2 rows, 3 columns), filling row by row
boxplot(formula = mandibular.ramus.width ~ species, data = otter)
boxplot(formula = mandibular.ramus.height ~ species, data = otter)
boxplot(formula = moment.arm.temporalis ~ species, data = otter)
boxplot(formula = outlever.at.carnassial ~ species, data = otter)
boxplot(formula = moment.arm.masseter ~ species, data = otter)
boxplot(formula = jaw.length ~ species, data = otter)
par(mfrow = c(1, 1), las = 0) # To restore graphing defaults
```


```{r}
boxplot(formula = jaw.length ~ species, data = otter)
boxplot(formula = mandibular.ramus.height ~ species, data = otter)
```

Our script should look like this:
```{r}
```

***

##Clustering with K-means

```{r}
# Scale data & drop species column
otter.scaled <- scale(x = otter[, -1])
wss <- numeric(10)
for (i in 1:10) {
  kmeans.fit <- kmeans(x = otter.scaled, centers = i, nstart = 10)
  wss[i] <- kmeans.fit$tot.withinss
}
plot(x = c(1:length(wss)), y = wss, xlab = "Number of clusters, K", ylab = "Total within SS", type = "b")

# Percentage of variance explained is the ratio of the between-group variance to the total 
# variance, also known as an F-test. A slight variation of this method plots the curvature 
# of the within group variance. https://en.wikipedia.org/wiki/Determining_the_number_of_clusters_in_a_data_set
# 

```

Abstracting the number of clusters to analyze
```{r}
```

Apply "Elbow" method based on percentage of variance explained
```{r}
# Advanced
```

See how clustering aligned with species
```{r}
# Run kmeans again for K = 4, to get clustering info
otter.k.4 <- kmeans(x = otter.scaled, centers = 4, nstart = 10)
otter.clusters <- data.frame(otter, cluster = otter.k.4$cluster)
assignments <- table(otter.clusters$species, otter.clusters$cluster)
assignments

```

***

###References

See: [Bibliographies in RStudio](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html)

***

Questions?  e-mail me at <a href="mailto:jcoliver@email.arizona.edu">jcoliver@email.arizona.edu</a>.