---
title: "Introduction to Multivariate Statistics in R"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    css: stylesheets/markdown-styles.css
---

##See: 

###PCA:

+ http://www.statmethods.net/advstats/factor.html
+ https://www.r-bloggers.com/computing-and-visualizing-pca-in-r/

###Clustering, K-means:

+ https://www.r-bloggers.com/k-means-clustering-in-r/
+ http://www.statmethods.net/advstats/cluster.html
+ https://stat.ethz.ch/R-manual/R-devel/library/stats/html/kmeans.html
+ http://stackoverflow.com/questions/15376075/cluster-analysis-in-r-determine-the-optimal-number-of-clusters

###Ignore these:

+ http://www.statmethods.net/stats/regression.html _ignore_
+ https://www.r-bloggers.com/r-tutorial-series-anova-tables/ _ignore_

An introduction to using the R statistics package and the RStudio interface for multivariate statistics.

####Learning objectives
1. Prepare data in spreadsheet program (e.g. Excel, LibreOffice Calc) for export to R
2. Read data from files into R
3. Run Principal Components Analysis (PCA) and graphically display results
4. Run K-means clustering analysis
5. Run DFA?

##Setup
First we need to setup our development environment. We need to create two folders: 'data' will store the data we will be analyzing, and 'output' will store the results of our analyses.
```{r, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

##Preparing data in external source

***

##Reading data into R

***

##Clustering with K-means

```{r}
# Scale data & drop species column
iris.scaled <- scale(x = iris[, -5])

# Try 1-8 clusters to find "elbow" in sum of squares
wss <- numeric(10)
for (i in 1:10) {
  iris.fit <- kmeans(x = iris.scaled, centers = i, nstart = 10)
  wss[i] <- iris.fit$tot.withinss
}
plot(x = c(1:10), y = wss, xlab = "Number of clusters, K", ylab = "Total within SS", type = "b")

```

```{r}
# Stolen from: http://www.statmethods.net/advstats/cluster.html
# K-Means Cluster Analysis
fit <- kmeans(mydata, 5) # 5 cluster solution
# get cluster means
aggregate(mydata,by=list(fit$cluster),FUN=mean)
# append cluster assignment
mydata <- data.frame(mydata, fit$cluster)
```

***

###References

See: [Bibliographies in RStudio](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html)

***

Questions?  e-mail me at <a href="mailto:jcoliver@email.arizona.edu">jcoliver@email.arizona.edu</a>.