---
title: "Introduction to Multivariate Statistics in R"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    css: stylesheets/markdown-styles.css
---

##See: 

###PCA:

+ http://www.statmethods.net/advstats/factor.html
+ https://www.r-bloggers.com/computing-and-visualizing-pca-in-r/

###Clustering, K-means:

+ https://www.r-bloggers.com/k-means-clustering-in-r/
+ http://www.statmethods.net/advstats/cluster.html
+ https://stat.ethz.ch/R-manual/R-devel/library/stats/html/kmeans.html
+ http://stackoverflow.com/questions/15376075/cluster-analysis-in-r-determine-the-optimal-number-of-clusters
+ [Wikipedia entry for the various ways of determining the optimal number of clusters](https://en.wikipedia.org/wiki/Determining_the_number_of_clusters_in_a_data_set)
+ [Programmatically identifying the "elbow"](http://stackoverflow.com/questions/2018178/finding-the-best-trade-off-point-on-a-curve)

###Ignore these:

+ http://www.statmethods.net/stats/regression.html _ignore_
+ https://www.r-bloggers.com/r-tutorial-series-anova-tables/ _ignore_

An introduction to using the R statistics package and the RStudio interface for multivariate statistics.

####Learning objectives
1. Prepare data in spreadsheet program (e.g. Excel, LibreOffice Calc) for export to R
2. Read data from files into R
3. Run Principal Components Analysis (PCA) and graphically display results
4. Run K-means clustering analysis
5. Run DFA?

##Setup
First we need to setup our development environment. We need to create two folders: 'data' will store the data we will be analyzing, and 'output' will store the results of our analyses.
```{r, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

##Preparing data in external source

***

##Reading data into R

***

##PCA

from: http://www.statmethods.net/advstats/factor.html
fit <- princomp(mydata, cor=TRUE)
summary(fit) # print variance accounted for
loadings(fit) # pc loadings
plot(fit,type="lines") # scree plot
fit$scores # the principal components
biplot(fit) 

```{r}
fit <- princomp(iris[, -5], cor = TRUE)
summary(fit)
loadings(fit)
plot(fit, type = "lines")
fit$scores
biplot(x = fit)
```

+ Standardize variables, mean = 0, variance = 1
+ eigenvalues are variance by each component
  + Get % variance for i^th^ component via eig~i~/n, where n is the total number of components (total number of variables)
+ Look at loadings to explain variation in data

***

##Clustering with K-means

```{r}
# Scale data & drop species column
iris.scaled <- scale(x = iris[, -5])

# Try 1-8 clusters to find "elbow" in sum of squares
wss <- numeric(10)
for (i in 1:10) {
  iris.fit <- kmeans(x = iris.scaled, centers = i, nstart = 10)
  wss[i] <- iris.fit$tot.withinss
}
plot(x = c(1:10), y = wss, xlab = "Number of clusters, K", ylab = "Total within SS", type = "b")

# Percentage of variance explained is the ratio of the between-group variance to the total 
# variance, also known as an F-test. A slight variation of this method plots the curvature 
# of the within group variance. https://en.wikipedia.org/wiki/Determining_the_number_of_clusters_in_a_data_set
# 

```

Abstracting the number of clusters to analyze
```{r}
# Scale data & drop species column
iris.scaled <- scale(x = iris[, -5])
k.max <- 10

# Try 1-8 clusters to find "elbow" in sum of squares
wss <- numeric(k.max)
for (i in 1:k.max) {
  iris.fit <- kmeans(x = iris.scaled, centers = i, nstart = 10)
  wss[i] <- iris.fit$tot.withinss
}
plot(x = c(1:k.max), y = wss, xlab = "Number of clusters, K", ylab = "Total within SS", type = "b")
```

Apply "Elbow" method based on percentage of variance explained
```{r}
# Advanced
```

See how clustering aligned with species
```{r}
# Run kmeans again for K = 3, to get clustering info
iris.k.3 <- kmeans(x = iris.scaled, centers = 3, nstart = 10)

iris.clusters <- data.frame(iris, cluster = iris.k.3$cluster)
assignments <- table(iris.clusters$Species, iris.clusters$cluster)
assignments
```

***

###References

See: [Bibliographies in RStudio](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html)

***

Questions?  e-mail me at <a href="mailto:jcoliver@email.arizona.edu">jcoliver@email.arizona.edu</a>.