---
title: "Introduction to Multivariate Statistics in R"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    css: stylesheets/markdown-styles.css
---

##See: 

###PCA:

+ http://www.statmethods.net/advstats/factor.html
+ https://www.r-bloggers.com/computing-and-visualizing-pca-in-r/

###Clustering, K-means:

+ https://www.r-bloggers.com/k-means-clustering-in-r/
+ http://www.statmethods.net/advstats/cluster.html
+ https://stat.ethz.ch/R-manual/R-devel/library/stats/html/kmeans.html

###Ignore these:

+ http://www.statmethods.net/stats/regression.html _ignore_
+ https://www.r-bloggers.com/r-tutorial-series-anova-tables/ _ignore_

An introduction to using the R statistics package and the RStudio interface for multivariate statistics.

####Learning objectives {.faded}
1. Prepare data in spreadsheet program (e.g. Excel, LibreOffice Calc) for export to R
2. Read data from files into R
3. Run Principal Components Analysis (PCA) and graphically display results
4. Run 

##Setup
First we need to setup our development environment. We need to create two folders: 'data' will store the data we will be analyzing, and 'output' will store the results of our analyses.
```{r, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

##Preparing data in external source

***

##Reading data into R

***

##Multivariate linear regression

```{r}
gapminder <- read.csv(file = "data/gapminder-FiveYearData.csv")
gapminder$logGDP <- log10(gapminder$gdpPercap)
fit.one <- lm(formula = lifeExp ~ logGDP, data = gapminder)
fit.two <- lm(formula = lifeExp ~ year + logGDP, data = gapminder)
anova(fit.one, fit.two)
```

***

##Multivariate analysis with categorical predictors

***

Questions?  e-mail me at <a href="mailto:jcoliver@email.arizona.edu">jcoliver@email.arizona.edu</a>.