---
title: "Introduction to data visualization with ggplot"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    css: stylesheets/markdown-styles.css
---

An introduction to using the R statistics package and the RStudio interface for multivariate statistics.

#### Learning objectives
1. Install and use third-party packages for R
2. Use layering to add elements to a plot
3. Format plots with faceting

## Setup
### Workspace organization
First we need to setup our development environment. We need to create two folders: 'data' will store the data we will be analyzing, and 'output' will store the results of our analyses.
```{r, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

With this workspace organization, we can download the data, either manually or use R to automatically download it. For this lesson we'll do the latter, saving the file in the `data` directory we just created, and name the file `gapminder.csv`:

```{r, eval = FALSE}
download.file(url = "http://tinyurl.com/qb83k3z", destfile = "data/gapminder.csv")
```

From this point on, we want to keep track of what we have done, so we will restrict our use of the console and instead use script files. Start a new script with some brief header information at the very top. We want, at the very least, to include:
1. A short description of what the script does (no more than one line of text)
2. Your name
3. A means of contacting you (e.g. your e-mail address)
4. The date, preferably in ISO format YYYY-MM-DD
```{r}
# Plot gapminder data
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-02-23
```

### Installing additional packages
There are two steps to using additional packages in R:
1. Install the package through `install.packages()`
2. Load the package into active memory with `library()`

For this exercise, we will install the `ggplot2` package:

```{r eval = FALSE}
install.packages("ggplot2")
library("ggplot2")
```

It is important to note that for each computer you work on, the `install.packages` command need only be issued once, while the call to `library` will need to be issued for each session of R. Because of this, it is standard convention in scripts to comment out the `install.packages` command once it has been run on the machine, so our script now looks like:

```{r}
# Plot gapminder data
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-02-23

#install.packages("ggplot2")
library("ggplot2")
```

## Plot!

***

### Scatterplot
1. lifeExp vs. gdp
2. lifeExp vs. log(gdp)
3. lifeExp vs. log(gdp) color by continent
4. lifeExp vs. log(gdp) color by continent with custom colors
5. ggsave

```{r}
#install.packages("ggplot2")
library("ggplot2")

# Load data
gapminder <- read.csv(file = "data/gapminder.csv")

# Create plot object
lifeExp.plot <- ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  scale_color_manual(values = c("red", "orange", "forestgreen", "darkblue", "violet"))

print(lifeExp.plot)  
```
```{r, eval = FALSE}
ggsave(filename = "output/gdp-lifeExp-plot.png", plot = lifeExp.plot)
```


***

### Violin plot
1. Y: gdp, X: continent
2. Y: log(gdp), X: continent
3. Y: log(gdp), X: continent, color with scale_fill_gradient for log(gdp)
4. Add theme_bw()
ggsave

***

## Additional resources

***

<a href="index.html">Back to learn-r main page</a>
  
Questions?  e-mail me at <a href="mailto:jcoliver@email.arizona.edu">jcoliver@email.arizona.edu</a>.