---
title: "Introduction to data visualization with ggplot"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    css: stylesheets/markdown-styles.css
---

An introduction to using the R statistics package and the RStudio interface for multivariate statistics.

#### Learning objectives
1. Install and use third-party packages for R
2. Use layering to add elements to a plot
3. Format plots with faceting

## Setup
### Workspace organization
First we need to setup our development environment. We need to create two folders: 'data' will store the data we will be analyzing, and 'output' will store the results of our analyses.
```{r, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

From this point on, we want to keep track of what we have done, so we will restrict our use of the console and instead use script files. Start a new script with some brief header information at the very top. We want, at the very least, to include:
1. A short description of what the script does (no more than one line of text)
2. Your name
3. A means of contacting you (e.g. your e-mail address)
4. The date, preferably in ISO format YYYY-MM-DD
```{r}
# Plot diamond data
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-02-23
```

### Installing additional packages

## Plot!

***

## Additional resources

***

### Installing R packages
Note that if a package is not installed on your machine, a call to `library` will throw an error, indicating the package has to be installed first. For example, if we were to load a package called `mustelid` without installing it first:
```{r error = TRUE}
library("mustelid")
```
So how do I install packages? you ask. With `install.packages`:
```{r eval = FALSE}
install.packages("mustelid")
library("mustelid")
```

<a href="index.html">Back to learn-r main page</a>
  
Questions?  e-mail me at <a href="mailto:jcoliver@email.arizona.edu">jcoliver@email.arizona.edu</a>.