---
title: "Introduction to data visualization with ggplot"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    css: stylesheets/markdown-styles.css
---

An introduction to using the R statistics package and the RStudio interface for multivariate statistics.

#### Learning objectives
1. Install and use third-party packages for R
2. Use layering to add elements to a plot
3. Format plots with faceting

## Setup
### Workspace organization
First we need to setup our development environment. We need to create two folders: 'data' will store the data we will be analyzing, and 'output' will store the results of our analyses.
```{r, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

With this workspace organization, we can download the data, either manually or use R to automatically download it. For this lesson we'll do the latter, saving the file in the `data` directory we just created, and name the file `gapminder.csv`:

```{r, eval = FALSE}
download.file(url = "http://tinyurl.com/qb83k3z", destfile = "data/gapminder.csv")
```

From this point on, we want to keep track of what we have done, so we will restrict our use of the console and instead use script files. Start a new script with some brief header information at the very top. We want, at the very least, to include:
1. A short description of what the script does (no more than one line of text)
2. Your name
3. A means of contacting you (e.g. your e-mail address)
4. The date, preferably in ISO format YYYY-MM-DD
```{r}
# Plot gapminder data
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-02-23
```

### Installing additional packages
There are two steps to using additional packages in R:
1. Install the package through `install.packages()`
2. Load the package into active memory with `library()`

For this exercise, we will install the `ggplot2` package:

```{r eval = FALSE}
install.packages("ggplot2")
library("ggplot2")
```

It is important to note that for each computer you work on, the `install.packages` command need only be issued once, while the call to `library` will need to be issued for each session of R. Because of this, it is standard convention in scripts to comment out the `install.packages` command once it has been run on the machine, so our script now looks like:

```{r}
# Plot gapminder data
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-02-23

#install.packages("ggplot2")
library("ggplot2")
```

## Plot!

***

```{r}
#install.packages("ggplot2")
library("ggplot2")

# Load data
gapminder <- read.csv(file = "data/gapminder.csv")
```

### Scatterplot
1. lifeExp vs. gdp
2. lifeExp vs. log(gdp)
3. lifeExp vs. log(gdp) color by continent
4. lifeExp vs. log(gdp) color by continent with custom colors
5. ggsave

One thing of interest is to include additional information in the plot, such as which continent each point comes from. We can color points by another value in our data through the `aes` parameter in the initial call to `ggplot`. So in addition to telling R which data to use for the x and y axes, we indicate which data to use for point colors:

```{r}
# Create plot object
lifeExp.plot <- ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  scale_x_log10()

# Draw plot
print(lifeExp.plot)  
```

At this point, we want to change the colors of the points in two ways: (1) we want to make them slightly transparent because there is considerable overlap in some part of the graph; we do this by setting the `alpha` parameter in the `geom_point` call; (2) we want to use custom colors for each continent; this is done with the `scale_color_manual` function:

```{r}
# Create plot object
lifeExp.plot <- ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  scale_color_manual(values = c("red", "orange", "forestgreen", "darkblue", "violet"))

# Draw plot
print(lifeExp.plot)  
```

And if we want to save the plot to a file, we call `ggsave`, passing a filename and the plot object we want to save (the latter is optional, if we don't indicate which plot to save, `ggsave` will save the last plot that was displayed):

```{r, eval = FALSE}
# Save plot to png file
ggsave(filename = "output/gdp-lifeExp-plot.png", plot = lifeExp.plot)
```

Also note that `ggsave` will guess the type of file to save from the extension. For example, if we instead wanted to save a TIF instead of a PNG, we change the extension for the `filename` argument:

```{r, eval = FALSE}
# Save plot to tif file
ggsave(filename = "output/gdp-lifeExp-plot.tif", plot = lifeExp.plot)
```

Our final script for this scatterplot is then:

```{r, eval = FALSE}
# Plot gapminder data
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-02-23

#install.packages("ggplot2")
library("ggplot2")

# Load data
gapminder <- read.csv(file = "data/gapminder.csv")

# Create plot object
lifeExp.plot <- ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  scale_color_manual(values = c("red", "orange", "forestgreen", "darkblue", "violet"))

# Draw plot
print(lifeExp.plot)  

# Save plot to png file
ggsave(filename = "output/gdp-lifeExp-plot.png", plot = lifeExp.plot)
```

***

### Violin plot
1. Y: gdp, X: continent
2. Y: log(gdp), X: continent
3. Y: log(gdp), X: continent, color with scale_fill_gradient for log(gdp)
4. Add theme_bw()
ggsave

***

## Additional resources

***

<a href="index.html">Back to learn-r main page</a>
  
Questions?  e-mail me at <a href="mailto:jcoliver@email.arizona.edu">jcoliver@email.arizona.edu</a>.