---
title: "Heatmaps in R"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
urlcolor: blue
---

A worked example of making heatmaps in R with the `ggplot` package, as well as 
some data wrangling to easily format the data needed for the plot.

#### Learning objectives

1. Manipulate data into a 'tidy' format
2. Visualize data in a heatmap
3. Become familiar with `ggplot` syntax for customizing plots

## Heatmaps & data wrangling

Heatmaps are a great way of displaying three-dimensional data in only two 
dimensions. But how can we easily translate tabular data into a format for 
heatmap plotting? By taking advantage of "data munging" and graphics packages, 
heatmaps are relatively easy to produce in R. 

***

## Getting started

First we need to setup our development environment. Open RStudio and create a 
new project via:

+ File > New Project...
+ Select 'New Directory'
+ For the Project Type select 'New Project'
+ For Directory name, call it something like "r-heatmaps" (without the quotes)
+ For the subdirectory, select somewhere you will remember (like "My Documents" 
or "Desktop")

Next create the two folders we'll use to organize our efforts. The two folders 
should be `data` and `output` and will store...data and output.

```{r setup-folders, eval = FALSE}
dir.create("data")
dir.create("output")
```

For this lesson we will use data from a diversity survey of microbial life in a 
[chrysotile asbestos](https://en.wikipedia.org/wiki/Chrysotile) mine. Driscoll 
and colleagues 
[doi: 10.1016/j.gdata.2016.11.004](https://doi.org/10.1016/j.gdata.2016.11.004) 
used next generation sequencing to identify and categorize bacterial diversity 
in a flooded pit of the abandoned mine. The data are normalized read counts for 
microbes found at the sites. A subset of the data can be downloaded from
[https://tinyurl.com/mine-data-csv](https://tinyurl.com/mine-data-csv) (which 
itself redirects to [https://jcoliver.github.io/learn-r/data/mine-microbe-class-data.csv](https://jcoliver.github.io/learn-r/data/mine-microbe-class-data.csv)). 
You can download this file to your computer with the `download.file` function:

```{r download-data, eval = FALSE}
download.file(url = "https://tinyurl.com/mine-data-csv", 
              destfile = "data/mine-data.csv")
```

This downloads the data and saves it in the `data` directory as `mine-data.csv`.  

***

## Data Wrangling

We'll need to start by reading the data into memory then formatting it for use 
by the ggplot package. We want all our work to be reproducible, so create a 
script where we can store all the commands we use to create the heatmap. We 
begin this script with brief information about the purpose of the script:

```{r script-header, eval = FALSE}
# Heatmap of mine pit microbe diversity
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-06-05
```

Now we read those abundance data into memory:

```{r read-data}
mine.data <- read.csv(file = "data/mine-data.csv")
```

Take a quick look at the structure of the data, using the `str` command:

```{r data-structure}
str(mine.data)
```

The data frame has `r nrow(mine.data)` rows ("obs.") and `r ncol(mine.data)` 
columns ("variables"). The first three columns have information about the 
observation (`r colnames(mine.data)[1:3]`), and the remaining columns have the 
abundance for each of `r ncol(mine.data) - 3` classes of bacteria.

We ultimately want a heatmap where the different sites are shown along the 
x-axis, the classes of bacteria are shown along the y-axis, and the shading of 
the cell reflects the abundance. This latter value, the abundance of a bacteria 
at specific depths at different site, is really just a third dimension. 
However, instead of creating a 3-dimensional plot that can be difficult to 
visualize, we instead use shading for our "z-axis". To this end, we need our 
data formatted so we have a column corresponding to each of these three 
dimensions:

+ X: Sample identity
+ Y: Bacterial class
+ Z: Abundance

The challenge is that our data are _not_ formatted like this. While the 
`Sample.name` column corresponds to what we would like for our x-axis, we do 
not have columns that correspond to what is needed for the y- and z-axes. All 
the data are in our data frame, but we need to take a table that looks like 
this:

| Site.name | Actinobacteria | Cytophagia |
|:----------|:---------------|:-----------|
| 1-S       | 373871         | 8052       |
| 2-S       | 332856         | 28561      |
| ...       | ...            | ...        |

And transform it to one with a column for bacterial class and a column for 
abundance, like this:

| Site.name | Class          | Abundance |
|:----------|:---------------|:----------|
| 1-S       | Actinobacteria | 373871    |
| 1-S       | Cytophagia     | 8052      |
| 2-S       | Actinobacteria | 332856    |
| 2-S       | Cytophagia     | 28561     |
| ...       | ...            | ...       |

Thankfully, there is an R package that does this for us! The `tidyr` package is 
designed for creating this type of "tidy"" data. 

```{r install-tidyr, eval = FALSE}
install.packages("tidyr")
```

```{r load-tidyr}
library("tidyr")
```

In fact, we can do this transformation in one line with the `pivot_longer` 
function:

```{r pivot-longer-01, error = TRUE}
mine.long <- pivot_longer(data = mine.data, 
                          cols = everything(),
                          names_to = "Class", 
                          values_to = "Abundance")
```

Uh oh. That error is preventing us from transforming our data to long format. 
What R is telling us is that we tried to create a column with multiple data 
types in it, which is a big no-no in R. That is, it attempted to make a single 
column that had both the data from the "Site" column (which are numbers) with 
the data from the "Sample.name" column, which is a factor (factors are what R 
calls categories).

Consider our original data structure:

```{r structure-long}
str(mine.data)
```

Notice there are columns that don't need to be transformed for our heatmap. 
Namely, Site, Depth, and Sample.name can all remain as-is. To do this, we 
instruct `pivot_longer` to ignore the first three columns of the data frame by 
negating those columns (`-c(1:3`):

```{r pivot-longer-02}
mine.long <- pivot_longer(data = mine.data, 
                          cols = -c(1:3), 
                          names_to = "Class", 
                          values_to = "Abundance")
head(mine.long)
```

Another approach, to accomplish the same thing, especially if the columns to 
ignore are not adjacent, is to explicitly name the columns to ignore:

```{r pivot-longer-03, eval = FALSE}
mine.long <- pivot_longer(data = mine.data, 
                    cols = -c(Site, Depth, Sample.name), 
                    names_to = "Class", 
                    values_to = "Abundance")
```

To recap, at this point we read in the data and transformed it for easy use 
with heatmap tools:

```{r status-01, eval = FALSE}
# Heatmap of mine pit microbe diversity
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-06-05

# Load dependancies
library("tidyr")

# Read data and format for heatmap
mine.data <- read.csv(file = "data/mine-data.csv")
mine.long <- pivot_longer(data = mine.data, 
                    cols = -c(1:3), 
                    names_to = "Class", 
                    values_to = "Abundance")
```

Now the data are ready - on to the plot!

***

## Plotting heatmaps

### The ggplot package

To plot a heatmap, we are going to use the `ggplot2` package.

```{r install-ggplot2, eval = FALSE}
install.packages("ggplot2")
```

```{r load-ggplot2}
library("ggplot2")
```

For this plot, we are going to first create the heatmap object with the 
`ggplot` function, then print the plot. We create the object by assigning the 
output of the `ggplot` call to the variable `mine.heatmap`, then entering the 
name of this object to print it to the screen.

```{r heatmap-basic}
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Abundance)) +
  geom_tile() +
  xlab(label = "Sample")

mine.heatmap
```

Let's dissect these commands:

+ `ggplot` is the initial call, where we provide the following information:
    + `data` is the data frame that contains the data we want to plot
    + `mapping` tells ggplot what to plot where; that is, in this call, it says 
    we want the `Sample.name` column on the x-axis, the bacterial `Class` on 
    the y-axis, and the shading, or fill, (the z-axis) to reflect the value in 
    the `Abundance` column.
+ `geom_tile`, which is appended to `ggplot` with a plus sign (+), tells ggplot 
that we want a heatmap
+ `xlab` provides a label to use for the x-axis

This command follows the commonly used syntax for creating ggplot objects, 
where an initial call to `ggplot` sets up preliminary plotting information, and 
details are appended with plus signs:

```{r ggplot-syntax, eval = FALSE}
# Create plot object
plot.object.name <- ggplot(data, mapping) +
  layer.one() +
  layer.two() +
  layer.three()

# Draw plot
plot.object.name
```

### Faceting a plot

The ggplot package includes a great way to visualize different categories of 
data. This approach, called "faceting" requires one additional layer to the 
plotting command called `facet_grid`. With `facet_grid` we indicate which 
column contains the categories we want to use for the plot:

```{r facet}
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Abundance)) +
  geom_tile() +
  xlab(label = "Sample") +
  facet_grid(~ Depth)

mine.heatmap
```

But this leaves a bit to be desired. We only want columns displayed for which 
there are data, so we have to add some additional information in our 
`facet_grid` call:

```{r facet-free}
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Abundance)) +
  geom_tile() +
  xlab(label = "Sample") +
  facet_grid(~ Depth, scales = "free_x", space = "free_x")

mine.heatmap
```

+ Passing `"free_x"` to the `scales` argument instructs R to allow different 
x-axis for each of our sub-plots. In this case, the default is for each site to 
appear in each sub-plot. Specifying `scales = "free_x"` removes any site from a 
sub-plot where there is no corresponding data. That is, it removes sites 1-S, 
2-S, and 3-S from the 3.5 and 25 depth sub-plots, removes 1-M, 2-M, and 3-M 
from the 0.5 and 25 depth sub-plots, and removes 2-B and 3-B from the 0.5 and 
3.5 depth sub-plots.
+ Passing `"free_x"` to `space` ensures that each column in the plot is the 
same width. Try it without specifying the `space` argument to see the effect of 
this behavior.

While we are here, we can also move the boxes labelling each sub-plot (0.5, 
3.5, and 25) from the top of the plot to the bottom of the plot with the 
`switch` argument in the `facet_grid()` function:

```{r facet-switch}
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Abundance)) +
  geom_tile() +
  xlab(label = "Sample") +
  facet_grid(~ Depth, switch = "x", scales = "free_x", space = "free_x")

mine.heatmap
```


That's better. There is still some room for improvement, but we'll get to that 
later.

### Improving aesthetics

#### Scale

First let's deal with a scaling issue. Most values are dark blue, but this is 
partly caused by the distribution of values - there are a few really high 
values in the Abundance column. We can transform the data for display by taking 
the square root of abundance and plotting that. To do this, we need to add 
another column to our data frame and update our call to ggplot to reference 
this new column, `Sqrt.abundance`:

```{r sqrt-transform}
mine.long$Sqrt.abundance <- sqrt(mine.long$Abundance)
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Sqrt.abundance)) +
  geom_tile() +
  xlab(label = "Sample") +
  facet_grid(~ Depth, switch = "x", scales = "free_x", space = "free_x")

mine.heatmap
```

#### Color

What if we want to use a different set of colors besides the default? For 
example, how can we change the shading so low values are light and high values 
are dark? We change colours with `scale_fill_gradient`. In this case we pass 
three objects to `scale_fill_gradient`: `name` is a label to use in the legend 
and `low` and `high` are hex codes for color. In this case, we use white 
(#FFFFFF) for the low values and a dark blue (#012345) for high values.

```{r change-color}
mine.long$Sqrt.abundance <- sqrt(mine.long$Abundance)
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Sqrt.abundance)) +
  geom_tile() +
  xlab(label = "Sample") +
  facet_grid(~ Depth, switch = "x", scales = "free_x", space = "free_x") +
  scale_fill_gradient(name = "Sqrt(Abundance)",
                      low = "#FFFFFF",
                      high = "#012345")

mine.heatmap
```

#### Theme elements

A number of elements of the plot are controlled through the `theme` function of 
ggplot. We can use this function to alter axis titles and the placement of the 
facet titles. Remember the facets? We grouped the graphs based on the depth of 
the samples. The graph currently shows the depth labels (the boxes with 0.5, 
3.5, or 25) between the graph and the sample name. We'd like to swap these 
positions, so the categories are shown at the very bottom. While we're at it, 
we should update the x-axis label, too. Setting `strip.placement` to "outside" 
should accomplish this.

```{r theme-elements}
mine.long$Sqrt.abundance <- sqrt(mine.long$Abundance)
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Sqrt.abundance)) +
  geom_tile() +
  xlab(label = "Depth (m)") +
  facet_grid(~ Depth, switch = "x", scales = "free_x", space = "free_x") +
  scale_fill_gradient(name = "Sqrt(Abundance)",
                      low = "#FFFFFF",
                      high = "#012345") +
  theme(strip.placement = "outside")

mine.heatmap
```

#### Plot title

Plot titles can be added with `ggtitle`:

```{r plot-title}
mine.long$Sqrt.abundance <- sqrt(mine.long$Abundance)
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Sqrt.abundance)) +
  geom_tile() +
  xlab(label = "Depth (m)") +
  facet_grid(~ Depth, switch = "x", scales = "free_x", space = "free_x") +
  scale_fill_gradient(name = "Sqrt(Abundance)",
                      low = "#FFFFFF",
                      high = "#012345") +
  theme(strip.placement = "outside") +
  ggtitle(label = "Microbe Class Abundance")

mine.heatmap
```

But by default the title is left-aligned. To center-justify the title, we add 
another argument, `plot.title` to the `theme` call:

```{r plot-title-align}
mine.long$Sqrt.abundance <- sqrt(mine.long$Abundance)
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Sqrt.abundance)) +
  geom_tile() +
  xlab(label = "Depth (m)") +
  facet_grid(~ Depth, switch = "x", scales = "free_x", space = "free_x") +
  scale_fill_gradient(name = "Sqrt(Abundance)",
                      low = "#FFFFFF",
                      high = "#012345") +
  theme(strip.placement = "outside",
        plot.title = element_text(hjust = 0.5)) +
  ggtitle(label = "Microbe Class Abundance")

mine.heatmap
```

#### Miscellaneous Debris

There are a couple more things we will change in the plot:

1. Reverse the order of the y-axis, so 'Actinobacteria' is at the top
2. Remove the y-axis title
3. Change the colors of the boxes for the depth categories

The second two are done via the `theme` command, with the `axis.title.y` and
`strip.background` arguments. The first, though, requires consideration of how 
ggplot determines the order of the y-axis. When the axis values are categories, 
ggplot treats them as factors and places them, from bottom to top, in the order 
of the levels for the factor. Since we are using values in the `Class` column, 
take a look at the default order if the column is treated as a factor:

```{r y-axis-levels}
levels(as.factor(mine.long$Class))
```

Since `r head(levels(as.factor(mine.long$Class)), n = 1)` is the first level, 
it appears at the bottom of the y-axis; the last level is `r tail(levels(as.factor(mine.long$Class)), n = 1)` 
and appears at the top of the plot. To reverse this order, we use 
`scale_y_discrete` and pass the factor levels in reverse. What? Take another 
look at the output of the `levels` command. If we wanted to reverse this 
output, we can use the `rev` function:

```{r y-axis-levels-rev}
rev(levels(as.factor(mine.long$Class)))
```

Now we use the same command in `scale_y_discrete` for the `limits` argument:

```{r y-axis-order-rev}
mine.long$Sqrt.abundance <- sqrt(mine.long$Abundance)
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Sqrt.abundance)) +
  geom_tile() +
  xlab(label = "Depth (m)") +
  facet_grid(~ Depth, switch = "x", scales = "free_x", space = "free_x") +
  scale_fill_gradient(name = "Sqrt(Abundance)",
                      low = "#FFFFFF",
                      high = "#012345") +
  theme(strip.placement = "outside",
        plot.title = element_text(hjust = 0.5)) +
  ggtitle(label = "Microbe Class Abundance") +
  scale_y_discrete(limits = rev(levels(as.factor(mine.long$Class))))

mine.heatmap
```

And to adjust the y-axis and change the depth boxes, we add two more arguments 
to the `theme` command:

```{r theme-changes}
mine.long$Sqrt.abundance <- sqrt(mine.long$Abundance)
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Sqrt.abundance)) +
  geom_tile() +
  xlab(label = "Depth (m)") +
  facet_grid(~ Depth, switch = "x", scales = "free_x", space = "free_x") +
  scale_fill_gradient(name = "Sqrt(Abundance)",
                      low = "#FFFFFF",
                      high = "#012345") +
  theme(strip.placement = "outside",
        plot.title = element_text(hjust = 0.5),
        axis.title.y = element_blank(), # Remove y-axis title
        strip.background = element_rect(fill = "#EEEEEE", color = "#FFFFFF")) +
  ggtitle(label = "Microbe Class Abundance") +
  scale_y_discrete(limits = rev(levels(as.factor(mine.long$Class))))

mine.heatmap
```

Finally, there are many default themes in ggplot that affect background colors, 
grid lines, and fonts. You can see examples of them in use at
[https://ggplot2.tidyverse.org/reference/ggtheme.html](https://ggplot2.tidyverse.org/reference/ggtheme.html). 
One relatively simple theme is `theme_bw`, and we can apply it by adding the 
function to our ggplot object:

```{r theme-bw-01}
mine.long$Sqrt.abundance <- sqrt(mine.long$Abundance)
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Sqrt.abundance)) +
  geom_tile() +
  xlab(label = "Depth (m)") +
  facet_grid(~ Depth, switch = "x", scales = "free_x", space = "free_x") +
  scale_fill_gradient(name = "Sqrt(Abundance)",
                      low = "#FFFFFF",
                      high = "#012345") +
  theme(strip.placement = "outside",
        plot.title = element_text(hjust = 0.5),
        axis.title.y = element_blank(),
        strip.background = element_rect(fill = "#EEEEEE", color = "#FFFFFF")) +
  ggtitle(label = "Microbe Class Abundance") +
  scale_y_discrete(limits = rev(levels(as.factor(mine.long$Class)))) +
  theme_bw() # Use the black and white theme

mine.heatmap
```

Most notably, the different depth plots all have a black border around them. 
But if we take a closer look, a number of our changes have been undone! The 
shading of the boxes indicating Depth is darker, the labels for the samples are 
back at the bottom, and the y-axis title ("Class") has returned. 

Why did this happen? Didn't we fix all of that with the call to `theme`?

This demonstrates the importance of _order_ in building ggplot objects. When we 
use the function `theme_bw`, we are using the settings in theme_bw for **all** 
theme elements. Thus by adding `theme_bw` _after_ our call to `theme`, we 
effectively throw out all those changes we made. In order for our changes to 
stay in place, we will need to call `theme_bw` _before_ our call to `theme`:

```{r theme-bw-02}
mine.long$Sqrt.abundance <- sqrt(mine.long$Abundance)
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Sqrt.abundance)) +
  geom_tile() +
  xlab(label = "Depth (m)") +
  facet_grid(~ Depth, switch = "x", scales = "free_x", space = "free_x") +
  scale_fill_gradient(name = "Sqrt(Abundance)",
                      low = "#FFFFFF",
                      high = "#012345") +
  theme_bw() +
  theme(strip.placement = "outside",
        plot.title = element_text(hjust = 0.5),
        axis.title.y = element_blank(),
        strip.background = element_rect(fill = "#EEEEEE", color = "#FFFFFF")) +
  ggtitle(label = "Microbe Class Abundance") +
  scale_y_discrete(limits = rev(levels(as.factor(mine.long$Class))))

mine.heatmap
```
  
Our final script for this heatmap is then:

```{r final-script, eval = FALSE}
# Heatmap of mine pit microbe diversity
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-06-05

rm(list = ls())

# Load dependancies
library("tidyr")
library("ggplot2")

# Read data and format for heatmap
mine.data <- read.csv(file = "data/mine-data.csv")
mine.long <- pivot_wider(data = mine.data,
                         cols = -c(1:3),
                         names_to = "Class",
                         values_to = "Abundance")

# Transform abundance data for better visualization
mine.long$Sqrt.abundance <- sqrt(mine.long$Abundance)

# Plot abundance
mine.heatmap <- ggplot(data = mine.long, mapping = aes(x = Sample.name,
                                                       y = Class,
                                                       fill = Sqrt.abundance)) +
  geom_tile() +
  xlab(label = "Depth (m)") +
  # Facet on depth and drop empty columns
  facet_grid(~ Depth, switch = "x", scales = "free_x", space = "free_x") + 
  # Set colors different from default
  scale_fill_gradient(name = "Sqrt(Abundance)", 
                      low = "#FFFFFF",
                      high = "#012345") +
  theme_bw() +
  
  theme(strip.placement = "outside", # Move depth boxes to bottom of plot
        plot.title = element_text(hjust = 0.5), # Center-justify plot title
        axis.title.y = element_blank(), # Remove y-axis title
        strip.background = element_rect(fill = "#EEEEEE", color = "#FFFFFF")) +
  ggtitle(label = "Microbe Class Abundance") +
  scale_y_discrete(limits = rev(levels(as.factor(mine.long$Class))))

mine.heatmap
```

***

## Additional resources

+ Paper describing [tidy data](https://www.jstatsoft.org/article/view/v059i10)
+ A great introduction to [data tidying](http://garrettgman.github.io/tidying/)
+ A [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) for data wrangling
+ [Official documentation](http://ggplot2.tidyverse.org/reference/) for ggplot
+ A [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) for ggplot
+ [Documentation](http://ggplot2.tidyverse.org/reference/geom_bin2d.html) for `geom_bin2d`, to create heatmaps for continuous x- and y-axes
+ A [PDF version](https://jcoliver.github.io/learn-r/006-heatmaps.pdf) of this lesson

***

[Back to learn-r main page](index.html)

Questions?  e-mail me at <a href="mailto:jcoliver@arizona.edu">jcoliver@arizona.edu</a>.