---
title: "Introduction to Functional Programming"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: default
  pdf_document: default
---

An introduction to writing functions to improve your coding efficiency and optimize performance.

#### Learning objectives
1. Write and use functions in R
2. Document functions for easy re-use
3. Replace loops with functions optimized for vector calculations

## Don't Repeat Yourself
The [DRY principle](https://en.wikipedia.org/wiki/Don%27t_repeat_yourself) aims to reduce repetition in software engineering. By writing and using functions to accomplish a set of instructions multiple times, you reduce the opportunities for mistakes and often improve performance of the code you write. Functional programming makes it easy to apply the same analyses to different sets of data, without excessive copy-paste-update cycles that can introduce hard-to-detect errors.

***

## Writing functions
Why do we write functions? Usually, we create functions after writing some code with certain variables, then copy/pasting that code and changing the variable names. Then more copy/paste. Then we forget to change one of the variables and spend a day figuring out why our results don't make sense.

For example, consider the `iris` data set:

```{r iris-head}
head(iris)
```

And if we look at the values for each of the four measurements, some are quite different in their ranges.

```{r iris-summary}
summary(iris)
```

Given this variation in ranges and magnitudes, we will have to standardize the variables if we want to run analyses like principle components analysis (PCA) (PCA is beyond the scope of this lesson, but it is covered in a [separate lesson](https://jcoliver.github.io/learn-r/003-intro-multivariate.html)). Standardizing a variable means we transform the data so the mean value is zero and the variance is one.

We'll start by creating an R script to hold our work. R scripts are simple text files with a series of R commands and are a great way to make our work reproducible. For all scripts, it is a good idea to start off with a little information about what the script does (and who is responsible for it). So create a script (from the File menu, select New File > R Script or use the keyboard shortcut of Ctrl+Shift+N or Cmd+Shift+N).

```{r script-header}
# Standarize iris values
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2018-08-16
```

R will ignore anything that appears to the right of the pound sign, `#`. This information allows anyone who is reading the script to quickly know (1) what the script does and (2) who to contact with questions.

To standardize a variable, we subtract observed values by the mean, and divide that difference by the standard deviation ($(x_i - \mu)/\sigma$).

We start by performing the calculation on the petal length data, first calculating the two summary statistics (mean and standard deviation), then using them in the formula for standardization.

```{r standardize-petal-length}
# Standardize petal length
mean_petal_length <- mean(iris$Petal.Length)
sd_petal_length <- sd(iris$Petal.Length)
standard_petal_length <- (iris$Petal.Length - mean_petal_length)/sd_petal_length
```

We can use the `summary` command to ensure the mean is zero,

```{r check-standardization-mean}
summary(standard_petal_length)
```

and the `sd` function to make sure the standard deviation is one.

```{r check-standardization-sd}
sd(standard_petal_length)
```

Now that we have the calculation for petal length, we can copy the code and update it to perform calculation for petal width. 

```{r petal-width-standardization}
# Standardize petal width
mean_petal_width <- mean(iris$Petal.Width)
sd_petal_width <- sd(iris$Petal.Width)
standard_petal_width <- (iris$Petal.Width - mean_petal_width)/sd_petal_width
summary(standard_petal_width)
```

But we had to change three instances of "Length" to "Width" and seven (seven!?) instances of "length" to "width". Did copy/paste really save time? And how many opportunities for mistakes?

Here we have a perfect opportunity to write a function that will do those three steps for any data we want. _And_ we can avoid the inevitable errors that arise with the copy-paste-update process.

### Write a function!

When we write a function, it is very much like assigning a value to a variable. However, in this instance, we assign the value produced by a special function in R called `function` (shocking, I know). When writing functions, I generally start by naming the function and creating a function with nothing in it. We will call our function `standardize`:

```{r function-skeleton, eval = FALSE}
standardize <- function(data) {
  
}
```

Note that there is a variable `data` in the call to `function`. This is referring to the data that we will perform our standarization calculation on. The next step is to outline what actions the function will peform. I start with comments to help with this process.

```{r function-steps, eval = FALSE}
standardize <- function(data) {
  # Calculate mean of data
  # Calculate standard deviation of data
  # Calculate standardized scores
}
```

These comments then provide a roadmap of the code that needs to be written inside the function. We start by adding the mean calculation to the function:

```{r function-mean, eval = FALSE}
standardize <- function(data) {
  # Calculate mean of data
  mean_data <- mean(data)
  # Calculate standard deviation of data
  # Calculate standardized scores
}
```

Next we fill in the standard deviation and standardization code:

```{r function-sd, eval = FALSE}
standardize <- function(data) {
  # Calculate mean of data
  mean_data <- mean(data)
  # Calculate standard deviation of data
  sd_data <- sd(data)
  # Calculate standardized scores
  stand_data <- (data - mean_data)/sd_data
}
```

At this point, remember that we want to use this function with the iris data. However, at no point in the function are we referring to the iris data frame. This is because we want our function to be _generalizable_ to other data as well. The last thing we need to do is to send the standardized values back to the user. This is known as "returning" a value, and we use the function `return` to do so:

```{r function-finish, eval = FALSE}
standardize <- function(data) {
  # Calculate mean of data
  mean_data <- mean(data)
  # Calculate standard deviation of data
  sd_data <- sd(data)
  # Calculate standardized scores
  stand_data <- (data - mean_data)/sd_data
  return(stand_data)
}
```

At this point, our function is almost ready for use. The last thing we need to do is to read it into memory. We do this like we are running any other line of code in R: place your cursor on one of the curly braces and run the code with Ctrl-Enter or Cmd-Enter.

Now we can use the function. We can replace the three lines that performed the calculation for petal length with a single line that calls our `standarize` function:

```{r use-function-length}
standard_petal_length <- standardize(data = iris$Petal.Length)
```

Like all code, we want to provide enough information so it is easy for someone else to use. So we want to add documentation that consists at a minimum of three things:

1. A brief explanation of what the function does
2. A description of the information needs to be provided to the function
3. An explanation of what the function returns

```{r function-documentation}
# Standardize a vector of numerical data
# data: vector of numerical data
# returns: vector of original data standardized to have mean = 0, sd = 1
standardize <- function(data) {
  # Calculate mean of data
  mean_data <- mean(data)
  # Calculate standard deviation of data
  sd_data <- sd(data)
  # Calculate standardized scores
  stand_data <- (data - mean_data)/sd_data
  return(stand_data)
}
```

Now we have a function we can use on each of the columns in our data frame. Yay, right? Well, almost. We still have to run the `standarize` function four times on the iris data frame. What if we had 100 columns of data? Or 1000 columns? That is simply too much copy and paste. 

Use apply to do it to multiple columns
```{r apply-function-01, eval = FALSE}
iris_stand <- apply(X = iris, MARGIN = 2, FUN = standardize)
```

```{r apply-function-02}
iris_stand <- apply(X = iris[, 1:4], MARGIN = 2, FUN = standardize)
iris_stand <- as.data.frame(iris_stand)
```

Add the species column
```{r add-species-column}
iris_stand$Species <- iris$Species
```

Write a function to do it all
(but do this in a separate file - move our other function there, too)
```{r standardize-df}
# data: data frame to standardize
# ignore: (optional) vector of columns to skip in standardization process
# returns: data frame with all non-ignored columns standardized to have mean = 0, sd = 1
standardize_df <- function(data, ignore = c()) {
  ignored_cols <- data[, ignore]
  df_stand <- apply(X = data[, -ignore], MARGIN = 2, FUN = standardize)
  df_stand <- cbind(df_stand, ignored.cols)
  return(df_stand)
}

iris_stand <- standardize_df(data = iris, ignore = 5)
```


```{r standardize-df-names}
# data: data frame to standardize
# ignore: (optional) vector of columns to skip in standardization process
# returns: data frame with all non-ignored columns standardized to have mean = 0, sd = 1
standardize_df <- function(data, ignore = c()) {
  # Do standardization calculation on all columns except those indicated by 
  # ignore vector
  df_stand <- apply(X = data[, -ignore], MARGIN = 2, FUN = standardize)

  # To add the ignored columns back to the data frame of standarized values, 
  # need to extract column names, so we know the order they should appear and 
  # the column names of the ignored columns only
  orig_colnames <- colnames(data)
  ignored_colnames <- colnames(data)[ignored]
  
  # Extract the ignored columns and ensure they are a data frame (if only one 
  # column is ignored, extracting it alone results in a vector, not a data 
  # frame)
  ignored_columns <- as.data.frame(data[, ignore])
  
  # Update column names for this data frame; really only necessary when a single
  # column is ignored
  colnames(ignored_columns) <- ignored_colnames
  
  df_stand <- cbind(df_stand, ignored_columns)
  
  # Ensure columns appear in original order
  df_stand <- df_stand[, orig_colnames]
  
  return(df_stand)
}

iris_stand <- standardize_df(data = iris, ignore = 5)
```

***
# Old Lesson

#### Learning objectives
1. Write and use functions in R
2. Document functions for easy re-use
3. Replace loops with functions optimized for vector calculations

## Don't Repeat Yourself
The [DRY principle](https://en.wikipedia.org/wiki/Don%27t_repeat_yourself) aims to reduce repetition in software engineering. By writing and using functions to accomplish a set of instructions multiple times, you reduce the opportunities for mistakes and often improve performance of the code you write. Functional programming makes it easy to apply the same analyses to different sets of data, without excessive copy-paste-update cycles that can introduce hard-to-detect errors.

***

## Writing functions
Why do we write functions? Usually, we create functions after writing some code with certain variables, then copy/pasting that code and changing the variable names. Then more copy/paste. Then we forget to change one of the variables and spend a day figuring out why our results don't make sense.

For example, consider the `airquality` data set:

```{r, echo = FALSE}
head(airquality)
```

Let's start by writing a script to run linear regression to see if the solar radiation (the `Solar.R` column) predicts the ozone level (the `Ozone` column). Create a new file called "airquality-regression.R":

```{r}
# Analyze air quality data
# Jeffrey Oliver
# jcoliver@email.arizona.edu
# 2017-06-22

# Relationship between ozone and solar radiation
simple <- lm(airquality[, "Ozone"] ~ airquality[, "Solar.R"])

```

Now we are interested in just the correlation coefficient (r^2^) and the p-value for this relationship, so we use `summary` and extract those values:

```{r}
# Extract the model parameters
simple.summary <- summary(simple)
simple.r2 <- simple.summary$r.squared
simple.p <- simple.summary$coefficients[2, 4]
```

And finally we can print out these two values with the `cat` command:

```{r, results = "hold"}
cat("Solar r^2 =", simple.r2)
cat("Solar p =", simple.p)  
```

If we want to do the same thing for the relationship between ozone and wind, we can copy/paste this code and just change the predictor column specification in the `lm` command from `"Solar.R"` to `"Wind"` and update the message in the `cat` commands:

```{r, eval = FALSE}
simple <- lm(airquality[, "Ozone"] ~ airquality[, "Wind"])
simple.summary <- summary(simple)
simple.r2 <- simple.summary$r.squared
simple.p <- simple.summary$coefficients[2, 4]
cat("Wind r^2 =", simple.r2)
cat("Wind p =", simple.p)  
```

This isn't too great an effort, but as the number of predictors grows, the time and opportunty for mistakes also grows. Since we are doing the same exact process for each predictor variable (run `lm`, run `summary`, extract values of interest, then print values of interest), we can encapsulate this in a function.

### Behold, a function!

To define a function, create a new file called "regression-functions.R". While functions do not necessarily have to exist outside of the R script in which they are called, it is generally considered good practice. Start this file with the usual header containing information about the contents:

```{r, eval = FALSE}
# Functions to automate linear regression
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-06-15
```

And we define a function pretty much the same way we assign values to a variable, but here we use the `function` function:

```{r, eval = FALSE}
RegressSimple <- function() {
}
```

We have a general idea of what the function should do, so write that as a brief comment above the function:

```{r, eval = FALSE}
# Run linear regression for all predictors and a response variable in a data frame
RegressSimple <- function() {
}
```

Now what do we do? How does one actually go about writing functions? Maybe you know exactly what your function should do and all the inputs and outputs. I rarely find myself in that case. Rather, I start by doing the same thing as I did before: I copy paste the portion of code I want to run. So in this case, from the airquality-regression.R file, copy the section for the solar radiation analysis and paste it into the body of the `RegressSimple` function. That is, paste it between the pair of curly braces `{ }`:

```{r, eval = FALSE}
# Run linear regression for all predictors and a response variable in a data frame
RegressSimple <- function() {
  simple <- lm(airquality[, "Ozone"] ~ airquality[, "Solar.R"])
  simple.summary <- summary(simple)
  simple.r2 <- simple.summary$r.squared
  simple.p <- simple.summary$coefficients[2, 4]
  cat("Solar r^2 =", simple.r2)
  cat("Solar p =", simple.p)  
}
```

We don't want the function to print out values from the models, so delete the code that extracts the values and prints them with the two `cat` statements:
```{r, eval = FALSE}
# Run linear regression for all predictors and a response variable in a data frame
RegressSimple <- function() {
  simple <- lm(airquality[, "Ozone"] ~ airquality[, "Solar.R"])
  simple.summary <- summary(simple)
}
```

At this point, we can consider input and output of the function. We'll need to give the function two pieces of input: the data to work with and which variable to use as the response. We specify input by declaring the names of the values (`data` and `response`) in the `function` call:

```{r, eval = FALSE}
# Run linear regression for all predictors and a response variable in a data frame
RegressSimple <- function(data, response) {
  simple <- lm(airquality[, "Ozone"] ~ airquality[, "Solar.R"])
  simple.summary <- summary(simple)
}
```

And while we're at it, we need to document these inputs:
```{r, eval = FALSE}
# Run linear regression for all predictors and a response variable in a data frame
# data: the data frame to analyze
# response: the name of the response variable
RegressSimple <- function(data, response) {
  simple <- lm(airquality[, "Ozone"] ~ airquality[, "Solar.R"])
  simple.summary <- summary(simple)
}
```

Now that we have inputs, we can update the variables in the code we copied from the regression script. What do we need to update?

+ The name of the data frame we used, `airquality` is replaced by `data`
+ The model specification now uses the abstracted response variable, stored in `response` instead of the hard coded `"Ozone"`; note that we _do not_ put `response` in double quotes

```{r, eval = FALSE}
# Run linear regression for all predictors and a response variable in a data frame
# data: the data frame to analyze
# response: the name of the response variable
RegressSimple <- function(data, response) {
  simple <- lm(data[, response] ~ data[, "Solar.R"])
  simple.summary <- summary(simple)
}
```

The goal of the function is to run linear regression for _all_ the predictors in the data frame, so how can we do this? The simplest way is to use a `for` loop for all the columns in the data frame, updating the `lm` call with the column specification of the predictor variable:

```{r, eval = FALSE}
# Run linear regression for all predictors and a response variable in a data frame
# data: the data frame to analyze
# response: the name of the response variable
RegressSimple <- function(data, response) {
  for (predictor in 1:ncol(data)) {
    simple <- lm(data[, response] ~ data[, predictor])
    simple.summary <- summary(simple)
  }
}
```

For output, we probably want the results of the linear model for each of our predictors. In this case, we'll use a `list` object and assigning the output of `summary` to an element in that list. Note because we are using a list, we use two-bracket notation `[[ ]]` to indicate an element in the list. And finally, we sent back these results with the `return` function:

```{r, eval = FALSE}
# Run linear regression for all predictors and a response variable in a data frame
# data: the data frame to analyze
# response: the name of the response variable
RegressSimple <- function(data, response) {
  model.summaries <- list()
  for (predictor in 1:ncol(data)) {
    simple <- lm(data[, response] ~ data[, predictor])
    element.name <- colnames(data)[predictor]
    model.summaries[[element.name]] <- summary(simple)
  }
  return(model.summaries)
}
```

Before we try this out, update the documentation with a description of the output
```{r}
# Run linear regression for all predictors and a response variable in a data frame
# data: the data frame to analyze
# response: the name of the response variable
# returns: a list where each element is the output from summary(lm)
RegressSimple <- function(data, response) {
  model.summaries <- list()
  for (predictor in 1:ncol(data)) {
    simple <- lm(data[, response] ~ data[, predictor])
    element.name <- colnames(data)[predictor]
    model.summaries[[element.name]] <- summary(simple)
  }
  return(model.summaries)
}
```

### Using our function

So how do we use this? We need to load this function into memory so we can use it. Go back to the script with our original regression analyses, "airquality-regression.R". Comment out our previous linear regression code and add a call to `source` to load our function file. Hint: in RStudio, you can select multiple lines and comment them out with the shortcut Shift-Ctrl-C.

```{r, eval = FALSE}
# Analyze air quality data
# Jeffrey Oliver
# jcoliver@email.arizona.edu
# 2017-06-22

source(file = "regression-functions.R")

# Relationship between ozone and solar radiation
# simple <- lm(airquality[, "Ozone"] ~ airquality[, "Solar.R"])

# Extract the model parameters
# simple.summary <- summary(simple)
# simple.r2 <- simple.summary$r.squared
# simple.p <- simple.summary$coefficients[2, 4]
# cat("Solar r^2 =", simple.r2)
# cat("Solar p =", simple.p)  
```

And we can now use this function, passing `airquality` as the `data` and `"Ozone"` as the `response`:

```{r}
airquality.models <- RegressSimple(data = airquality, response = "Ozone")
```

Hmmmm...that's an odd warning. Let's add some reporting code to see if we can figure out what it's doing. Update the `RegressSimple` function to print the name of the predictor using the `cat` function:

```{r, eval = FALSE}
# Run linear regression for all predictors and a response variable in a data frame
# data: the data frame to analyze
# response: the name of the response variable
# returns: a list where each element is the output from summary(lm)
RegressSimple <- function(data, response) {
  model.summaries <- list()
  for (predictor in 1:ncol(data)) {
    simple <- lm(data[, response] ~ data[, predictor])
    element.name <- colnames(data)[predictor]
    model.summaries[[element.name]] <- summary(simple)
    cat("Predictor:", element.name, "\n")
  }
  return(model.summaries)
}
```

Now go back our airquality-regression.R script and run the `RegressSimple` command again.

```{r}
airquality.models <- RegressSimple(data = airquality, response = "Ozone")
```

Drat. Same warning, but no message printed? Why not? Because we made changes to the regression-functions.R file, but did not load them into memory with the `source` command, R is using the old version of the `RegressSimple` function. So we need to run the `source` command first:

```{r, eval = FALSE}
source(file = "regression-functions.R")
```

```{r, echo = FALSE}
# Run linear regression for all predictors and a response variable in a data frame
# data: the data frame to analyze
# response: the name of the response variable
# returns: a list where each element is the output from summary(lm)
RegressSimple <- function(data, response) {
  model.summaries <- list()
  for (predictor in 1:ncol(data)) {
    simple <- lm(data[, response] ~ data[, predictor])
    element.name <- colnames(data)[predictor]
    model.summaries[[element.name]] <- summary(simple)
    cat("Predictor:", element.name, "\n")
  }
  return(model.summaries)
}
```

Then the `RegressSimple` command:
```{r}
airquality.models <- RegressSimple(data = airquality, response = "Ozone")
```

OK, so there are some problems. First, we aren't really interested in the effect of `Day` or `Month` on ozone levels, so when we call `RegressSimple`, we should only pass it data we want to analyze. Here we drop the fifth and sixth columns, which are the Month and Day columns, respectively:

```{r, eval = FALSE}
airquality.models <- RegressSimple(data = airquality[, -c(5:6)], response = "Ozone")
```

But look at the output from function call again. Our function actually ran regression on a model using ozone to predict ozone - that's probably what caused the warning message "summary may be unreliable". But we don't want to exclude ozone from the data we pass to the function, because that is the response variable. We therefore need to update our function definition so we don't run an `ozone ~ ozone` model. More generally, we need to make sure our response variable is not treated as a predictor. To do this, we:

1. Find out which column is the `response` variable
2. Create a vector of predictor variable names using `colnames`
3. Update our `for` loop to only use those predictor variables
4. Use the predictor variable name for the `model.summaries` list element name

Open the regression-functions.R file and update `RegressSimple`:

```{r}
# Run linear regression for all predictors and a response variable in a data frame
# data: the data frame to analyze
# response: the name of the response variable
# returns: a list where each element is the output from summary(lm)
RegressSimple <- function(data, response) {
  response.index <- which(colnames(data) == response)
  predictors <- colnames(data)[-response.index]
  model.summaries <- list()
  for (predictor in predictors) {
    simple <- lm(data[, response] ~ data[, predictor])
    model.summaries[[predictor]] <- summary(simple)
    cat("Predictor:", predictor, "\n")
  }
  return(model.summaries)
}
```

Now when we run `RegressSimple`, there are only three predictors used in the models, as we expect:

```{r, eval = FALSE}
source(file = "regression-functions.R")
airquality.models <- RegressSimple(data = airquality[, -c(5:6)], response = "Ozone")
```

```{r, echo = FALSE}
airquality.models <- RegressSimple(data = airquality[, -c(5:6)], response = "Ozone")
```

Since it `RegressSimple` is now only running the models we want, remove the `cat` command from the function:

```{r}
# Run linear regression for all predictors and a response variable in a data frame
# data: the data frame to analyze
# response: the name of the response variable
# returns: a list where each element is the output from summary(lm)
RegressSimple <- function(data, response) {
  response.index <- which(colnames(data) == response)
  predictors <- colnames(data)[-response.index]
  model.summaries <- list()
  for (predictor in predictors) {
    simple <- lm(data[, response] ~ data[, predictor])
    model.summaries[[predictor]] <- summary(simple)
  }
  return(model.summaries)
}
```

Because `airquality.models` is a list, we can access the objects using double-bracket notation:

```{r, results = "hold"}
solar.model <- airquality.models[["Solar.R"]]
solar.corr <- solar.model$r.squared
solar.p <- solar.model$coefficients[2, 4]
cat("Solar r^2 = ", solar.corr, "\n")
cat("Solar p = ", solar.p, "\n")
```

### Make it Class-y

But now we're back to the copy-paste-update cycle if we want to get all the correlation coefficients and p-values. If we know that's all we want, we can create another function, one that does the work of heavy lifting of extracting coefficients and printing them out to the screen.

In the file with our `RegressSimple` function, create _another_ function, and call it `print.RegressSimple`:

```{r, eval = FALSE}
print.RegressSimple <- function(x, ...) {
}
```

Briefly, what we are doing is creating a function that specifies the output of anything that is `class RegressSimple` (we'll get to how we make that happen in a moment). The thing to note now is that in a `print.____` function, the first argument, `x` is the object we wish to print; in this case, a product of the function `RegressSimple`. Stick with me here, it will become clearer. For the purposes of this lesson, just remember that in the `print.RegressSimple` function, the variable `x` is the list that is produced from a call to `RegressSimple`. We want this function to extract the r^2^ and p-values for each model and display a table of those values for each predictor variable. Something like:

```{r, echo = FALSE}
cat("  ", "variable", "r2  ", "p", "\n", 
    "  ", "Solar.R ", "0.12", "1.7e-04", "\n",
    "  ", "Wind    ", "0.36", "9.2e-13", "\n", 
    "  ", "Temp    ", "0.48", "2.9e-18", "\n", sep = "\t")
##      variable r2        p           
## [1,] Solar.R  0.1213419 1.793109e-04
## [2,] Wind     0.3618582 9.271974e-13
## [3,] Temp     0.4877072 2.931897e-18
```

Add a brief explanation of what this function does, then add code to get the names of the variables and set up a data frame to hold the values we want to print:

```{r, eval = FALSE}
# Print values of interest from each predictor in a RegressSimple object
print.RegressSimple <- function(x, ...) {
  # Get a vector of the elements' names
  predictors <- names(x)
  # Set up a dataframe for results of interest
  model.results <- data.frame(variable = predictors,
                           r2 = 0,
                           p = 0)
}
```

Now extract the values we want to print for each element in the `RegressSimple` object (which is the variable `x` in `print.RegressSimple`):

```{r, eval = FALSE}
# Print values of interest from each predictor in a RegressSimple object
print.RegressSimple <- function(x, ...) {
  # Get a vector of the elements' names
  predictors <- names(x)
  # Set up a dataframe for results of interest
  model.results <- data.frame(variable = predictors,
                           r2 = 0,
                           p = 0)
  # Extract r-squared and p-values
  model.results$r2 <- sapply(x, "[[", "r.squared")
  model.coeffs <- sapply(x, "[[", "coefficients")
  model.results$p <- model.coeffs[8, ]
}
```

Finally, we add `cat` and `print` statements to output the values.

```{r}
# Print values of interest from each predictor in a RegressSimple object
print.RegressSimple <- function(x, ...) {
  # Get a vector of the elements' names
  predictors <- names(x)
  # Set up a dataframe for results of interest
  model.results <- data.frame(variable = predictors,
                           r2 = 0,
                           p = 0)
  # Extract r-squared and p-values
  model.results$r2 <- sapply(x, "[[", "r.squared")
  model.coeffs <- sapply(x, "[[", "coefficients")
  model.results$p <- model.coeffs[8, ]

  # Print values
  cat("Regression results: ", "\n")
  print(as.matrix(model.results), quote = FALSE)
}
```

Now when we run our code, we can just enter the name of the variable to have it print out our nice table of just the r^2^ and p-values.

```{r, eval = FALSE}
source(file = "regression-functions.R")
airquality.models <- RegressSimple(data = airquality[, -c(5:6)], response = "Ozone")
airquality.models
```

```{r, echo = FALSE}
airquality.models <- RegressSimple(data = airquality[, -c(5:6)], response = "Ozone")
airquality.models
```

Well that didn't work. It just printed out the each element of the list. That's because we need to take one final step. The `print.RegressSimple` only works on objects that are of `class RegressSimple`. So what is the class of our `airquality.models`?

```{r}
class(airquality.models)
```

It's a `r class(airquality.models)`, _not_ a `RegressSimple` object. So how do we instruct R to make the output of `RegressSimple` to be an object of `class RegressSimple`? Big surprise here, we use the `class` function! In our `RegressSimple` function, right before we return the `model.summaries` list, we set the class to be "RegressSimple":

```{r}
# Run linear regression for all predictors and a response variable in a data frame
# data: the data frame to analyze
# response: the name of the response variable
# returns: a list where each element is the output from summary(lm)
RegressSimple <- function(data, response) {
  response.index <- which(colnames(data) == response)
  predictors <- colnames(data)[-response.index]
  model.summaries <- list()
  for (predictor in predictors) {
    simple <- lm(data[, response] ~ data[, predictor])
    model.summaries[[predictor]] <- summary(simple)
  }
  
  # Set class and return results
  class(model.summaries) <- "RegressSimple"
  return(model.summaries)
}
```

Now re-run the lines in our regression script:
```{r, eval = FALSE}
source(file = "regression-functions.R")
airquality.models <- RegressSimple(data = airquality[, -c(5:6)], response = "Ozone")
airquality.models
```

```{r, echo = FALSE}
airquality.models <- RegressSimple(data = airquality[, -c(5:6)], response = "Ozone")
airquality.models
```

There it is, our table of values! Our final two files will then be:

airquality-regression.R:

```{r, eval = FALSE}
# Analyze air quality data
# Jeffrey Oliver
# jcoliver@email.arizona.edu
# 2017-06-22

source(file = "regression-functions.R")
airquality.models <- RegressSimple(data = airquality[, -c(5:6)], response = "Ozone")
airquality.models
```

regression-functions.R:

```{r, eval = FALSE}
# Functions to automate linear regression
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-06-15

################################################################################
# Run linear regression for all predictors and a response variable in a data frame
# data: the data frame to analyze
# response: the name of the response variable
# returns: a list where each element is the output from summary(lm)
RegressSimple <- function(data, response) {
  response.index <- which(colnames(data) == response)
  predictors <- colnames(data)[-response.index]
  model.summaries <- list()
  for (predictor in predictors) {
    simple <- lm(data[, response] ~ data[, predictor])
    model.summaries[[predictor]] <- summary(simple)
  }
  
  # Set class and return results
  class(model.summaries) <- "RegressSimple"
  return(model.summaries)
}

################################################################################
# Print values of interest from each predictor in a RegressSimple object
print.RegressSimple <- function(x, ...) {
  # Get a vector of the elements' names
  predictors <- names(x)
  # Set up a dataframe for results of interest
  model.results <- data.frame(variable = predictors,
                           r2 = 0,
                           p = 0)
  # Extract r-squared and p-values
  model.results$r2 <- sapply(x, "[[", "r.squared")
  model.coeffs <- sapply(x, "[[", "coefficients")
  model.results$p <- model.coeffs[8, ]

  # Print values
  cat("Regression results: ", "\n")
  print(as.matrix(model.results), quote = FALSE)
}
```

***

## Additional resources

+ A deeper dive to [functional programming](http://adv-r.had.co.nz/Functional-programming.html)
+ An _even deeper_ dive into [functional programming](http://www.brodrigues.co/fput/)
+ Some [opinions and suggestions](http://adv-r.had.co.nz/Style.html) for naming things like functions (see the 'Object names' section)
+ A [PDF version](https://jcoliver.github.io/learn-r/007-intro-functional-programming.pdf) of this lesson

***

<a href="index.html">Back to learn-r main page</a>
  
Questions?  e-mail me at <a href="mailto:jcoliver@email.arizona.edu">jcoliver@email.arizona.edu</a>.