---
title: "Introduction to Functional Programming"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

An introduction to writing functions to improve your coding efficiency and optimize performance.

#### Learning objectives
1. Replace loops with functions optimized for vector calculations
2. Write and use functions in R
3. Document functions for easy re-use

## Don't Repeat Yourself
The [DRY principle](https://en.wikipedia.org/wiki/Don%27t_repeat_yourself) aims to reduce repetition in software engineering. By writing and using functions to accomplish a set of instructions multiple times, you reduce the opportunities for mistakes and often improve performance of the code you write. Functional programming makes it easy to apply the same analyses to different sets of data, without excessive copy-paste-update cycles that can introduce hard-to-detect errors.

***

## Getting started

***

## R is built for vectors

Loop first, then apply
Introduce at least some of the \*apply or \*aply functions
Use airquality data and get the column means (colMeans) for each month separately? `by(airquality[, 1:4], airquality$Month, colMeans)`

***

## Writing your own functions

Do linear regression with multiple variables, get output for all of them...

Maybe regression predicting Ozone with Solar.R, Wind, and Temp in the `airquality` dataset?

+ Write some functions to do something, ideally some sort of 2-5 step process that we need to do some data manipulation for, then produce a result.
+ Create something then class it and add a print.<class> method. See [Programmers' Niche](https://cran.r-project.org/doc/Rnews/Rnews_2004-1.pdf)
+ Deal with scope? Like global variables and those inside the function; check out python rules for variables
+ If possible, get to the point where we return a `list`
+ Deal with `require` and `stop`?

```{r}
# Analyze air quality data
# Jeffrey Oliver
# jcoliver@email.arizona.edu
# 2017-06-15

# Loop over each predictor and get ....

simple <- lm(airquality[, "Ozone"] ~ airquality[, "Solar.R"])
simple.summary <- summary(simple)
simple.r2 <- simple.summary$r.squared
simple.p <- simple.summary$coefficients[2, 4]
cat("Solar r^2 =", simple.r2)
cat("Solar p =", simple.p)  
    
simple <- lm(airquality[, "Ozone"] ~ airquality[, "Wind"])
simple.summary <- summary(simple)
simple.r2 <- simple.summary$r.squared
simple.p <- simple.summary$coefficients[2, 4]
cat("Wind r^2 =", simple.r2)
cat("Wind p =", simple.p)  

simple <- lm(airquality[, "Ozone"] ~ airquality[, "Temp"])
simple.summary <- summary(simple)
simple.r2 <- simple.summary$r.squared
simple.p <- simple.summary$coefficients[2, 4]
cat("Temp r^2 =", simple.r2)
cat("Temp p =", simple.p)
```

```{r}
# Functions to automate linear regression
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-06-15

################################################################################
# Remember to ONLY pass variables we want to analyze; so omit Month and Day when passing data
RegressSimple <- function(data, response) {
  response.index <- which(colnames(data) == response)
  predictors <- colnames(data)[-response.index]
  model.summaries <- list()
  
  for (predictor in predictors) { # Run this once with just cat("Predictor: ", predictor, "\n")
    simple <- lm(airquality[, "Ozone"] ~ airquality[, predictor])
    simple.summary <- summary(simple)
    model.summaries[[predictor]] <- simple.summary
  }
  
  # Don't do class statement until after we've run the function once
  class(model.summaries) <- "RegressSimple"
  return(model.summaries)
}

################################################################################
print.RegressSimple <- function(x, ...) {
  # Get a vector of the elements' names
  predictors <- names(x)
  # Set up a dataframe for results of interest
  model.results <- data.frame(variable = predictors,
                           r2 = 0,
                           p = 0)
  # Loop over all the elements in x and extract r-squared and p values
  for (predictor in predictors) {
    # Pull out the model from the list
    simple.model <- x[[predictor]]
    # Find the corresponding row in the model.results data.frame
    row.number <- which(model.results$variable == predictor)
    # Store the values of interest in the data frame
    model.results[row.number, "r2"] <- simple.model$r.squared
    model.results[row.number, "p"] <- simple.model$coefficients[2, 4]
  }
  cat("Regression results: ", "\n")
  print(as.matrix(model.results), quote = FALSE)
}

ozone.models <- RegressSimple(data = airquality[, -c(5:6)], response = "Ozone")
ozone.models

```

***

## Additional resources

+ [resource one](url-one)
+ [resource two](url-two)

Example showing timing difference between loop and apply:
https://gist.github.com/KevCaz/d1b252265209be63d1e2

H. Wickham chapter: http://adv-r.had.co.nz/Functional-programming.html
http://www.brodrigues.co/fput/

tidyr lesson: https://github.com/datasciencelabs/2016/blob/master/lectures/wrangling/data-wrangling-with-tidyr.Rmd

H. Wickham's take on naming things (see 'Object names' section in: http://adv-r.had.co.nz/Style.html)


***

<a href="index.html">Back to learn-r main page</a>
  
Questions?  e-mail me at <a href="mailto:jcoliver@email.arizona.edu">jcoliver@email.arizona.edu</a>.