---
title: "Clusters and Heatmaps"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: default
  pdf_document:
    latex_engine: xelatex
urlcolor: blue
---

Want to combine the visualization of quantitative data with clustering 
algorithms? Unsatisfied with the options provided by the base R packages? In 
this hands-on workshop, we'll use the `ggendro` package for R to make
publication-quality graphics.

#### Learning objectives

1. Run clustering analysis and display dedrogram
2. Visualize data in a heatmap
3. Use `grid` package to create multi-plot figures

***

## Getting started

First we need to setup our development environment. Open RStudio and create a 
new project via:

+ File > New Project...
+ Select 'New Directory'
+ For the Project Type select 'New Project'
+ For Directory name, call it something like "r-graphing" (without the quotes)
+ For the subdirectory, select somewhere you will remember (like "My Documents" 
or "Desktop")

We need to create two folders: 'data' will store the data we will be analyzing, 
and 'output' will store the results of our analyses. In the RStudio console:

```{r setup, eval = FALSE}
dir.create("data")
dir.create("output")
```

We will also need a few R packages that are not included in the standard
distribution of R. Those packages are `ggplot2`, `ggdendro`, `tidyr`, and 
`grid` and can be installed with the `install.packages()` function. Note these
packages need only be installed once on your machine.

```{r install-packages, eval = FALSE}
install.packages("ggplot2")
install.packages("ggdendro")
install.packages("grid")
install.packages("tidyr")
```

Finally, download data file from
[https://jcoliver.github.io/learn-r/data/otter-mandible-data.csv](https://jcoliver.github.io/learn-r/data/otter-mandible-data.csv) 
or [http://tinyurl.com/otter-data-csv](http://tinyurl.com/otter-data-csv) (the
latter just re-directs to the former). These data are a subset of those used in 
a study on skull morphology and diet specialization in otters 
[doi: 10.1371/journal.pone.0143236](http://dx.doi.org/10.1371/journal.pone.0143236). **Save this file in the "data" folder we just created**.

***

## Data preparation

For any work that we do, we want to record all the steps we take, so instead of
typing commands directly into the R console, we keep all our work in an R 
script. These scripts are just text files with R commands; by convention, we 
start the script with a brief bit of information about what the script does.

```{r read-data}
# Cluster & heatmap on otter data
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-08-15

# Load dependencies
library("ggplot2")
library("ggdendro")
library("tidyr")
library("grid")

# Read in data
otter <- read.csv(file = "data/otter-mandible-data.csv",
                  stringsAsFactors = TRUE)
```

If we look at the first few rows of data, we see the first three columns have
information about the specimen from which the measurements were taken, while 
columns 4-9 have data for six mandible measurements (`m1` - `m6`):

```{r otter-head}
head(otter)
```

For our purposes, we only want to look at the data for two species, so we 
subset the data, including only those rows that have either "A. cinerea" or 
"L. canadensis" in the `species` column. Note, this _does not_ alter the data 
in the original file we read into R; it only alters the data object `otter`
currently in R's memory.

```{r subset-data}
two_species <- c("A. cinerea", "L. canadensis")
otter <- otter[otter$species %in% two_species, ]
```

The last thing we need to do is scale the data variables (columns 4-9) so
measurements have a mean value of 0 and a variance of 1.

```{r rescale-data}
otter_scaled <- otter
otter_scaled[, c(4:9)] <- scale(otter_scaled[, 4:9])
```

***

## Clustering

To make our figure, we will build the two plots (the cluster diagram and the
heatmap) separately, then use the `grid` framework to put them together. We 
start by making the dendrogram (or cluster).

```{r run-clustering}
# Run clustering
otter_matrix <- as.matrix(otter_scaled[, -c(1:3)])
rownames(otter_matrix) <- otter_scaled$accession
otter_dendro <- as.dendrogram(hclust(d = dist(x = otter_matrix)))

# Create dendro
dendro_plot <- ggdendrogram(data = otter_dendro, rotate = TRUE)

# Preview the plot
print(dendro_plot)
```

We can make those labels a bit smaller:

```{r axis-labels}
dendro_plot <- dendro_plot + theme(axis.text.y = element_text(size = 6))

# Preview the plot
print(dendro_plot)
```

That's it for the cluster image (we'll make a few more changes later on).

***

## Heatmap

We need to start by transforming data to a "long" format, where each row only
contains values for one measurement. For example, our data started in this 
format:

```{r print-wide, echo = FALSE}
otter_scaled[c(1:2), c(1:5)]
```

We want to transform it so there is only a single measurement in each row:

```{r print-long, echo = FALSE}
# For demonstration purposes only; print the first few rows and columns
otter_long_demo <- melt(otter_scaled[c(1:2), c(1:5)], 
                        id = c("species", "museum", "accession"))
otter_long_demo <- pivot_longer(data = otter_scaled[c(1:2), c(1:5)],
                                cols = -c(species, museum, accession),
                                names_to = "measurement",
                                values_to = "value")

otter_long_demo
```

This new data has columns that have information about the specimen (the 
"species", "museum", and "accession" columns), but now has "measurement" and 
"value" columns. Instead of one column for each mandible measurement, each row 
only has one measurement in the "value" column, and the measurement type is 
stored in the "measurement" column.  

We use the `tidyr` package's `pivot_longer` function to handle the data 
transformation:

```{r to-long}
# Heatmap
# Data wrangling
otter_long <- pivot_longer(data = otter_scaled,
                           cols = -c(species, museum, accession),
                           names_to = "measurement",
                           values_to = "value")
```

Now we can use that new data frame, `otter_long` to create our heatmap. In the
`ggplot` package, we use the `geom_tile` layer for creating a heatmap. Given 
our prior experience with the y-axis labels being large, we will again use 
`theme` to make the accession numbers (the y-axis labels) a little smaller:

```{r heatmap-1}
heatmap_plot <- ggplot(data = otter_long, aes(x = measurement, y = accession)) +
  geom_tile(aes(fill = value)) +
  scale_fill_gradient2() +
  theme(axis.text.y = element_text(size = 6))

# Preview the heatmap
print(heatmap_plot)
```

***

## Putting it all together

OK. We made a dendrogram. We made a heatmap. Now how to get them into the same
figure? The `grid` package offers a way to combine separate plots.

```{r grid-plot-1}
grid.newpage()
print(heatmap_plot, 
      vp = viewport(x = 0.4, y = 0.5, width = 0.8, height = 1.0))
print(dendro_plot, 
      vp = viewport(x = 0.90, y = 0.445, width = 0.2, height = 1.0))
```

OK, that's a start, but there are a few things we need to address:

1. Notice the tips of the dendrogram are not in the same order as the y-axis of 
the heatmap. We'll need to re-order the heatmap to match the structure of the
clustering.
2. It would be nice to have the dendrogram and heatmap closer together, without 
a legend separating them.
3. The dendrogram should be vertically stretched so each tip lines up with a 
row in the heatmap plot.

### Re-order heatmap rows to match dendrogram

We'll use the order of the tips in the dendrogram to re-order the rows in our
heatmap. The order of those tips are stored in the `otter_dendro` object, and 
we can extract it with the `order.dendro` function:

```{r order-data}
otter_order <- order.dendrogram(otter_dendro)
```

And now for the fun part. By default, `ggplot` use the level order of the 
y-axis labels as the means of ordering the rows in the heatmap. That is, level 
1 of the factor is plotted in the top row, level 2 is plotted in the second 
row, level 3 in the third row and so on. So we can dictate the order in which 
these rows are plotted by _setting_ the order of the levels in the column we 
use for the y-axis labels in the heatmap. The data we used for the heatmap was 
the long-format, `otter_long` object, and the column we used for the y-axis is
`accession`. Using the order of the dendrogram tips (we stored this above in 
the `otter_order` vector), we can re-level the `otter_long$accession` column to
match the order of the tips in the trees.

```{r reorder-levels}
# Order the levels according to their position in the cluster
otter_long$accession <- factor(x = otter_long$accession,
                               levels = otter_scaled$accession[otter_order], 
                               ordered = TRUE)

```

At this point, we need to re-create the heatmap, because the underlying data
(`otter_long`) has changed:

```{r heatmap-2}
# Heatmap
heatmap_plot <- ggplot(data = otter_long, aes(x = measurement, y = accession)) +
  geom_tile(aes(fill = value)) +
  scale_fill_gradient2() +
  theme(axis.text.y = element_text(size = 6))

# Preview the heatmap
print(heatmap_plot)
```

Now the rows have been re-arranged and we can see right away that the heatmap 
now does a better job of grouping "red" rows together and "blue" rows together.

The combined figure shows the accession numbers in the heatmap are now in the 
same order as the acession numbers in the dendrogram:

```{r grid-plot-2}
grid.newpage()
print(heatmap_plot, 
      vp = viewport(x = 0.4, y = 0.5, width = 0.8, height = 1.0))
print(dendro_plot, 
      vp = viewport(x = 0.90, y = 0.445, width = 0.2, height = 1.0))
```

### Re-position legend

We would like the legend to be somewere else besides in between the two graphs, 
so we will move it to the top of the heatmap. We do this with through the 
`theme` layer when creating the heatmap, setting the value of `legend.position` 
to "top":

```{r heatmap-3}
# Heatmap
heatmap_plot <- ggplot(data = otter_long, aes(x = measurement, y = accession)) +
  geom_tile(aes(fill = value)) +
  scale_fill_gradient2() +
  theme(axis.text.y = element_text(size = 6),
        legend.position = "top")

# Preview the heatmap
print(heatmap_plot)
```

The combined figure now has the heatmap and dendrogram closer together:

```{r grid-plot-3}
grid.newpage()
print(heatmap_plot, 
      vp = viewport(x = 0.4, y = 0.5, width = 0.8, height = 1.0))
print(dendro_plot, 
      vp = viewport(x = 0.90, y = 0.445, width = 0.2, height = 1.0))
```

### Align dendrogram tips with heatmap rows

Finally, we will need to adjust the height and alignment of dendrogram to 
ensure the tips line up with the corresponding rows in the heatmap. 
Unfortunately, because the two figures are actually independent graphics 
objects, this part involves considerable trial and error. In general, we will 
need to change two parameters in the `viewport` function when we call `print` 
on `dendro_plot`:  

1. `y`: This argument sets the vertical justification. A value of 0.5 centers 
the graphic in the middle of the viewport to which it is assigned. Values 
closer to zero move the graphic towards the bottom of the viewport and values 
closer to one move the graphic towards the top of the viewport. In our case, we 
will probably want to move the dendrogram down a little bit, so we use a value 
below 0.5.
2. `height`: This is the proportion of the viewport's vertical space the 
graphic should use. We initially set it to use all the vertical space 
(`height = 1.0`), but since the legend now occupies the real estate at the top 
of the graphic, we will need to make the dendrogram height smaller.

**Note:** The actual values you use will depend on the graphics device you
ultimately use; that is, the values presented here are designed for an html to 
be viewed on the web, but if you are displaying on the screen or writing to a 
pdf file, you may need to use different values for `y` and `height` (e.g. when
calling `print` on `dendro_plot`, it may work better on your screen to set 
`y = 0.4` and `height = 0.85`).

```{r grid-plot-4}
grid.newpage()
print(heatmap_plot, 
      vp = viewport(x = 0.4, y = 0.5, width = 0.8, height = 1.0))
print(dendro_plot, 
      vp = viewport(x = 0.90, y = 0.43, width = 0.2, height = 0.92))
```

And lastly, as a bonus, we can remove the accession numbers from the heatmap, 
as they are already displayed at the tips of the dendrogram (and we made sure 
the heatmap rows and dendrogram tips match, above). We do so by setting the 
relevant theme elements to `element_blank()` in our call to `ggplot`:

```{r heatmap-grid}
# Heatmap
heatmap_plot <- ggplot(data = otter_long, aes(x = measurement, y = accession)) +
  geom_tile(aes(fill = value)) +
  scale_fill_gradient2() +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "top")

grid.newpage()
print(heatmap_plot, 
      vp = viewport(x = 0.4, y = 0.5, width = 0.8, height = 1.0))
print(dendro_plot, 
      vp = viewport(x = 0.90, y = 0.43, width = 0.2, height = 0.92))
```

Our final script would look like this:

```{r final-script, eval = FALSE}
# Cluster & heatmap on otter data
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2017-08-15

################################################################################

# Load dependencies
library("ggplot2")
library("ggdendro")
library("tidyr")
library("grid")

# Read in data
otter <- read.csv(file = "data/otter-mandible-data.csv",
                  stringsAsFactors = TRUE)

# Restrict the data to only two species, A. cinerea & L. canadensis
two_species <- c("A. cinerea", "L. canadensis")
otter <- otter[otter$species %in% two_species, ]

# Force re-numbering of the rows after subsetting
rownames(otter) <- NULL

# Scale each measurement (independently) to have a mean of 0 and variance of 1
otter_scaled <- otter
otter_scaled[, c(4:9)] <- scale(otter_scaled[, 4:9])

# Run clustering
otter_matrix <- as.matrix(otter_scaled[, -c(1:3)])
rownames(otter_matrix) <- otter_scaled$accession
otter_dendro <- as.dendrogram(hclust(d = dist(x = otter_matrix)))

# Create dendrogram plot
dendro_plot <- ggdendrogram(data = otter_dendro, rotate = TRUE) + 
  theme(axis.text.y = element_text(size = 6))

# Heatmap

# Data wrangling
otter_long <- pivot_longer(data = otter_scaled,
                           cols = -c(species, museum, accession),
                           names_to = "measurement",
                           values_to = "value")
# Extract the order of the tips in the dendrogram
otter_order <- order.dendrogram(otter_dendro)
# Order the levels according to their position in the cluster
otter_long$accession <- factor(x = otter_long$accession,
                               levels = otter_scaled$accession[otter_order], 
                               ordered = TRUE)

# Create heatmap plot
heatmap_plot <- ggplot(data = otter_long, aes(x = measurement, y = accession)) +
  geom_tile(aes(fill = value)) +
  scale_fill_gradient2() +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "top")

# All together
grid.newpage()
print(heatmap_plot, 
      vp = viewport(x = 0.4, y = 0.5, width = 0.8, height = 1.0))
print(dendro_plot, 
      vp = viewport(x = 0.90, y = 0.43, width = 0.2, height = 0.92))
```

***
## Additional resources

+ Documentation for the [ggdendro package](https://cran.r-project.org/web/packages/ggdendro/vignettes/ggdendro.html)
+ An [example](https://stackoverflow.com/questions/6673162/reproducing-lattice-dendrogram-graph-with-ggplot2) on StackOverflow
+ A [pdf version](https://jcoliver.github.io/learn-r/008-ggplot-dendrograms-and-heatmaps.pdf) of this lesson

***

[Back to learn-r main page](index.html)

Questions?  e-mail me at <a href="mailto:jcoliver@arizona.edu">jcoliver@arizona.edu</a>.