---
title: "A very brief introduction to species distribution models in R"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
urlcolor: blue
---

Predicting ranges of species from latitude and longitude coordinates has become
increasingly easier with a suite of R packages. This introductory tutorial will 
show you how to turn your coordinate data into a range map.

#### Learning objectives

1. Install packages for species distribution modeling
2. Run species distribution models using `bioclim` approach
3. Visualize model predictions on a map

Species distribution modeling in becoming an increasingly important tool to
understand how organisms might respond to current and future environmental 
changes. There is an ever-growing number of approaches and literature for 
species distribution models (SDMs), and you are encouraged to check out the
[Additional Resources](#additional-resources) section for a few of these 
resources. The vignette for the `dismo` package is especially useful, and 
Jeremy Yoder's introduction is another great place to start. In this tutorial, 
we'll use publicly available data to build, evaluate, and visualize a 
distribution model for the saguaro cactus.

***

## Getting started

Before we do anything, we will need to make sure we have necessary software, 
set up our workspace, download example data, and install additional packages 
that are necessary to run the models and visualize their output. 

### Necessary software

The packages necessary for species distribution modeling will likely require 
additional, non-R software to work properly. Which software will depend on the 
operating system of your computer.

#### Linux

On Debian Linux systems, you will likely need to install the libgdal-dev 
package. You can do this through the terminal via 
`sudo apt-get install libgdal-dev`.

#### Windows

On Windows machines, you should probably install Rtools. You can find downloads 
and instructions at 
[https://cran.r-project.org/bin/windows/Rtools/](https://cran.r-project.org/bin/windows/Rtools/).

#### Mac OS

To use the raster package on Mac OS, you'll need to install Xcode Command Line 
Tools package. You can do this through a terminal via `xcode-select --install`.

### Workspace organization

First we need to setup our development environment. Open RStudio and create a 
new project via:

+ File > New Project...
+ Select 'New Directory'
+ For the Project Type select 'New Project'
+ For Directory name, call it something like "r-sdm" (without the quotes)
+ For the subdirectory, select somewhere you will remember (like "My Documents" 
or "Desktop")

We need to create two folders: 'data' will store the data we will be analyzing, 
and 'output' will store the results of our analyses. In the RStudio console:

```{r workspace-setup, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

It is good practice to keep input (i.e. the data) and output separate. 
Furthermore, any work that ends up in the `output` folder should be completely
disposable. That is, the combination of data and the code we write should allow 
us (or anyone else, for that matter) to reproduce any output.

### Example data

The data we are working with are observations of the 
[saguaro, _Carnegiea gigantea_](https://en.wikipedia.org/wiki/Saguaro). We are 
using a subset of records available from [GBIF](https://www.gbif.org/), the 
Global Biodiversity Information Facility. You can download the data from
[https://tinyurl.com/saguaro-obs](https://tinyurl.com/saguaro-obs); save it in 
the 'data' folder that you created in the step above.

### Install additional R packages

Next, there are _five_ additional R packages that will need to be installed:

+ dismo
+ maptools
+ rgdal
+ raster
+ sp

To install these, run:

```{r install-libraries, eval = FALSE}
install.packages("dismo")
install.packages("maptools")
install.packages("rgdal")
install.packages("raster")
install.packages("sp")
```

***

## Components of the model

The basic idea behind species distribution models is to take two sources of
information to model the conditions in which a species is expected to occur. 
The two sources of information are:

1. Occurrence data: these are usually latitude and longitude geographic 
coordinates where the species of interest has been observed. These are known as
'presence' data. Some models also make use of 'absence' data, which are 
geographic coordinates of locations where the species is known to _not_ occur.
Absence data are a bit harder to come by, but are required by some modeling
approaches. For this lesson, we will use the occurrence data of the saguaro 
that you downloaded earlier.
2. Environmental data: these are descriptors of the environment, and can 
include abiotic measurements of temperature and precipitation as well as biotic
factors, such as the presence or absence of other species (like predators,
competitors, or food sources). In this lesson we will focus on the 19 abiotic
variables available from [WorldClim](http://www.worldclim.org/bioclim). Rather 
than downloading the data from WorldClim, we'll use functions from the `dismo`
package to download these data (see below).

***

## Data and quality control

We'll start our script by loading those five libraries we need. And of course 
adding a little bit of information at the very top of our script that says what 
the script does and who is responsible!

```{r load-dependancies}
# Species distribution modeling for saguaro
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2018-02-27

library("sp")
library("raster")
library("maptools")
library("rgdal")
library("dismo")
```

There is a good chance you might have seen some red messages print out to the
screen, especially when loading the maptools or rgdal libraries. This is 
normal, and as long as none of the messages include 
"<span style="color:red">ERROR</span>", you can just hum right through those
messages. If loading the libraries _does_ result in an 
<span style="color:red">ERROR</span> message, check to see that the libraries 
were installed properly.

Now that we have those packages loaded, we can download the bioclimatic 
variable data with the `getData` function:

```{r get-bioclim-data}
bioclim.data <- getData(name = "worldclim",
                        var = "bio",
                        res = 2.5,
                        path = "data/")
```

We're giving `getData` four critical pieces of information:

1. `name = "worldclim"`: This indicates the name of the data set we would like 
to download
2. `var = "bio"`: This tells `getData` that we want to download all 19 of the
bioclimatic variables, rather than individual temperature or precipitation
measurements
3. `res = 2.5`: This is the resolution of the data we want to download; in this
case, it is 2.5 minutes of a degree. For other resolutions, you can check the
documentation by typing `?getData` into the console.
4. `path = "data/"`: Finally, this sets the location to which the files are
downloaded. In our case, it is the `data` folder we created at the beginning.

Note also that after the files are downloaded to the `data` folder, the are 
read into memory and stored in the variable called `bioclim.data`

```{r load-saguaro-data}
# Read in saguaro observations
obs.data <- read.csv(file = "data/Carnegiea-gigantea-GBIF.csv")

# Check the data to make sure it loaded correctly
summary(obs.data)
```

Notice that there are three `NA` values in the `latitude` and `longitude` 
columns. Those records will not be of any use to us, so we can remove them from 
our data frame:

```{r remove-NAs}
# Notice NAs - drop them before proceeding
obs.data <- obs.data[!is.na(obs.data$latitude), ]

# Make sure those NA's went away
summary(obs.data)
```

When we look at the `obs.data` data frame now there are no `NA` values, so we 
are ready to proceed.

To make species distribution modeling more streamlined, it is useful to have an 
idea of how widely our species is geographically distributed. We are going to 
find general latitudinal and longitudinal boundaries and store this information 
for later use:

```{r geographic-extent}
# Determine geographic extent of our data
max.lat <- ceiling(max(obs.data$latitude))
min.lat <- floor(min(obs.data$latitude))
max.lon <- ceiling(max(obs.data$longitude))
min.lon <- floor(min(obs.data$longitude))
geographic.extent <- extent(x = c(min.lon, max.lon, min.lat, max.lat))
```

Before we do any modeling, it is also a good idea to run a reality check on 
your occurrence data by plotting the points on a map.

```{r plot-occurrence-01}
# Load the data to use for our base map
data(wrld_simpl)

# Plot the base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95")

# Add the points for individual observation
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "olivedrab", 
       pch = 20, 
       cex = 0.75)
# And draw a little box around the graph
box()
```

Looking good!

***

## Building a model and visualizing results

Now that our occurrence data look OK, we can use the bioclimatic variables to 
create a model. The first thing we want to do though is limit our consideration 
to a reasonable geographic area. That is, for our purposes we are not looking 
to model saguaro habitat suitability _globally_, but rather to the general 
southwest region. So we can restrict the biolimatic variable data to the 
geographic extent of our occurrence data:

```{r bioclim-crop, error = TRUE}
# Crop bioclim data to geographic extent of saguaro
bioclim.data <- crop(x = bioclim.data, y = geographic.extent)

# Build species distribution model
bc.model <- bioclim(x = bioclim.data, p = obs.data)
```

Uh oh. That's not good. It looks like the data we passed to `bioclim` is not in 
the right format. The clue comes in the second line of the error message: 
`## Found these dimensions: 400, 3`. This is referring to the `obs.data` data 
frame, which does indeed have 400 rows and three columns. From the 
documentation from `bioclim` (see for yourself via `?bioclim` in the console): 

>**Usage**  
bioclim(x, p, ...)  
**Arguments**  
x	Raster* object or matrix  
p	two column matrix or SpatialPoints* object  
...	Additional arguments

So whatever we pass to `p` should only have **two** columns. Let's modify the
`obs.data` so it only has two columns. The first column is the GBIF identifier,
which we will not need, so we drop it using the negation operator (i.e. the 
minus sign). Then we can run the species distribution model.

```{r drop-gbif-id, error = TRUE}
# Drop unused column
obs.data <- obs.data[, c("latitude", "longitude")]

# Build species distribution model
bc.model <- bioclim(x = bioclim.data, p = obs.data)
```

What the...? OK, this error message is tougher to figure out. But let's 
consider what our `obs.data` data frame looks like now:

```{r check-obs-data}
head(obs.data)
```

The first column is latitude and the second column is longitude, which seems 
fine. That is, until we think about how R generally deals with coordinates. 
When we plot something, we generally use syntax like this:

```{r plot-syntax-example, eval = FALSE}
plot(x, y)
```

The thing to note is that the first argument we pass is data for the **x-axis** 
and the second argument is for the **y-axis**. The `bioclim` function is 
looking for data in the _same order_. That is, it looks at whatever we passed 
to `p` and assumes the first column is for the x-axis and the second column is 
for the y-axis. But our data is in the opposite order: the first column is
_latitude_, essentially the y-axis data, and our second column is longitude,
corresponding to x-axis data. So we need to reverse the column order before we 
pass `obs.data` to `bioclim`:

```{r reorder-columns}
# Reverse order of columns
obs.data <- obs.data[, c("longitude", "latitude")]

# Build species distribution model
bc.model <- bioclim(x = bioclim.data, p = obs.data)
```

Woo-hoo! No errors here (hopefully).

There's one more step we need to take before we plot the model on a map. We 
need to generate an object that has the model's probability of occurrence for
saguaros. We use the `predict` model from the `dismo` package:

```{r predict-presence}
# Predict presence from model
predict.presence <- dismo::predict(object = bc.model, 
                                   x = bioclim.data, 
                                   ext = geographic.extent)
```

You might be wondering about why we use `dismo::predict` rather than just 
`predict`. Not surprisingly, different packages sometimes use the same function 
name to perform very different operations. In the case of `predict`, there are 
at least three packages loaded into memory that have a `predict` function: 
dismo, sp, and stats. Although we _probably_ would have been fine just using
`predict` (R should have used the version from the dismo package), specifying 
the `dismo` version explicitly communicates this fact to anyone else reading 
the code. So, rather than leaving others (or your future self!) guessing, we 
can use the `dismo::predict` syntax.

Enough! It's time to plot. We start as we did before, with a blank gray map, 
add the model, and if we feel like it, add the original observations as points.

```{r plot-sdm-probabilities}
# Plot base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95")

# Add model probabilities
plot(predict.presence, add = TRUE)

# Redraw those country borders
plot(wrld_simpl, add = TRUE, border = "grey5")

# Add original observations
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "olivedrab", 
       pch = 20, 
       cex = 0.75)
box()
```

This plot shows the probability of occurrence of saguaros across the map. Note 
the values are all quite below 1.0; in fact, the maximum probability anywhere 
on the map is only `r round(predict.presence@data@max, 2)`, according to the 
model. However, we are pretty sure that saguaros are found across a pretty 
broad area of the Sonoran Desert - after all, we have the observations to prove
that! If we want our map to better reflect this, we will need to re-run our
analyses, but this time include some absence points, where saguaros are known 
to _not_ occur. The problem is, we only have presence data for saguaros.

```{r plot-occurrence-02, echo = FALSE}
# Plot the base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95",
     main = "Saguaro observations")

# Add the points for individual observation
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "olivedrab", 
       pch = 20, 
       cex = 0.75)
# And draw a little box around the graph
box()
```

## The pseudo-absence point

One common work around for coercing presence-only data for use with 
presence/absence approaches is to use pseudo-absence, or "background" points. 
While "pseudo-absence" sounds fancy, it really just means that one randomly 
samples points from a given geographic area and treats them like locations 
where the species of interest is absent. A great resource investigating the
influence and best practices of pseudo-absence points is a study by 
Barbet-Massin _et al._ (2012) (see 
[Additional Resources](#additional-resources) below for full details).

For our purposes, we are going to create a set of background (aka 
pseudo-absence) points at random, with as many points as we have observations. 
We are going to use the bioclim data files for determining spatial resolution 
of the points, and restrict the sampling area to the general region of the
observations of saguaros.

```{r pseudo-absence-sampling}
# Use the bioclim data files for sampling resolution
bil.files <- list.files(path = "data/wc2-5", 
                          pattern = "*.bil$", 
                          full.names = TRUE)

# We only need one file, so use the first one in the list of .bil files
mask <- raster(bil.files[1])

# Set the seed for the random-number generator to ensure results are similar
set.seed(20210707)

# Randomly sample points (same number as our observed points)
background <- randomPoints(mask = mask,     # Provides resolution of sampling points
                   n = nrow(obs.data),      # Number of random points
                   ext = geographic.extent, # Spatially restricts sampling
                   extf = 1.25)             # Expands sampling a little bit
```

Take a quick look at the `background` object we just created:

```{r pseudo-absence-check}
head(background)
```

We can also visualize them on a map, like we did for the observed points:

```{r pseudo-absence-plot}
# Plot the base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95",
     main = "Presence and pseudo-absence points")

# Add the background points
points(background, col = "grey30", pch = 1, cex = 0.75)

# Add the observations
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "olivedrab", 
       pch = 20, 
       cex = 0.75)

box()
```

Now that we have our pseudo-absence points, we need to take one more step. 
Getting a more traditional-range-map-looking figure requires _post hoc_ 
evaluation of the model. To do this evaluation, we are going to build the model
using only _part_ of our data (the **training** data), reserving a portion of 
the data for evaluation of the model after it is build (the **testing** data). 
We are going to reserve 20% of the data for testing, so we use the `kfold` 
function in the dismo package to evenly assign each observation to a random 
group.

```{r separate-training-testing-01}
# Arbitrarily assign group 1 as the testing data group
testing.group <- 1

# Create vector of group memberships
group.presence <- kfold(x = obs.data, k = 5) # kfold is in dismo package
```

Now pause for a minute and take a look at that `group.presence` vector we just
created:

```{r check-training-testing}
head(group.presence)

# Should see even representation in each group
table(group.presence)
```

The output of `table` shows how many points have been assigned to each of the 
five groups. In this case, we can see that the points have been evenly 
distributed, with 20% of the points in group 1, our testing group.

We use the `group.presence` vector with the observed data to separate our
observations into a training data set and a testing data set:

```{r separate-training-testing-02}
# Separate observations into training and testing groups
presence.train <- obs.data[group.presence != testing.group, ]
presence.test <- obs.data[group.presence == testing.group, ]

# Repeat the process for pseudo-absence points
group.background <- kfold(x = background, k = 5)
background.train <- background[group.background != testing.group, ]
background.test <- background[group.background == testing.group, ]
```

## Training and testing the model

Now that we have (1) our pseudo-absence points and (2) separate training and 
testing data, we can re-build the model, evaluate its performance, and draw a 
more aesthetically pleasing map. We build the model with the `bioclim` function 
as before, but instead of using all the observations in `obs.data` we only use 
the training data stored in `presence.train`:

```{r bioclim-training}
# Build a model using training data
bc.model <- bioclim(x = bioclim.data, p = presence.train)

# Predict presence from model (same as previously, but with the update model)
predict.presence <- dismo::predict(object = bc.model, 
                                   x = bioclim.data, 
                                   ext = geographic.extent)
```

We now take that model, and evaluate it using the observation data and the
pseudo-absence points we reserved for model _testing_. We then use this test to
establish a cutoff of occurrence probability to determine the boundaries of the
saguaro range.

```{r testing-model}
# Use testing data for model evaluation
bc.eval <- evaluate(p = presence.test,   # The presence testing data
                    a = background.test, # The absence testing data
                    model = bc.model,    # The model we are evaluating
                    x = bioclim.data)    # Climatic variables for use by model

# Determine minimum threshold for "presence"
bc.threshold <- threshold(x = bc.eval, stat = "spec_sens")
```

The `threshold` function offers a number of means of determining the threshold
cutoff through the `stat` parameter. Here we chose `"spec_sens"`, which sets 
"the threshold at which the sum of the sensitivity (true positive rate) and
specificity (true negative rate) is highest." For more information, check out 
the documentation for `threshold` (`?threshold`, remember?).
  
And _finally_, we can use that threshold to paint a map with the predicted 
range of the saguaro!

```{r plot-sdm-threshold-bad-color}
# Plot base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95")

# Only plot areas where probability of occurrence is greater than the threshold
plot(predict.presence > bc.threshold, 
     add = TRUE, 
     legend = FALSE, 
     col = "olivedrab")

# And add those observations
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "black",
       pch = "+", 
       cex = 0.75)

# Redraw those country borders
plot(wrld_simpl, add = TRUE, border = "grey5")
box()
```

Hmmm...that doesn't look right. It plotted a large portion of the map green. 
Let's look at what we actually asked R to plot, that is, we plot the value of
`predict.presence > bc.threshold`. So what is that?

```{r investigate-raster-comparison}
predict.presence > bc.threshold
```

The comparison of these two rasters produces another raster with values of only 
0 or 1: 0 where the comparison evaluates as `FALSE` (i.e., when the value in a 
grid cell of `predict.presence` is less than or equal to the value in the
corresponding grid cell of `bc.threshold`) and 1 where the comparison evaluates 
at `TRUE`. Since there are two values in this comparison (the 0 and 1 in the
`values` field), we need to update what we pass to the `col` parameter in our 
plot call. Instead of just passing a single value, we provide a color for 0 
(`NA`) and a color for 1 (`"olivedrab"`):

```{r plot-sdm-threshold-good-color}
# Plot base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95")

# Only plot areas where probability of occurrence is greater than the threshold
plot(predict.presence > bc.threshold, 
     add = TRUE, 
     legend = FALSE, 
     col = c(NA, "olivedrab"))

# And add those observations
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "black",
       pch = "+", 
       cex = 0.75)

# Redraw those country borders
plot(wrld_simpl, add = TRUE, border = "grey5")
box()
```

A final note on our approach: the map we have drawn presents a categorical
classification of whether a particular point on the landscape will be suitable 
or not for the species of interest. This classification relies quite heavily on 
the value of the threshold (see `bc.threshold` and the documentation for
`threshold`) _and_ the pseudo-absence points. Given that we used random 
sampling to generate those pseudo-absence points, there is potential for 
variation in the predicted range if you run this code more than once (try it! 
if you re-run the code from the point of creating the pseudo-absence points, 
you are almost guaranteed a different map.). There are a number of approaches 
to dealing with this variation, and the paper by 
[Barbet-Massin et al. (2012)](https://dx.doi.org/10.1111/j.2041-210X.2011.00172.x)
is a great resource. I'll leave it as homework for you to determine which 
approach is most appropriate here!

Our final script, generating the model, determining the threshold, and 
visualizing the results:

```{r final-script, eval = FALSE}
# Species distribution modeling for saguaro
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2018-02-27

rm(list = ls())

# Load additional packages
library("sp")
library("raster")
library("maptools")
library("rgdal")
library("dismo")

# Download bioclim data and store in bioclim.data variable
bioclim.data <- getData(name = "worldclim",
                        var = "bio",
                        res = 2.5,
                        path = "data/")

# Read in saguaro observations
obs.data <- read.csv(file = "data/Carnegiea-gigantea-GBIF.csv")

# Drop any rows with NAs
obs.data <- obs.data[!is.na(obs.data$latitude), ]

# Only pull out those columns of interest and in the order we want them
obs.data <- obs.data[, c("longitude", "latitude")]

# Determine geographic extent of our data
max.lat = ceiling(max(obs.data$latitude))
min.lat = floor(min(obs.data$latitude))
max.lon = ceiling(max(obs.data$longitude))
min.lon = floor(min(obs.data$longitude))
geographic.extent <- extent(x = c(min.lon, max.lon, min.lat, max.lat))

# Crop the bioclim data to geographic extent of saguaro
bioclim.data <- crop(x = bioclim.data, y = geographic.extent)

# Create pseudo-absence, or background, points
# Use the bioclim data files for sampling resolution
bil.files <- list.files(path = "data/wc2-5", 
                          pattern = "*.bil$", 
                          full.names = TRUE)

# We only need one file, so use the first one in the list of .bil files
mask <- raster(bil.files[1])

# Randomly sample points (same number as our observed points)
background <- randomPoints(mask = mask,     # Provides resolution of sampling points
                   n = nrow(obs.data),      # Number of random points
                   ext = geographic.extent, # Spatially restricts sampling
                   extf = 1.25)             # Expands sampling a little bit

# Arbitrarily assign group 1 as the testing data group
testing.group <- 1

# Create vector of group memberships
group.presence <- kfold(x = obs.data, k = 5) # kfold is in dismo package

# Separate observations into training and testing groups
presence.train <- obs.data[group.presence != testing.group, ]
presence.test <- obs.data[group.presence == testing.group, ]

# Repeat the process for pseudo-absence points
group.background <- kfold(x = background, k = 5)
background.train <- background[group.background != testing.group, ]
background.test <- background[group.background == testing.group, ]

# Build a model using training data
bc.model <- bioclim(x = bioclim.data, p = presence.train)

# Predict presence from model
predict.presence <- dismo::predict(object = bc.model, 
                                   x = bioclim.data, 
                                   ext = geographic.extent)

# Use testing data for model evaluation
bc.eval <- evaluate(p = presence.test,   # The presence testing data
                    a = background.test, # The absence testing data
                    model = bc.model,    # The model we are evaluating
                    x = bioclim.data)    # Climatic variables for use by model

# Determine minimum threshold for "presence"
bc.threshold <- threshold(x = bc.eval, stat = "spec_sens")

# Load map data for plotting
data(wrld_simpl)

# Plot base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95")

# Only plot areas where probability of occurrence is greater than the threshold
plot(predict.presence > bc.threshold, 
     add = TRUE, 
     legend = FALSE, 
     col = c(NA, "olivedrab"))

# And add those observations
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "black",
       pch = "+", 
       cex = 0.6)

# Redraw those country borders
plot(wrld_simpl, add = TRUE, border = "grey5")
box()
```

***

## Advanced: Forecasting distributions

Now that you have a species distribution model, you can make predictions about 
the distribution under different climate scenarios. Let us pause for a moment 
and be very clear about this approach. With all kinds of math wizardry on our 
side, we are attempting to predict the future. Which means _any_ predictions we 
make should be interpreted with extreme caution. If you are going to go about 
an approach such as this, it would be wise to run a variety of different models
and a variety of different climate scenarios. There are links to such resources 
in the [Additional Resources](#additional-resources) section, below.

### Forecast climate data

We will need to download climate data for the time period of interest. For the 
purposes of this lesson, we will look a climate projections for the year 2070. 
We will use the GFDL-ESM2G model with an RCP of 4.5 CO2. For an examination of 
different forecast models, see 
[McSweeney et al. 2015](https://link.springer.com/article/10.1007/s00382-014-2418-8). 
We can download the data like we did before, using the raster package's 
`getData()` function, and send it do the data folder:

```{r download-forecast-real, echo = FALSE}
suppressWarnings(expr = {
forecast.data <- getData(name = "CMIP5", # forecast data
                         var = "bio",    # bioclim
                         res = 2.5,      # 2.5 minute resolution
                         path = "data/", # destination directory
                         model = "GD",   # GFDL-ESM2G
                         rcp = "45",     # CO2 increase 4.5
                         year = 70)      # 2070
})
```

```{r download-forecast-example, eval = FALSE}
forecast.data <- getData(name = "CMIP5", # forecast data
                         var = "bio",    # bioclim
                         res = 2.5,      # 2.5 minute resolution
                         path = "data/", # destination directory
                         model = "GD",   # GFDL-ESM2G
                         rcp = "45",     # CO2 increase 4.5
                         year = 70)      # 2070
```

You may see warning messages about "Discarded datum unknown in Proj4 
definition", but we can ignore those for now.

**Note:** As of this writing (2021-07-07), the forecast data download described 
above retrieves the CMIP5 data, which are now obsolete. The [CMIP6](https://pcmdi.llnl.gov/CMIP6/) 
data are now available at [https://www.worldclim.org/data/cmip6/cmip6climate.html](https://www.worldclim.org/data/cmip6/cmip6climate.html) 
and users are recommended to download the bioclim data directly from the site. 
You can find the corresponding 2.5 minute resolution data at [https://www.worldclim.org/data/cmip6/cmip6_clim2.5m.html](https://www.worldclim.org/data/cmip6/cmip6_clim2.5m.html); bioclim data are indicated as "bc".

We need to do one more thing before we can use our data, and that is to make 
sure the names in our model of bioclimatic variables line up with the names in 
the forecast data object

```{r name-forecast}
names(forecast.data) <- names(bioclim.data)
```

### Get out the crystal ball

Now that we have the forecast data, we can apply the model we build above, 
`bc.model`, to the forecast climate data:

```{r predict-forecast}
# Predict presence from model with forecast data
forecast.presence <- dismo::predict(object = bc.model, 
                                    x = forecast.data, 
                                    ext = geographic.extent)

```

If you want to look at the predicted probabilities of occurrence, you can 
modify the code we used above.

```{r map-predicted-probs}
# Plot base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95")

# Add model probabilities
plot(forecast.presence, add = TRUE)

# Redraw those country borders
plot(wrld_simpl, add = TRUE, border = "grey5")

# Add original observations
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "olivedrab", 
       pch = 20, 
       cex = 0.75)
box()
```

We can also map our predictions for presence / absence, using the same 
threshold that we did for predictions based on current climate data.

```{r map-predicted-pa}
# Plot base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95")

# Only plot areas where probability of occurrence is greater than the threshold
plot(forecast.presence > bc.threshold, 
     add = TRUE, 
     legend = FALSE, 
     col = c(NA, "olivedrab"))

# And add those observations
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "black",
       pch = "+", 
       cex = 0.6)

# Redraw those country borders
plot(wrld_simpl, add = TRUE, border = "grey5")
box()
```

Oof. Things do not look great for saguaros under this climate forecast. Try 
downloading other climate models to see how predictions differ. And remember to 
interpret these results cautiously.

***

## Additional resources

+ [Vignette for dismo package](https://rspatial.org/raster/sdm/)
+ [Species distribution models in R](http://www.molecularecologist.com/2013/04/species-distribution-models-in-r/)
+ [Fast and flexible Bayesian species distribution modelling using Gaussian processes](http://onlinelibrary.wiley.com/doi/10.1111/2041-210X.12523/pdf)
+ [Run a range of species distribution models](https://rdrr.io/cran/biomod2/man/BIOMOD_Modeling.html)
+ [SDM polygons on a Google map](https://rdrr.io/rforge/dismo/man/gmap.html)
+ [R package 'maxnet' for functionality of Java maxent package](https://cran.r-project.org/web/packages/maxnet/maxnet.pdf)
+ A [study on the effect of pseudo-absences in SDMs (Barbet-Massin et al. 2012)](https://dx.doi.org/10.1111/j.2041-210X.2011.00172.x)
+ A [PDF version](https://jcoliver.github.io/learn-r/011-species-distribution-models.pdf) of this lesson

***

[Back to learn-r main page](index.html)

Questions?  e-mail me at <a href="mailto:jcoliver@arizona.edu">jcoliver@arizona.edu</a>.