---
title: "A very brief introduction to species distribution models in R"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
---

Outline:

+ What they are (introductory stuff + description)
+ What do we need?
    + Data
    + Dependencies
+ Preparing data?
+ Quick plot of points on map
+ Pseudo-absence points
+ Models
    + Quick glance at loadings?
+ Maps
    + Do just probability map, without any pseudo-absence points
    + Do threshold map, with pseudo-absence points

[INTRODUCTORY SENTENCE]

#### Learning objectives
1. Install packages for species distribution modeling
2. Run species distribution models using `bioclim` approach
3. Visualize model predictions on a map

## [DESCRIPTION OR MOTIVATION; 2-4 sentences that would be used for an announcement]

***

## Getting started
Before we do anything, we will need to set up our workspace, download example data, and install additional packages that are necessary to run the models and visualize their output. 
### Workspace organization
So, to start, create a pair of folders in your workspace:
```{r workspace-setup, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```
It is good practice to keep input (i.e. the data) and output separate. Furthermore, any work that ends up in the 'output' folder should be completely disposable. That is, the combination of data and the code we write should allow us (or anyone else, for that matter) to reproduce any output.

### Example data
The data we are working with are observations of the Saguaro, _Carnegiea gigantea_, downloaded from [GBIF](https://www.gbif.org/), the Global Biodiversity Information Facility. You can download the data from [](); save it in the 'data' folder that you created in the step above.

### Install additional R packages
Next, there are _five_ additional R packages that will need to be installed:

+ dismo
+ maptools
+ rgdal
+ raster
+ sp

To install these, run:
```{r install-dependencies, eval = FALSE}
install.packages("dismo")
install.packages("maptools")
install.packages("rgdal")
install.packages("raster")
install.packages("sp")
```

***

## [TOPIC ONE]

***

## [TOPIC TWO]

***


```{r}
library("sp") # required by raster
library("raster") # extent
library("maptools") # wrld_simpl
library("rgdal")
library("dismo")

obs.data <- read.delim(file = "data/Carnegiea-gigantea-GBIF.csv")
summary(obs.data)
obs.data <- obs.data[!is.na(obs.data$latitude), ]
obs.data <- obs.data[, -1]
obs.data <- obs.data[sample(x = nrow(obs.data), size = 400), ] # TODO: Fix original data so there are only 400 observations

# Determine geographic extent of our data
max.lat = ceiling(max(obs.data$latitude))
min.lat = floor(min(obs.data$latitude))
max.lon = ceiling(max(obs.data$longitude))
min.lon = floor(min(obs.data$longitude))
geographic.extent <- extent(x = c(min.lon, max.lon, min.lat, max.lat))

# Reality check - plot points
data(wrld_simpl)
# Debugging note: Have to run multiple successive plotting lines all at once for it to work
plot(wrld_simpl, xlim = c(min.lon, max.lon), ylim = c(min.lat, max.lat), axes = TRUE, col = "lightyellow")
points(obs.data$longitude, obs.data$latitude, col = "olivedrab", pch = 20, cex = 0.75)
box()

# Get the biolim data
bioclim.data <- getData(name = "worldclim",
                        var = "bio",
                        res = 2.5,
                        path = "data/")
bioclim.data <- crop(x = bioclim.data, y = geographic.extent)

# TODO: run this without re-ordering first
# Have to swap order of latitude & longitude, as bioclim method 
# expects first column to be longitude, second to be latitude
obs.data <- obs.data[, c(2:1)]

# Do species distribution model and draw a plot
bc.model <- bioclim(x = bioclim.data, p = obs.data)
predict.presence <- predict(x = bioclim.data, object = bc.model, ext = geographic.extent, progress = "")

# TODO: May need some more graphics massaging here
plot(predict.presence, xlim = c(min.lon, max.lon), ylim = c(min.lat, max.lat), main = "BIOCLIM, raw")
plot(wrld_simpl, add = TRUE, border = "dark grey")
points(obs.data$longitude, obs.data$latitude, col = "olivedrab", pch = "+", cex = 0.75)
box()

# Create pseudo-absence points
bil.files <- list.files(path = "data/wc2-5", 
                          pattern = "*.bil$", 
                          full.names = TRUE)
mask <- raster(bil.files[1])

# Random points (same number as our observed points)
bg <- randomPoints(mask = mask, n = nrow(obs.data), ext = geographic.extent, extf = 1.25)

# Do a quick plot to see the points
plot(wrld_simpl, xlim = c(min.lon, max.lon), ylim = c(min.lat, max.lat), axes = TRUE, col = "light yellow")
points(bg, col = "grey30", pch = "+", cex = 0.75)
points(obs.data, col = "olivedrab", pch = 20, cex = 0.75)
box()

# Now re-run model with only the training data

# Data for observation sites (presence and background)
presence.values <- extract(x = bioclim.data, y = obs.data)
absence.values <- extract(x = bioclim.data, y = bg)

# Separate training & testing data
group.presence <- kfold(x = obs.data, k = 5) # kfold is in dismo package
testing.group <- 1
presence.train <- obs.data[group.presence != testing.group, ]
presence.test <- obs.data[group.presence == testing.group, ]
group.bg <- kfold(x = bg, k = 5)
bg.train <- bg[group.bg != testing.group, ]
bg.test <- bg[group.bg == testing.group, ]

# Do species distribution model and draw a plot
bc.model <- bioclim(x = bioclim.data, p = presence.train)


bc.eval <- evaluate(p = presence.test, a = bg.test, model = bc.model, x = bioclim.data)
bc.threshold <- threshold(x = bc.eval, stat = "spec_sens")
predict.presence <- predict(x = bioclim.data, object = bc.model, ext = geographic.extent, progress = "")
plot(predict.presence > bc.threshold, main = "Presence/Absence")
plot(wrld_simpl, add = TRUE, border = "dark grey")
box()
```

***

## Additional resources

+ [Vignette for `dismo` package](https://cran.r-project.org/web/packages/dismo/vignettes/sdm.pdf)
+ [Fast and flexible Bayesian species distribution modelling using Gaussian processes](http://onlinelibrary.wiley.com/doi/10.1111/2041-210X.12523/pdf)
+ [Species distribution models in R](http://www.molecularecologist.com/2013/04/species-distribution-models-in-r/)
+ [Run a range of species distribution models](https://rdrr.io/cran/biomod2/man/BIOMOD_Modeling.html)
+ [SDM polygons on a Google map](https://rdrr.io/rforge/dismo/man/gmap.html)
+ [R package 'maxnet' for functionality of Java maxent package](https://cran.r-project.org/web/packages/maxnet/maxnet.pdf)
+ A [study on the effect of pseudo-absences in SDMs (Barbet-Massin et al. 2012)](https://dx.doi.org/10.1111/j.2041-210X.2011.00172.x)
+ A [PDF version](https://jcoliver.github.io/learn-r/011-species-distribution-models.pdf) of this lesson

***

<a href="index.html">Back to learn-r main page</a>
  
Questions?  e-mail me at <a href="mailto:jcoliver@email.arizona.edu">jcoliver@email.arizona.edu</a>.