---
title: "Introduction to tidyverse packages"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
urlcolor: blue
---

Make your life easier and your code faster with the suite of "tidyverse" 
packages, including ggplot, tidyr, and dplyr.

#### Learning objectives
1. Manipulate data with `group_by` and `summarize` to extract information from
datasets
2. Combine multiple commands with piping functionality
3. Create publication-quality data visualizations with `ggplot`

## Data science: more fun, less pain

R is a powerful language for managing, analyzing, and visualizing complex data.
However, some of the commands in R are esoteric or just plain confusing. The
[tidyverse](https://www.tidyverse.org/) package for R includes several helpful
functions that make it easier to manipulate, summarize, and visualize your 
data. In this lesson we'll use these functions to create plots from summary
statistics.

***

## Getting started

First we need to setup our development environment. Open RStudio and create a 
new project via:

+ File > New Project...
+ Select 'New Directory'
+ For the Project Type select 'New Project'
+ For Directory name, call it something like "r-tidyverse" (without the quotes)
+ For the subdirectory, select somewhere you will remember (like "My Documents" 
or "Desktop")

We are using the tidyverse package, which itself is really just a collection of 
six different packages. However, we can install them all with one command:

```{r install-tidyverse, eval = FALSE}
install.packages("tidyverse")
```

Our ultimate goal is to use the pre-loaded `iris` data to create a plot of the 
data stored in that data frame. The `iris` data were collected by botanist 
Edgar Anderson and used in the early statistical work of R.A. Fisher. 

We want to make a chart that looks like this:

```{r load-tidyverse, echo = FALSE}
suppressMessages(library("tidyverse"))
```

```{r goal-graphic, echo = FALSE}
iris %>% 
  pivot_longer(cols = -Species, 
         names_to = "trait", 
         values_to = "measurement") %>%
  group_by(Species, trait) %>%
  summarize(trait.mean = mean(measurement),
            trait.se = sd(measurement/sqrt(n()))) %>%
  ggplot(mapping = aes(x = Species, y = trait.mean)) +
  geom_point() +
  geom_errorbar(mapping = aes(ymin = trait.mean - trait.se,
                              ymax = trait.mean + trait.se),
                width = 0.3) +
  ylab(label = "Trait mean values") +
  facet_wrap(~ trait, scales = "free_y")
```

We want to keep all of our work stored in an R script, so we open a new script 
and start with an explanation of what we are going to do:

```{r script-header, eval = FALSE}
# Plot iris trait measurements
# Jeffrey C. Oliver
# jcoliver@email.arizona.edu
# 2018-05-17
```

***

## Summarizing the data

So what will we need? Break this down into the component parts.

+ Means
+ Standard errors
+ For each species
+ For each trait

### The hard way

Let's start with getting the means for a single column, `Sepal.Length`. If we 
do this in base R, we need to pull out the values for _each_ species, then 
calculate the means for each. This looks something like this:

```{r species-means-base-R, eval = FALSE}
setosa.mean <- mean(iris$Sepal.Length[iris$Species == "setosa"])
versicolor.mean <- mean(iris$Sepal.Length[iris$Species == "versicolor"])
virginica.mean <- mean(iris$Sepal.Length[iris$Species == "virginica"])
```

Which is a little cumbersome, especially if we also need to do the additional 
step of putting all these means into a single data frame. 

```{r build-means-df, eval = FALSE}
# Add these back into a data.frame
iris.means <- data.frame(Species = c("setosa", "versicolor", "virginica"),
                         SL.mean = c(setosa.mean, versicolor.mean, virginica.mean))
```

### There's a better way

A pair of commands can make this much easier: `group_by` and `summarize`. The 
first, `group_by` imposes structure on our data; for our immediate purposes, we 
will use it to group the data by the `Species` column:

```{r group-by-species-quiet, echo = FALSE}
iris.grouped <- group_by(iris, Species)
```

```{r group-by-species, echo = TRUE, eval = FALSE}
# Load the tidyverse packages
library("tidyverse")

# Group the iris data by values in the Species column
iris.grouped <- group_by(iris, Species)
```

If we look at the first few rows of these data with the `head` command, 

```{r head-grouped}
head(iris.grouped)
```

it looks similar to the `iris` data we started with, but now, instead of a
data.frame, this is actually a `tibble`. We don't need to worry much about that 
now, only to notice that Species is listed as a group and that below the column
names is an indication of the data types (there are only numbers `<dbl>` and 
factors `<fct>`).

The second function we want to use is `summarize`, which does exactly that: it
provides some summary of the data we pass to it. Let's get the mean value of 
sepal length for each species:

```{r sepal-length-means}
iris.means <- summarise(iris.grouped, SL.mean = mean(Sepal.Length))
iris.means
```

Note that we did _not_ have to tell `summarize` to calculate the mean for each
species separately. As part of the `tidyverse` package, `summarize` knows how 
to deal with grouped data.

These two functions made it easier, after all we went from this:

```{r species-means-base-R-redux, eval = FALSE}
# Calcuate the mean for each species
setosa.mean <- mean(iris$Sepal.Length[iris$Species == "setosa"])
versicolor.mean <- mean(iris$Sepal.Length[iris$Species == "versicolor"])
virginica.mean <- mean(iris$Sepal.Length[iris$Species == "virginica"])

# Add these back into a data.frame
iris.means <- data.frame(Species = c("setosa", "versicolor", "virginica"),
                         SL.mean = c(setosa.mean, versicolor.mean, virginica.mean))
```

To this:

```{r grouped-means, eval = FALSE}
iris.grouped <- group_by(iris, Species)
iris.means <- summarise(iris.grouped, SL.mean = mean(Sepal.Length))
```

But there is another operator that can make our life even easier. If you are
familiar with the bash shell, you might be familiar with the pipe character, 
`|`, which is used to re-direct output. A similar operator in R is `%>%` and is 
used to send whatever is on the left-side of the operator to the first argument 
of the function on the right-side of the operator. So, these two statements are
effectively the same:

```{r pipe-intro, eval = FALSE}
# This:
iris %>% group_by(Species)
# is the same as:
group_by(iris, Species)
```

But here comes the really cool part! We can chain these pipes together in a 
string of commands, sending the output of one command directly to the next. So
instead of the two-step process we used to first group the data by species, 
then calculate the means, we can do it all at once with pipes:

```{r pipe-group-summarize}
iris.means <- iris %>%
  group_by(Species) %>%
  summarize(SL.mean = mean(Sepal.Length))
```

Let's break apart what we just did, line by line:

+ `iris.means <- iris %>%` We did two things here. First, we instructed R to 
assign the final output to the variable `iris.means` _and_ we sent the `iris` 
data to whatever command is coming on the next line.
+ `group_by(Species)` This line is effectively the same as `group_by(.data = 
iris, Species)`, because we sent `iris` data to `group_by` through the pipe, 
`%>%`. We then sent _this_ grouped data to the next line.
+ `summarize(SL.mean = mean(Sepal.Length))` This used the grouped data from the
preceding `group_by` command to calculate the mean values of sepal length for 
each species.
+ The final output of `summarize` was then assigned to the variable 
`iris.means`.

Remember our plot:

```{r echo = FALSE}
iris %>% 
  pivot_longer(cols = -Species, 
         names_to = "trait", 
         values_to = "measurement") %>%
  group_by(Species, trait) %>%
  summarize(trait.mean = mean(measurement),
            trait.se = sd(measurement/sqrt(n()))) %>%
  ggplot(mapping = aes(x = Species, y = trait.mean)) +
  geom_point() +
  geom_errorbar(mapping = aes(ymin = trait.mean - trait.se,
                              ymax = trait.mean + trait.se),
                width = 0.3) +
  ylab(label = "Trait mean values") +
  facet_wrap(~ trait, scales = "free_y")
```

Where are we with our necessary components?

+ Means
+ Standard errors
+ For each species
+ For each trait

Well, we have the means for each species, but we don't have the standard errors 
and we only have data for one trait (sepal length). Let's start by calculating 
the standard error. Remember the formula for the standard error is the standard
deviation divided by the square root of the sample size:

$$
SE = \frac{\sigma}{\sqrt{n}}
$$
Base R has the function `sd` which calculates the standard deviation, but we 
need another function from tidyverse, `n`, which counts the number of 
observations in the current group. So to caluclate the standard error, we can 
use `sd(Sepal.Length)/sqrt(n())`. But where? It turns out that `summarize` is 
not restricted to a single calculation. That is, we can summarize data in 
multiple ways with a single call to `summarize`. We can update our previous 
code to include a column for standard errors in our output:

```{r summarize-mult}
iris.means <- iris %>%
  group_by(Species) %>%
  summarize(SL.mean = mean(Sepal.Length),
            SL.se = sd(Sepal.Length)/sqrt(n()))
iris.means
```

***

## Visualize!

At this point, we should go ahead and start trying to plot our data. Another part of the tidyverse package is `ggplot2`, a great package for making high-quality visualizations. `ggplot2` uses special syntax for making graphs. We start by telling R _what_ we want to plot in the graph:

```{r plot-sepal-length-empty}
ggplot(data = iris.means, mapping = aes(x = Species, y = SL.mean))
```

But our graph is empty! This is because we did not tell R _how_ to plot the 
data. That is, do we want a bar chart? A scatterplot? Maybe a heatmap? We are 
going to plot the means as points, so we use `geom_point()`. Note also the
specialized syntax where we add components to our plot with the plus sign, "+":

```{r plot-sepal-length-points}
ggplot(data = iris.means, mapping = aes(x = Species, y = SL.mean)) +
  geom_point()
```

Great! So now we also need to add those error bars. We'll use another 
component, `geom_errorbar` to do this.

```{r plot-sepal-length-errorbar}
ggplot(data = iris.means, mapping = aes(x = Species, y = SL.mean)) +
  geom_point() +
  geom_errorbar(mapping = aes(ymin = SL.mean - SL.se,
                              ymax = SL.mean + SL.se))
```

Note that for the error bars, the calculations for the positions of the bars (1
standard error above and below the mean) are actually performed _inside_ the
`ggplot` command.

Those error bars are a little outrageous, so let's make them narrower:

```{r plot-sepal-length-errorbar-nicer}
ggplot(data = iris.means, mapping = aes(x = Species, y = SL.mean)) +
  geom_point() +
  geom_errorbar(mapping = aes(ymin = SL.mean - SL.se,
                              ymax = SL.mean + SL.se),
                width = 0.3)
```

Our plot looks good for now. Let's move on to getting all our traits in a 
single graph.

***

## The _long_ way

The iris data are organized like this:

```{r show-wide, echo = FALSE}
head(iris)
```

But in order to capitalize on `ggplot` functionality, we need to reorganize the 
data so each row only has data for a _single trait_, like this:

```{r show-long, echo = FALSE}
head(pivot_longer(data = iris, 
                  cols = -Species, 
                  names_to = "trait", 
                  values_to = "measurement"))
```

This is known as "long" format, where each row only has a single trait 
observation. To make this data conversion, we use the the `pivot_longer` 
function:

```{r pivot-longer-intro}
iris.long <- pivot_longer(data = iris, 
                    cols = -Species,
                    names_to = "trait",
                    values_to = "measurement")
```
The arguments we pass to `pivot_longer` are:

+ `data = iris` this indicates `iris` is the data frame we want to transform
+ `cols = -Species` tells `pivot_longer` _not_ to treat the value in the 
`Species` column as a separate variable
+ `names_to = "trait"` `trait` is the column name for the variable names (e.g. "Sepal.Length", "Sepal.Width", etc.)
+ `values_to = "measurement"` `measurement` is the column name for the actual 
values 

Let's take this `pivot_longer` functionality and combine it with the `group_by` 
and `summarize` commands we used previously. Recall our earlier code to 
generate species' means and standard errors:

```{r group-summarize-reminder, eval = FALSE}
iris.means <- iris %>%
  group_by(Species) %>%
  summarize(SL.mean = mean(Sepal.Length),
            SL.se = sd(Sepal.Length)/sqrt(n()))
```

We'll want to update this, inserting the `pivot_longer` function and updating 
the values used for calculating the mean and standard deviation:

```{r by-species-stats}
iris.means <- iris %>%
  pivot_longer(cols = -Species, 
               names_to = "trait", 
               values_to = "measurement") %>%
  group_by(Species, trait) %>%
  summarize(trait.mean = mean(measurement),
            trait.se = sd(measurement)/sqrt(n()))
```
Note the insertion of `pivot_longer` and the changes to `group_by` and 
`summarize`:

+ `group_by`: We add an additional term, `trait`, to indicate to create another
grouping, based on each trait
+ `summarize`: We replace `SL.mean` with `trait.mean` and `SL.se` with 
`trait.se`

Our data frame now has a row for each species and each trait:

```{r show-long-stats, echo = FALSE}
iris.means
```

Great! So now we need to use these summary statistics to create our plot. 
Recall the code we used to plot the sepal lengths:

```{r plot-reminder, eval = FALSE}
ggplot(data = iris.means, mapping = aes(x = Species, y = SL.mean)) +
  geom_point() +
  geom_errorbar(mapping = aes(ymin = SL.mean - SL.se,
                              ymax = SL.mean + SL.se),
                width = 0.3)
```

We need to update:

+ The specification of what to plot on the y-axis in the `ggplot` function
    + In `ggplot` command: change `SL.mean` to `trait.mean`
+ The values for error bar boundaries
    + In `geom_errorbar` command: replace `SL.mean` with `trait.mean` and 
    `SL.se` with `trait.se`

```{r plot-all-traits-in-one}
ggplot(data = iris.means, mapping = aes(x = Species, y = trait.mean)) +
  geom_point() +
  geom_errorbar(mapping = aes(ymin = trait.mean - trait.se,
                              ymax = trait.mean + trait.se),
                width = 0.3)
```

Something isn't quite right. We actually want four separate charts, one for 
each of the traits. To do so, we need to tell R how to break apart the data 
into separate charts. We do this with the `facet_wrap` component of `ggplot`:

```{r plot-all-traits-one-y}
ggplot(data = iris.means, mapping = aes(x = Species, y = trait.mean)) +
  geom_point() +
  geom_errorbar(mapping = aes(ymin = trait.mean - trait.se,
                              ymax = trait.mean + trait.se),
                width = 0.3) +
  facet_wrap(~ trait)
```

OK, there are a few more things we want to change:

1. The names of each subplot reflect the trait names, which were column names 
in the original data. Let's update the values so the two words for each trait 
are separated by a period, not a space (e.g. "Petal.Width" becomes "Petal 
Width").
2. We should make that Y-axis title a little nicer. We'll use `ylab` for that.
3. All four charts are using the same y-axis scale; note all the petal width 
values are below 3, but the maximum value of the y-axis is 6. Since we won't be
doing comparisons on actual values _between_ the charts, we can give each chart 
its own, independent y-axis scale. We'll add this information to the 
`facet_wrap` command.

```{r plot-all-traits}
# Update trait names, replacing period with space
iris.means$trait <- gsub(pattern = ".",
                         replacement = " ",
                         x = iris.means$trait,
                         fixed = TRUE) # if fixed = FALSE, evaluates as regex

ggplot(data = iris.means, mapping = aes(x = Species, y = trait.mean)) +
  geom_point() +
  geom_errorbar(mapping = aes(ymin = trait.mean - trait.se,
                              ymax = trait.mean + trait.se),
                width = 0.3) +
  ylab(label = "Trait mean values") + # update the y-axis title
  facet_wrap(~ trait, scales = "free_y") # allow y-axis scale to vary
```

With the use of tidyverse functions, we created a publication-quality graphic 
with just a few lines of code.

Our final script looks like:

```{r final-script, eval = FALSE}
# Plot iris trait measurements
# Jeffrey C. Oliver
# jcoliver@email.arizona.edu
# 2018-05-17

rm(list = ls())

# Load the tidyverse packages
library("tidyverse")

# Create data of summary statistics
iris.means <- iris %>%
  pivot_longer(cols = -Species,
               names_to = "trait",
               values_to = "measurement") %>%
  group_by(Species, trait) %>%
  summarize(trait.mean = mean(measurement),
            trait.se = sd(measurement)/sqrt(n()))

# Update trait names, replacing period with space
iris.means$trait <- gsub(pattern = ".",
                         replacement = " ",
                         x = iris.means$trait,
                         fixed = TRUE)

# Plot each trait separately
ggplot(data = iris.means, mapping = aes(x = Species, y = trait.mean)) +
  geom_point() +
  geom_errorbar(mapping = aes(ymin = trait.mean - trait.se,
                              ymax = trait.mean + trait.se),
                width = 0.3) +
  ylab(label = "Trait mean values") +
  facet_wrap(~ trait, scales = "free_y")
```

***

## Additional resources

+ Official page for the [tidyverse](https://www.tidyverse.org/) package
+ [Cheatsheet for data wrangling with the dpylr package](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf)
+ An [opinionated discussion about "tidy" data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
+ A [PDF version](https://jcoliver.github.io/learn-r/012-intro-tidyverse.pdf) of this lesson
+ Software Carpentry lessons on [dplyr](http://swcarpentry.github.io/r-novice-gapminder/13-dplyr/), [tidyr](http://swcarpentry.github.io/r-novice-gapminder/14-tidyr/), and [ggplot2](http://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2/) packages

***

[Back to learn-r main page](index.html)

Questions?  e-mail me at <a href="mailto:jcoliver@arizona.edu">jcoliver@arizona.edu</a>.