---
title: "LESSON TITLE"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

[INTRODUCTORY SENTENCE]

General idea:

Load some data: temperature and precipitation
Look at the temperature (via console & `plot()`)
Crop to southwest & recolor (YlOrBr is good for temperature)
> plot(temp_son, col = rev(hcl.colors(50, palette = "YlOrBr")))


Rough extent for Sonoran desert: -116, -110, 28, 35
Do data transformation (like converting C to F)

Categorize areas as desert with prec_raster < 250 (general definition) and 
then convert to polygons with terra::as.polygons

Add this to plot to outline desert areas

Get some point or **shape** data (points: Tucson, PHX, Hermosillo)
Hermosillo: 29.016, -110.926
Phoenix: 33.453, -111.947
Tucson: 32.233, -110.953

#### Learning objectives

After this lesson, learners should be able to:

1. Install packages for geospatial data visualization and analysis
2. Explain the difference between raster and shape data
3. Load geospatial raster data into memory
4. Visualize raster data
5. Change spatial projection of geospatial data
6. Add points to raster data visualization

## [DESCRIPTION OR MOTIVATION; 2-4 sentences that would be used for an announcement]

***

## Getting started

### Workspace organization

First we need to setup our development environment. Open RStudio and create a 
new project via:

+ File > New Project...
+ Select 'New Directory'
+ For the Project Type select 'New Project'
+ For Directory name, call it something like "intro-geo" (without the quotes)
+ For the subdirectory, select somewhere you will remember (like "My Documents" 
or "Desktop")

We need to create two folders: 'data' will store the data we will be analyzing, 
and 'output' will store the results of our analyses. In the RStudio console:

```{r workspace-setup, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

It is good practice to keep input (i.e. the data) and output separate. 
Furthermore, any work that ends up in the **output** folder should be completely
disposable. That is, the combination of data and the code we write should allow 
us (or anyone else, for that matter) to reproduce any output.

### Install additional R packages

**TODO** Update this section

There are _two_ additional R packages that will need to be installed:

+ dplyr
+ randomforest

To install these, run:

```{r install-libraries, eval = FALSE}
install.packages("dplyr")
install.packages("randomforest")
```

### Example data

***

## [TOPIC ONE]

***

## [TOPIC TWO]

***

## Additional resources

+ [resource one](url-one)
+ [resource two](url-two)
+ A [PDF version](https://jcoliver.github.io/learn-r/lesson-name.pdf) of this lesson
