---
title: "Intro to Geospatial data in R"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r library-check, echo = FALSE, results = "hide"}
#| output: false

libs <- c("terra")
libs_missing <- libs[!(libs %in% installed.packages()[, "Package"])]
if (length(libs_missing) > 0) {
  for (l in libs_missing) {
    install.packages(l)
  }
}
```

An introduction to working with two types of geospatial data in R, with 
examples of displaying data on maps.

#### Learning objectives

After this lesson, learners should be able to:

1. Install packages for geospatial data visualization and analysis
2. Explain the difference between raster and shape data
3. Load geospatial raster data into memory
4. Visualize raster data
5. Change spatial projection of geospatial data
6. Add points to raster data visualization

Geospatial data are data that have some connection to specific places or areas 
on Earth. In this lesson we will look at common geospatial data types and how 
we can work with them in R. We will really be scratching the surface today and 
you can find links to more resources at the end of the lesson.

***

## Getting started

### Workspace organization

First we need to setup our development environment. Open RStudio and create a 
new project via:

+ File > New Project...
+ Select 'New Directory'
+ For the Project Type select 'New Project'
+ For Directory name, call it something like "intro-geo" (without the quotes)
+ For the subdirectory, select somewhere you will remember (like "My Documents" 
or "Desktop")

We need to create two folders: 'data' will store the data we will be analyzing, 
and 'output' will store the results of our analyses. In the RStudio console:

```{r workspace-setup, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

It is good practice to keep input (i.e. the data) and output separate. 
Furthermore, any work that ends up in the **output** folder should be completely
disposable. That is, the combination of data and the code we write should allow 
us (or anyone else, for that matter) to reproduce any output.

### Install additional R packages

To work with geospatial data in R, we usually need to install one or more 
additional packages. The two most popular packages for geospatial data are 
[terra](https://rspatial.org/pkg/) and [sf](https://r-spatial.github.io/sf/). 
In this lesson, we will only be working with terra, and you can find links for 
more information on terra and on sf at the bottom of the lesson. In order to 
use the terra package, we must first install the package with 
`install.packages()`:

```{r terra-install, eval = FALSE}
install.packages("terra")
```

And we load it into memory with the `library()` command:

```{r terra-load, echo = FALSE}
suppressPackageStartupMessages(library(terra))
```

```{r terra-load-print, eval = FALSE}
library(terra)
```

### Example data



***

## Geospatial raster data

***

## Geospatial vector data

***

## Additional resources

+ A really nice [guide to using terra in R](https://rspatial.org/).
+ A cheat-sheet of sorts for 
[terra functions](https://rspatial.github.io/terra/reference/terra-package.html).
+ The [sf package](https://r-spatial.github.io/sf/) is another way to work with 
geospatial data in R. It offers a lot of functionality and works well (most of 
the time) with terra objects.
+ The [OpenStreetMap](https://www.openstreetmap.org) (OSM) project also has an 
R package, [osmdata](https://docs.ropensci.org/osmdata/), that allows you to 
download and use data available from OSM.
+ To really impress your friends, the [tidyterra](https://dieghernan.github.io/tidyterra/)
package offers functions for using terra objects (SpatRasters, SpatVectors, 
etc.) with the ggplot2 package.
+ For a more in-depth look at geospatial applications in R, check out the Data 
Carpentry lesson introducing 
[raster & vector data in R](https://datacarpentry.org/r-raster-vector-geospatial/).
+ A [PDF version](https://jcoliver.github.io/learn-r/019-intro-geo.pdf) of this lesson
