---
title: "LESSON TITLE"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

## TODO: Update library check

Add any libraries that are used to the `libs` vector.
If lesson is base R only, remove this chunk entirely.

```{r library-check, echo = FALSE, results = "hide"}
#| output: false

libs <- c("randomForest", "dplyr")
libs_missing <- libs[!(libs %in% installed.packages()[, "Package"])]
if (length(libs_missing) > 0) {
  for (l in libs_missing) {
    install.packages(l)
  }
}
```

[INTRODUCTORY SENTENCE]

#### Learning objectives

After this lesson, learners should be able to:

1. one
2. two
3. three

## [DESCRIPTION OR MOTIVATION; 2-4 sentences that would be used for an announcement]

## Setup

First we need to setup our development environment. Open RStudio and create a 
new project via:

+ File > New Project...
+ Select 'New Directory'
+ For the Project Type select 'New Project'
+ For Directory name, call it something like "r-stats" (without the quotes)
+ For the subdirectory, select somewhere you will remember (like "My Documents" 
or "Desktop")

We need to create two folders: 'data' will store the data we will be analyzing, 
and 'output' will store the results of our analyses.

```{r setup, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

***

## Getting started

+ Library installation (if necessary)
+ Download data (if necessary)

***

## [TOPIC ONE]

The following was suggested by Anthropic's Claude Haiku 4.5

```{r nnet-test}
# Load the nnet package
library(nnet)

# Load the iris dataset NO, GOD NO
# data(iris)

# Set a seed for reproducibility
set.seed(123)

# Create the neural network model
# Formula: Species ~ all measurement columns
# size=5 creates one hidden layer with 5 nodes
mod <- nnet(Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, 
            data = iris, 
            size = 5,
            maxit = 100)

# To visualize the network, can use plotnet
library(NeuralNetTools)
plotnet(mod)

# To make predictions
predictions <- predict(mod, iris, type = "class")

# Check accuracy
table(predictions, iris$Species)

# Ha! Too easy.
```

Another dataset is heart disease from Cleveland hospital. Retrieved from 
https://archive.ics.uci.edu/dataset/45/heart+disease

```{r heart-disease}
# Using the processed.cleveland.data file (a csv)
# Has bad column names
cd <- read.csv(file = "~/Desktop/processed.cleveland.data")
colnames(cd) <- c("age", "sex", "cp", "trestbps", "chol", "fbs", "restecg",
                  "thalach", "exang", "oldpeak", "slope", "ca", "thal", "num")
# Change num (heart disease or not)
cd$num <- ifelse(cd$num == 0,
                 0, 1)

# Run logistic regression
cd_lr <- glm(num ~ ., family = binomial, data = cd)
# Get predictions of probabilities
cd_predict <- predict(object = cd_lr, newdata = cd, type = "response")
# Turn predictions into binary values
cd_predict <- round(cd_predict, digits = 0)
# Show confusion matrix
table(cd_predict, cd$num)

# How well does neural network do?
cd_nnet <- nnet(num ~ ., 
            data = cd, 
            size = 5,
            maxit = 100)

# To visualize the network, can use plotnet
plotnet(cd_nnet)

# To make predictions
predictions <- predict(cd_nnet, cd)

# Turn predictions into binary values
predictions <- round(predictions, digits = 0)

# Check accuracy
table(predictions, cd$num)

## Add more nodes to network
# size 5 -> 10
cd_nnet <- nnet(num ~ ., 
            data = cd, 
            size = 10,
            maxit = 100)

# To make predictions
predictions <- predict(cd_nnet, cd)

# Turn predictions into binary values
predictions <- round(predictions, digits = 0)

# Check accuracy
table(predictions, cd$num)
```

```{r hd-test-train}
fold <- rep(1:5, length.out = nrow(cd))

lr_acc <- numeric(5)
nn_acc <- numeric(5)
for (i in 1:5) {
  testing <- cd[fold == i, ]
  training <- cd[fold != i, ]
  lr <- glm(num ~ ., data = training, family = binomial)
  nn <- nnet(num ~ ., data = training, size = 10, maxit = 100)
  lr_predict <- predict(lr, newdata = testing, type = "response")
  lr_predict <- round(lr_predict, digits = 0)
  nn_predict <- predict(nn, newdata = testing)
  nn_predict <- round(nn_predict, digits = 0)
  lr_acc[i] <- sum(abs(lr_predict - testing$num))/nrow(testing)
  nn_acc[i] <- sum(abs(nn_predict - testing$num))/nrow(testing)
}
mean(lr_acc)
mean(nn_acc)
# HA. Logistic regression is better.
```

```{r}
# Try a wine dataset from
# https://archive.ics.uci.edu/dataset/186/wine+quality
# or
# https://archive.ics.uci.edu/dataset/109/wine

wine_data <- read.csv(file = "~/Desktop/wine/wine.data", header = FALSE)
names(wine_data) <- c("cultivar", "ethanol", "malic_acid", "ash", 
                      "ash_alkalinity", "magnisium", "total_phenols", 
                      "flavanoids", "nonflavanoid_phenols", "proanthocyanins",
                      "color_intensity", "color_hue", "od_ratio", "proline")
library(nnet)

wine_scaled <- apply(X = wine_data[, -1], FUN = scale, MARGIN = 2)
wine_scaled <- data.frame(wine_scaled)
wine_scaled <- cbind(wine_data[, 1], wine_scaled)
colnames(wine_scaled)[1] <- "cultivar"

wine_nnet <- nnet(cultivar ~ ., 
                  data = wine_scaled, 
                  size = 4,
                  maxit = 100)

wine_predict <- predict(wine_nnet, newdata = wine_scaled)
wine_predict <- round(wine_predict)
colSums(wine_predict)
table(wine_scaled$cultivar)

# Try a random forest
library(randomForest)
wine_rf <- randomForest(cultivar ~ ., data = wine_scaled)

wine_rf_predict <- predict(wine_rf, newdata = wine_scaled)
table(wine_rf_predict)
```

# Too easy

The problems described above (heart disease, wine categorization) are too easy. 
That is, things like random forest and logistic regression perform just as well 
(or better) than the neural network (rf and nn had 100% accuracy on the wine 
data).

+ This really might be a case for image classification. MNIST is the standard; 
ten handwritten digits in 28x28 pixel images (grayscale, values 0-255). 
https://en.wikipedia.org/wiki/MNIST_database
+ A recent alternative is Fashion-MNIST (or FMNIST), which classifies images of 
pieces of clothing (same size images as MNIST).
https://en.wikipedia.org/wiki/Fashion-MNIST
Data are available at https://github.com/zalandoresearch/fashion-mnist

Going this route might really require keras/tensorflow...

The MNIST data are usually either distributed as images or as a series of 
28x28 arrays. At least one source has converted it to a csv 
https://pjreddie.com/projects/mnist-in-csv/ with 784 columns (28 x 28 = 784).
The pjreddie site above also includes a python script that does the conversion, 
so one could do this on the FMNIST data, too.

These datasets are too large to be handled by the nnet package.

There's a nice, simple lesson with keras+tensorflow at 
https://www.rpubs.com/sedzinfo/keras_mnist

The keras package includes a getting started vignette at 
https://cran.r-project.org/web/packages/keras/vignettes/index.html

For dealing with the data, we can see the gist at 
https://gist.github.com/brendano/39760
This opens a binary file (the extracted .gz file that ends in -ubyte) and reads
it into memory. Image data are called x (MNIST) and training (FMNIST), while 
label data are called y (MNIST) and label (FMNIST).

For our purposes, consider reducing the dataset to 10%. Could even have a 
"bonus track" section to do download of file, R.utils::gunzip to extract the 
gz file, then read in the file with readBin.

```{r first-mnist}
mnist_train <- read.csv(file = "~/Desktop/mnist_train.csv", header = FALSE)

```

***

## [TOPIC TWO]

***

## Additional resources

+ [resource one](url-one)
+ [Neural Networks and Deep Learning](http://neuralnetworksanddeeplearning.com/index.html)
+ A [PDF version](https://jcoliver.github.io/learn-r/lesson-name.pdf) of this lesson

***
