---
title: "Parallel processing in R"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
urlcolor: blue
---

[INTRODUCTORY SENTENCE]

#### Learning objectives

1. Install packages for parallel processing in R
2. Write code to automate repetitive tasks
3. Translate iterative `for` loop code into parallel processing

## [DESCRIPTION OR MOTIVATION; 2-4 sentences that would be used for an announcement]

***

## Getting started

<making a new project>

***

## Repeating ourselves

<Something for each row in a data frame>
Do leave-one-out linear regression

```{r}
# Run linear regression on iris data, using leave-one-out error estimation, 
# in this example, we use Petal.Length to predict Sepal.Length

# Vector to hold error values
errors <- numeric(nrow(iris))

##### for loop
# The for loop iterative approach
for (i in 1:nrow(iris)) {
  # Estimate the model, leaving out the ith row of data
  one_model <- lm(Sepal.Length ~ Petal.Length, data = iris[-i, ])
  # Use the model to predict value for that ith row of data
  predicted_fit <- predict(one_model, newdata = iris[i, ])
  # Calculate difference between observed and predicted value
  errors[i] <- iris$Sepal.Length[i] - predicted_fit[1]
}
# Calculate mean squared errors
mse <- mean(errors^2)

##### lapply
# The lapply vectorized approach. Start by creating a function to do the work; 
# The one argument to lm_mse (x) will be the row number (x); ignore, for the 
# moment, the bad practice of referring to things inside the function that are 
# not passed to or created inside the function (the iris data frame and columns
# therein).
lm_mse <- function(x) {
  # Estimate the model, leaving out the ith row of data
  one_model <- lm(Sepal.Length ~ Petal.Length, data = iris[-x, ])
  # Use the model to predict value for that ith row of data
  predicted_fit <- predict(one_model, newdata = iris[x, ])
  # Calculate difference between observed and predicted value
  error <- iris$Sepal.Length[x] - predicted_fit[1]
  # Send back the error value
  return(error)
}
# Now use that function with lapply to do the work without a for loop
mse_list <- lapply(X = 1:nrow(iris),
                   FUN = lm_mse)
# lapply returns a list, so we need to convert to a vector before squaring
mse <- mean(unlist(mse_list)^2)

##### parallel processing
library(parallel)

# Use two fewer cores than there are available
n <- detectCores() - 2

# Setup the cluster
clust <- makeCluster(n)
# Use the function we defined above with parLapply
mse_list <- parLapply(cl = clust,
                      X = 1:nrow(iris),
                      fun = lm_mse)
# Stop the cluster
stopCluster(cl = clust)
# parLapply also returns a list
mse <- mean(unlist(mse_list)^2)

# Same as above, but keeping track of times
system.time({
  clust <- makeCluster(n)
  a <- parLapply(cl = clust,
                 X = 1:nrow(iris),
                 fun = function(x) {
                   m <- lm(Sepal.Length ~ Petal.Length, data = iris[-x, ])
                   p <- predict(m, newdata = iris[x, ])
                   e <- iris$Sepal.Length[x] - p[1]
                   return(e)
                 })
  stopCluster(cl = clust)
})

mean(unlist(a)^2)


```

***

## Make it functional

<convert the stuff from the for loop into a function>

***

## "Embarrasingly parallizable"

```{r}
# Takes about a second
system.time(
  for (i in 1:300) {
    x <- rnorm(n = 10000)
    y <- x + rnorm(n = length(x), sd = 0.1)
    l <- lm(y ~ x)
    s <- summary(l)
  }
)

library(parallel)

n <- detectCores() - 2
clust <- makeCluster(n)
system.time({
  a <- parLapply(cl = clust,
                 X = rnorm(n = 10000),
                 fun = function(x) {
                   y <- x + rnorm(n = length(x), sd = 0.1)
                   l <- lm(y ~ x)
                   s <- summary(l)
                 })
})
stopCluster(cl = clust)

```

***

## Additional resources

+ [Overview](https://dept.stat.lsa.umich.edu/~jerrick/courses/stat701/notes/parallel.html#using-sockets-with-parlapply) of parallel processing in R, with comparsions of 
different approaches
+ Examples of parallel processing with [parallel, doParallel, and foreach packages](https://www.r-bloggers.com/2018/09/simple-parallel-processing-in-r/)
+ A [PDF version](https://jcoliver.github.io/learn-r/017-intro-parallel.pdf) of this lesson

***

<a href="index.html">Back to learn-r main page</a>
  
Questions?  e-mail me at <a href="mailto:jcoliver@arizona.edu">jcoliver@arizona.edu</a>.