---
title: "Dynamic reports with quarto"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r library-check, echo = FALSE, results = "hide"}
#| output: false

libs <- c("quarto")
libs_missing <- libs[!(libs %in% installed.packages()[, "Package"])]
if (length(libs_missing) > 0) {
  for (l in libs_missing) {
    install.packages(l)
  }
}
```

An introduction to using the `quarto` package in R to produce reproducible, 
dynamic reports. 

# TODO: Add link to knitr lesson?

#### Learning objectives

After this lesson, learners should be able to:

1. Install and use third-party packages for R
2. Become familar with R Markdown syntax
3. Write dynamic reports including code and visualizations

## [DESCRIPTION OR MOTIVATION; 2-4 sentences that would be used for an announcement]

## Setup

# TODO: Do we need to install quarto first? Probably...

# TODO: Maybe this diverges, given that it is Quarto...

First we need to setup our development environment. Open RStudio and create a 
new project via:

+ File > New Project...
+ Select 'New Directory'
+ For the Project Type select 'New Project'
+ For Directory name, call it something like "r-stats" (without the quotes)
+ For the subdirectory, select somewhere you will remember (like "My Documents" 
or "Desktop")

We need to create two folders: 'data' will store the data we will be analyzing, 
and 'output' will store the results of our analyses.

```{r setup, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

***

## Getting started

+ Library installation (if necessary)
+ Download data (if necessary)

***

## [TOPIC ONE Introduction to markdown syntax]

# TODO: Look at default file and describe parts

# TODO: Render the doc

# TODO: A couple of formatting things?

***

## [TOPIC TWO Code chunks]

***

## [TOPIC THREE Template files (doc)]

***

## Additional resources

+ [resource one](url-one)
+ [resource two](url-two)
+ A [PDF version](https://jcoliver.github.io/learn-r/lesson-name.pdf) of this lesson

***
