---
title: "LESSON TITLE"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

## TODO: Update library check

Add any libraries that are used to the `libs` vector.
If lesson is base R only, remove this chunk entirely.

```{r library-check, echo = FALSE, results = "hide"}
#| output: false

libs <- c("randomForest", "dplyr")
libs_missing <- libs[!(libs %in% installed.packages()[, "Package"])]
if (length(libs_missing) > 0) {
  for (l in libs_missing) {
    install.packages(l)
  }
}
```

[INTRODUCTORY SENTENCE]

#### Learning objectives

After this lesson, learners should be able to:

1. one
2. two
3. three

## [DESCRIPTION OR MOTIVATION; 2-4 sentences that would be used for an announcement]

## Setup

First we need to setup our development environment. Open RStudio and create a 
new project via:

+ File > New Project...
+ Select 'New Directory'
+ For the Project Type select 'New Project'
+ For Directory name, call it something like "r-stats" (without the quotes)
+ For the subdirectory, select somewhere you will remember (like "My Documents" 
or "Desktop")

We need to create two folders: 'data' will store the data we will be analyzing, 
and 'output' will store the results of our analyses.

```{r setup, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

***

## Getting started

+ Library installation (if necessary)
+ Download data (if necessary)

***

## [TOPIC ONE]

The following was suggested by Anthropic's Claude Haiku 4.5

```{r nnet-test}
# Load the nnet package
library(nnet)

# Load the iris dataset NO, GOD NO
# data(iris)

# Set a seed for reproducibility
set.seed(123)

# Create the neural network model
# Formula: Species ~ all measurement columns
# size=5 creates one hidden layer with 5 nodes
mod <- nnet(Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, 
            data = iris, 
            size = 5,
            maxit = 100)

# To visualize the network, can use plotnet
library(NeuralNetTools)
plotnet(mod)

# To make predictions
predictions <- predict(mod, iris, type = "class")

# Check accuracy
table(predictions, iris$Species)

# Ha! Too easy.
```

Another dataset is heart disease from Cleveland hospital. Retrieved from 
https://archive.ics.uci.edu/dataset/45/heart+disease

```{r heart-disease}
# Using the processed.cleveland.data file (a csv)
# Has bad column names
cd <- read.csv(file = "~/Desktop/processed.cleveland.data")
colnames(cd) <- c("age", "sex", "cp", "trestbps", "chol", "fbs", "restecg",
                  "thalach", "exang", "oldpeak", "slope", "ca", "thal", "num")
# Change num (heart disease or not)
cd$num <- ifelse(cd$num == 0,
                 0, 1)

# Run logistic regression
cd_lr <- glm(num ~ ., family = binomial, data = cd)
# Get predictions of probabilities
cd_predict <- predict(object = cd_lr, newdata = cd, type = "response")
# Turn predictions into binary values
cd_predict <- round(cd_predict, digits = 0)
# Show confusion matrix
table(cd_predict, cd$num)

# How well does neural network do?
cd_nnet <- nnet(num ~ ., 
            data = cd, 
            size = 5,
            maxit = 100)

# To visualize the network, can use plotnet
plotnet(cd_nnet)

# To make predictions
predictions <- predict(cd_nnet, cd)

# Turn predictions into binary values
predictions <- round(predictions, digits = 0)

# Check accuracy
table(predictions, cd$num)

## Add more nodes to network
# size 5 -> 10
cd_nnet <- nnet(num ~ ., 
            data = cd, 
            size = 10,
            maxit = 100)

# To make predictions
predictions <- predict(cd_nnet, cd)

# Turn predictions into binary values
predictions <- round(predictions, digits = 0)

# Check accuracy
table(predictions, cd$num)
```

```{r hd-test-train}
fold <- rep(1:5, length.out = nrow(cd))

lr_acc <- numeric(5)
nn_acc <- numeric(5)
for (i in 1:5) {
  testing <- cd[fold == i, ]
  training <- cd[fold != i, ]
  lr <- glm(num ~ ., data = training, family = binomial)
  nn <- nnet(num ~ ., data = training, size = 10, maxit = 100)
  lr_predict <- predict(lr, newdata = testing, type = "response")
  lr_predict <- round(lr_predict, digits = 0)
  nn_predict <- predict(nn, newdata = testing)
  nn_predict <- round(nn_predict, digits = 0)
  lr_acc[i] <- sum(abs(lr_predict - testing$num))/nrow(testing)
  nn_acc[i] <- sum(abs(nn_predict - testing$num))/nrow(testing)
}
mean(lr_acc)
mean(nn_acc)
# HA. Logistic regression is better.
```

```{r}
# Try a wine dataset from
# https://archive.ics.uci.edu/dataset/186/wine+quality
# or
# https://archive.ics.uci.edu/dataset/109/wine

```

***

## [TOPIC TWO]

***

## Additional resources

+ [resource one](url-one)
+ [Neural Networks and Deep Learning](http://neuralnetworksanddeeplearning.com/index.html)
+ A [PDF version](https://jcoliver.github.io/learn-r/lesson-name.pdf) of this lesson

***
