---
title: "LESSON TITLE"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

## TODO: Update library check

Add any libraries that are used to the `libs` vector.
If lesson is base R only, remove this chunk entirely.

```{r library-check, echo = FALSE, results = "hide"}
#| output: false

libs <- c("randomForest", "dplyr")
libs_missing <- libs[!(libs %in% installed.packages()[, "Package"])]
if (length(libs_missing) > 0) {
  for (l in libs_missing) {
    install.packages(l)
  }
}
```

[INTRODUCTORY SENTENCE]

#### Learning objectives

After this lesson, learners should be able to:

1. one
2. two
3. three

## [DESCRIPTION OR MOTIVATION; 2-4 sentences that would be used for an announcement]

## Setup

First we need to setup our development environment. Open RStudio and create a 
new project via:

+ File > New Project...
+ Select 'New Directory'
+ For the Project Type select 'New Project'
+ For Directory name, call it something like "r-stats" (without the quotes)
+ For the subdirectory, select somewhere you will remember (like "My Documents" 
or "Desktop")

We need to create two folders: 'data' will store the data we will be analyzing, 
and 'output' will store the results of our analyses.

```{r setup, eval = FALSE}
dir.create(path = "data")
dir.create(path = "output")
```

***

## Getting started

+ Library installation (if necessary)
+ Download data (if necessary)

***

## [TOPIC ONE]

The following was suggested by Anthropic's Claude Haiku 4.5

```{r nnet-test}
# Load the nnet package
library(nnet)

# Load the iris dataset NO, GOD NO
# data(iris)

# Set a seed for reproducibility
set.seed(123)

# Create the neural network model
# Formula: Species ~ all measurement columns
# size=5 creates one hidden layer with 5 nodes
mod <- nnet(Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, 
            data = iris, 
            size = 5,
            maxit = 100)

# To visualize the network, can use plotnet
library(NeuralNetTools)
plotnet(mod)

# To make predictions
predictions <- predict(mod, iris, type = "class")

# Check accuracy
table(predictions, iris$Species)

# Ha! Too easy.
```

***

## [TOPIC TWO]

***

## Additional resources

+ [resource one](url-one)
+ [Neural Networks and Deep Learning](http://neuralnetworksanddeeplearning.com/index.html)
+ A [PDF version](https://jcoliver.github.io/learn-r/lesson-name.pdf) of this lesson

***
