---
title: "Interpreting Model Coefficients"
author: "Jeff Oliver"
date: "11/2/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

When interpreting coefficients of a model, I find it easiest to formalize the 
model mathematically first. For example, in the following model:

```{r full-model, eval = FALSE}
lm(formula = ppengag ~ depsym * sexparents.F * pemergencysupportnaF +
    pfinancialsupportnaF + pemotionalsupportnaF + childsex.F +
    mathardship + currentrelatwithBSFna.F + seenchildpastmonthna.F +
    livewithchildna.Fr, data = d)
```

we have 10 individual predictor variables, plus the interactions among three of 
the variables (for another four predictors). For the remainder of this 
explanation, we will focus on a simpler model, only including those variables 
that are involved in interaction terms:

```{r simple-model, eval = FALSE}
lm(formula = ppengag ~ depsym * sexparents.F * pemergencysupportnaF, data = d)
```

This model can be mathematically expressed as

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_1 \times depsym + \beta_2 \times sexparents.F + \\
  & \beta_3 \times pemergencysupportnaF + \\
  & \beta_4 \times depsym \times sexparents.F + \\
  & \beta_5 \times depsym \times pemergencysupportnaF + \\
  & \beta_6 \times sexparents.F \times pemergencysupportnaF + \\
  & \beta_7 \times depsym \times sexparents.F \times pemergencysupportnaF
\end{aligned}
$$

We can use this formulation of the model with the output of the `summary` 
command to see the predicted values for each situation. The output of `summary` 
produces (among other things) a coefficients table (If you want to skip to the 
R code without the math, head to the **Calculations** section near the end of 
this document). The table below shows how the $\beta$ values from the equation 
above correspond to values in the output of `summary`:

|$\beta$  | Name in table | value |
|:--------|:--------------|------:|
|$\beta_0$| (Intercept) | 5.9929707 |
|$\beta_1$| depsym | 0.0087369 |
|$\beta_2$| sexparents.FMale | -0.1730717 |
|$\beta_3$| pemergencysupportnaFYes | 0.0808494 |
|$\beta_4$| depsym:sexparents.FMale | -0.8209010 |
|$\beta_5$| depsym:pemergencysupportnaFYes | -0.0703307 |
|$\beta_6$| sexparents.FMale:pemergencysupportnaFYes | -1.3248714 |
|$\beta_7$| depsym:sexparents.FMale:pemergencysupportnaFYes | 0.8394927 |

For `sexparents.F` and `pemergencysupportnaF`, there are two levels. R treats 
the first level as a 0 and the second level as a 1. We can find out which level 
is the _second_ level by looking at the coefficients table. The value used as 
the second level shows up immediately after the name of the variable. So for 
`sexparents.F`, we know the second level is Male because the variable always 
shows up as `sexparents.FMale`. Similarly, the second level of 
`pemergencysupportnaF` is Yes, because that variable shows up in the 
coefficients table as `pemergencysupportnaFYes`.

***

We can now start creating separate formulas for the different cases. When 
`sexparents.F` is Female, we substitute in a 0 for each `sexparents.F`:

$sexparents.F = Female$

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_1 \times depsym + \beta_2 \times 0 + \\
  & \beta_3 \times pemergencysupportnaF + \\
  & \beta_4 \times depsym \times 0 + \\
  & \beta_5 \times depsym \times pemergencysupportnaF + \\
  & \beta_6 \times 0 \times pemergencysupportnaF + \\
  & \beta_7 \times depsym \times 0 \times pemergencysupportnaF
\end{aligned}
$$
Which reduces to:

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_1 \times depsym + \\
  & \beta_3 \times pemergencysupportnaF + \\
  & \beta_5 \times depsym \times pemergencysupportnaF
\end{aligned}
$$

***

For the case when `sexparents.F` is Male, we substitute in a 1 for each 
`sexparents.F`:

$sexparents.F = Male$

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_1 \times depsym + \beta_2 \times 1 + \\
  & \beta_3 \times pemergencysupportnaF + \\
  & \beta_4 \times depsym \times 1 + \\
  & \beta_5 \times depsym \times pemergencysupportnaF + \\
  & \beta_6 \times 1 \times pemergencysupportnaF + \\
  & \beta_7 \times depsym \times 1 \times pemergencysupportnaF
\end{aligned}
$$
Which simplifies to:
$$
\begin{aligned}
ppengag = & \beta_0 + \beta_1 \times depsym + \beta_2 + \\
  & \beta_3 \times pemergencysupportnaF + \\
  & \beta_4 \times depsym  + \\
  & \beta_5 \times depsym \times pemergencysupportnaF + \\
  & \beta_6 \times pemergencysupportnaF + \\
  & \beta_7 \times depsym \times pemergencysupportnaF
\end{aligned}
$$
And we can group like terms:

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_2 + \\
  & (\beta_1 + \beta_4) \times depsym + \\
  & (\beta_3 + \beta_6) \times pemergencysupportnaF + \\
  & (\beta_5 + \beta_7) \times depsym \times pemergencysupportnaF
\end{aligned}
$$

***

We now have two equations:

$sexparents.F = Female$

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_1 \times depsym + \\
  & \beta_3 \times pemergencysupportnaF + \\
  & \beta_5 \times depsym \times pemergencysupportnaF
\end{aligned}
$$

and 

$sexparents.F = Male$

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_2 + \\
  & (\beta_1 + \beta_4) \times depsym + \\
  & (\beta_3 + \beta_6) \times pemergencysupportnaF + \\
  & (\beta_5 + \beta_7) \times depsym \times pemergencysupportnaF
\end{aligned}
$$

***

Each of _those_ equations can be split into two equations, one each for the 
two values of `pemergencysupportnaF`.

For $sexparents.F = Female$ and $pemergencysupportnaF = No$, we substitute in 0 for 
`pemergencysupportnaF`:

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_1 \times depsym + \\
  & \beta_3 \times 0 + \\
  & \beta_5 \times depsym \times 0
\end{aligned}
$$
Which becomes

$$
\begin{aligned}
ppengag = \beta_0 + \beta_1 \times depsym
\end{aligned}
$$

***

For $sexparents.F = Female$ and $pemergencysupportnaF = Yes$, we substitute in 1 
for `pemergencysupportnaF`:

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_1 \times depsym + \\
  & \beta_3 \times 1 + \\
  & \beta_5 \times depsym \times 1
\end{aligned}
$$
Which becomes (grouping like terms):

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_3 + \\
  & (\beta_1 + \beta_5) \times depsym
\end{aligned}
$$

***

Turning to the other equation, where $sexparents.F = Male$, we start with the 
case where $pemergencysupportnaF = No$, substituting 0 for 
`pemergencysupportnaF`:

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_2 + \\
  & (\beta_1 + \beta_4) \times depsym + \\
  & (\beta_3 + \beta_6) \times 0 + \\
  & (\beta_5 + \beta_7) \times depsym \times 0
\end{aligned}
$$
Reducing to:

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_2 \\
  & (\beta_1 + \beta_4) \times depsym
\end{aligned}
$$

***

For the final case, where $sexparents.F = Male$ and 
$pemergencysupportnaF = Yes$, we again substitute 1 in for 
`pemergencysupportnaF`:

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_2 + \\
  & (\beta_1 + \beta_4) \times depsym + \\
  & (\beta_3 + \beta_6) \times 1 + \\
  & (\beta_5 + \beta_7) \times depsym \times 1
\end{aligned}
$$
Reducing and grouping like terms:

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_2 + \beta_3 + \beta_6 + \\
  & (\beta_1 + \beta_4 + \beta_5 + \beta_7) \times depsym
\end{aligned}
$$

***

We now have four equations that we can use to calculate each of the four 
different cases.


$sexparents.F = Female$ and $pemergencysupportnaF = No$

$$
\begin{aligned}
ppengag = \beta_0 + \beta_1 \times depsym
\end{aligned}
$$

***

$sexparents.F = Female$ and $pemergencysupportnaF = Yes$

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_3 + \\
  & (\beta_1 + \beta_5) \times depsym
\end{aligned}
$$

***

$sexparents.F = Male$ and $pemergencysupportnaF = No$

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_2 \\
  & (\beta_1 + \beta_4) \times depsym
\end{aligned}
$$

***

$sexparents.F = Male$ and $pemergencysupportnaF = Yes$

$$
\begin{aligned}
ppengag = & \beta_0 + \beta_2 + \beta_3 + \beta_6 + \\
  & (\beta_1 + \beta_4 + \beta_5 + \beta_7) \times depsym
\end{aligned}
$$

***

## Calculations

I'll focus on just the last case, where $sexparents.F = Male$ and 
$pemergencysupportnaF = Yes$. We will do calculations for two values of 
`depsym` the minimum observed value and the maximum observed value (if 
`depsym` is categorical with two levels, the minimum is 0 and maximum is 1). 
When indexing the $\beta$ values, we have to remember to add 1 because R starts 
numbering at 1, not zero. The $\beta$ values are stored in the first column of 
the coefficients table.

```{r calc-model, eval = FALSE}
# Extract coefficients table from model
lm_coeff <- summary(lmslopes)$coefficients
# Calculate value for smallest value of depsym
max_depsym <- lm_coeff[1, 1] + lm_coeff[3, 1] + lm_coeff[4, 1] + lm_coeff[7, 1] + 
  (lm_coeff[2, 1] + lm_coeff[5, 1] + lm_coeff[6, 1] + lm_coeff[8, 1]) * max(d$depsym) 
# Calculate value for smallest value of depsym
min_depsym <- lm_coeff[1, 1] + lm_coeff[3, 1] + lm_coeff[4, 1] + lm_coeff[7, 1] + 
  (lm_coeff[2, 1] + lm_coeff[5, 1] + lm_coeff[6, 1] + lm_coeff[8, 1]) * min(d$depsym) 
```
