---
title: "Land cover data for species distribution modeling"
author: "Jeff Oliver"
date: "2021-11-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

An investigation into the use of MODIS land-cover data for species distribution 
models.

# Setup

## Dependencies

The work below uses additional package to download, visualize, and analyze the 
data.

```{r load-libraries, echo = FALSE}
suppressPackageStartupMessages({
  library(MODIStsp)    # downloading land cover data
  library(raster)      # reading and transforming geospatial data
  library(ggplot2)     # data visualization
  library(dplyr)       # data wrangling
})
```

```{r load-libraries-display, eval = FALSE}
library(MODIStsp)    # downloading land cover data
library(raster)      # reading and transforming geospatial data
library(ggplot2)     # data visualization
library(dplyr)       # data wrangling
```

## Earthdata access

To access the data, you'll need to register with NASA's Earthdata system at 
[https://urs.earthdata.nasa.gov/users/new](https://urs.earthdata.nasa.gov/users/new). 
After registering, also be sure to navigate to your 
[profile page](https://urs.earthdata.nasa.gov/profile), click the on the 
Applications menu and the Authorized Apps submenu. Click the "Approve More 
Applications" button near the bottom of the page and click the "Authorize" 
button in the row labeled "LP DAAC Data Pool". Make a note of your username and 
password, as we will need those when downloading the data. Note the password 
is not super-securely stored by NASA, so don't use one that is especially 
important to you (I'm not going to lecture anyone about never re-using 
passwords - we all do it).

# Data retrieval

In this example, we use land cover data available from
[MODIS](https://modis.gsfc.nasa.gov/data/dataprod/). There are many means of 
categorizing land cover, and in this example we use the land cover scheme of 
the International Geosphere-Biosphere Programme (IGBP) that identifies 17 
classes of land cover. If you want to get details about data availability, you 
can run:

```{r modis-check}
MODIStsp::MODIStsp_get_prodlayers("MCD12Q1")
```

It's that first band that we want ("LC1", or "Land Cover Type 1 (IGBP)*"). To 
download the data, we use the `MODIStsp()` function from the MODIStsp package. 
Much of the code for downloading and visualizing these data from the example at 
[https://rspatialdata.github.io/land_cover.html](https://rspatialdata.github.io/land_cover.html).
Here's where you'll need to add your usename and password that you used to 
register for NASA's Earthdata site. In the example below, I stored them in a 
pair of plain text files called earthdata-username.txt and 
earthdata-password.txt. I then read in the values and stored them in variables 
`username` and `password` respectively.

```{r load-credentials}
username <- scan(file = "../keys/earthdata-username.txt", what = character())
password <- scan(file = "../keys/earthdata-password.txt", what = character())
```

You could also do this in the console if you don't want to store your username 
and/or password in files (replacing `<INSERT USERNAME HERE>` and 
`<INSERT PASSWORD HERE>` with your actual username and password):

```{r assign-credentials, eval = FALSE}
# Run in the console:
username <- "<INSERT USERNAME HERE>"
password <- "<INSERT PASSWORD HERE>"
```

We will download land cover data for a box that roughly encapsulates the 
Sonoran desert. The code below downloads the data to a folder called "data", so 
be sure that you have created such a folder first (i.e. `dir.create("data")`).

```{r download-sonoran-raster, eval = FALSE}
# Boundaries of interest (left, bottom, right, top); here we use very rough 
# coordinates for the Sonoran desert
bounds <- c(-114, 26, -109, 35)

# Need to specify which projection to use; we do this with the WKT (Well Known 
# Text) string for the WGS 84 projection. A terrifying thing, but it works.
wkt_proj <- 'GEOGCS["WGS 84",DATUM["WGS_1984",SPHEROID["WGS
84",6378137,298.257223563,AUTHORITY["EPSG","7030"]],AUTHORITY["EPSG","6326"]],
PRIMEM["Greenwich",0,AUTHORITY["EPSG","8901"]],UNIT["degree",
0.01745329251994328,AUTHORITY["EPSG","9122"]],AUTHORITY["EPSG","4326"]]'

MODIStsp::MODIStsp(gui = FALSE,
                   out_folder = "data",
                   out_folder_mod = "data",
                   selprod = "LandCover_Type_Yearly_500m (MCD12Q1)",
                   bandsel = "LC1", 
                   start_date = "2019.01.01", 
                   end_date = "2019.12.31", 
                   user = username,
                   password = password,
                   verbose = FALSE, # set to true if you want to see details
                   reprocess = TRUE,
                   spatmeth = "bbox",
                   bbox = bounds,
                   out_projsel = "User Defined",
                   output_proj = wkt_proj,
                   out_format = "GTiff")
```

The servers where these data live are notorious for timing out. Sometimes it 
takes a few times of running the above code for it to actually work. If you see 
multiple lines of 

`Error in curl::curl_fetch_memory(url, handle = handle): Timeout was reached`

Try running the `MODIStsp()` command again.

As long as the download was successful, we can do a reality check and plot the 
land cover data. We will need to read the data into memory and transform it 
into a data frame for easier plotting with ggplot

```{r plot-raster}
# Load in raster from file
IGBP_raster <- raster(x = "data/LandCover_Type_Yearly_500m_v6/LC1/MCD12Q1_LC1_2019_001.tif")

# Convert to a data frame for plotting
IGBP_df <- as.data.frame(IGBP_raster, xy = TRUE, na.rm = TRUE) %>%
  mutate(MCD12Q1_LC1_2019_001 = as.factor(round(MCD12Q1_LC1_2019_001)))

# Reset rownames so they are numbered, starting at 1
rownames(IGBP_df) <- c()

# Renaming IGBP classification levels, since incoming data just enumerates them
# 1-17
levels(IGBP_df$MCD12Q1_LC1_2019_001) <- c( "Evergreen needleleaf forests",
                                           "Evergreen broadleaf forests",
                                           "Deciduous needleleaf forests",
                                           "Deciduous broadleaf forests",
                                           "Mixed forests",
                                           "Closed shrublands",
                                           "Open shrublands",
                                           "Woody savannas",
                                           "Savannas",
                                           "Grasslands",
                                           "Permanent wetlands",
                                           "Croplands",
                                           "Urban and built-up lands",
                                           "Cropland/natural vegetation mosaics",
                                           "Snow and ice",
                                           "Barren",
                                           "Water bodies")
# Visualising using ggplot2
lc_plot <- ggplot(data = IGBP_df, 
                  mapping = aes(x = x, y = y, fill = MCD12Q1_LC1_2019_001)) + 
  geom_raster() +
  scale_fill_viridis_d(name = "Land Cover Type") + # the color scheme
  labs(subtitle = "Land cover 2019",
       x = "Longitude",
       y = "Latitude") +
  theme_minimal() +
  theme(legend.text = element_text(size = 8))
print(lc_plot)
```

# Species distribution modeling

Stay tuned!